---
title: "Sanger sequencing summary statistics and trees"

editor: visual
bibliography: ZosteropsSangerSeqsRefs.bib
---

## Goals

What follows is our workflow for the general descriptive analysis of the Sanger sequencing data set for Zosterops. In this analysis we will describe the variation across individual Sanger sequenced loci and construct trees for each of these loci using distance-based and maximum likelihood methods.

## Getting started

We are using R version 4.4.2, RStudio version 2024.09.1 for the analysis. Quarto version 1.6.39 for Mac OS was used to document the workflow. Set the working directory in RStudio to one where the necessary phased Sanger sequencing alignments and other data are accessible.

## General methods summary

### gDNA extraction

Whole gDNA was isolated from either blood or tissue samples using a magnetic bead based protocol (University of Kansas) [@rohland2012] or standard phenol-chloroform-isoamyl alcohol (PCI) protocol followed by ethanol precipitation (Cincinnati Museum Center and Marshall University) and quantified by a Qubit dsDNA Assay.

### PCR and Sanger sequencing

159 individual samples underwent polymerase chain reactions (PCRs) and Sanger sequencing across eight loci including two mtDNA and six nuclear protein-coding exons (nuDNA). Common sequence tags, either CS1 (5′-ACACTGACGACATGGTTCTACA) in the forward direction or CS2 (5′-TACGGTAGCAGAGACTTGGTCT) in the reverse direction, were added to all the oligonucleotide sequences in this study [@moonsamy2013]. The addition of common 5′ tags allowed for CS1 and CS2 oligonucleotides to be utilized as sequencing primers for all amplicons across all loci. The NADH dehydrogenase subunit 2 (ND2) and cytochrome c oxidase subunit 1 (COI) were amplified using the primers L5216 and H6313 [@sorenson1999] and BIRDF1 and BIRDR2 [@hebert2004], respectively. Six nuDNA coding loci (B3GNT9, BIRC2, CCDC, GPRC6, LACTBL, and UBP38) were amplified using primers from Liu *et al.* [@liu2018]. A complete list of the oligonucleotide sequences used as PCR primers in this study are shown below.

| Oligonucleotide name | 5′-CS tag + target sequence (5′ → 3′) | Locus description | Location/reference species | References |
|----|----|----|----|----|
| L5216-CS1 | ACACTGACGACATGGTTCTACAGGCCCATACCCCGRAAATG | NADH dehydrogenase subunit 2 (ND2) | mtDNA/avian | Sorenson et al. 1999 |
| H6313-CS2 | TACGGTAGCAGAGACTTGGTCTACTCCTRTTTAAGGCTTTGAAGGC | NADH dehydrogenase subunit 2 (ND2) | mtDNA/avian | Sorenson et al. 1999 |
| BIRDF1-CS1 | ACACTGACGACATGGTTCTACATTCTCCAACCACAAAGACATTGGCAC | Cytochrome c Oxidase Subunit 1 (COI) | mtDNA/avian | Hebert et al. 2004 |
| BIRDR2-CS2 | TACGGTAGCAGAGACTTGGTCTACTACATGTGAGATGATTCCGAATCCAG | Cytochrome c Oxidase Subunit 1 (COI) | mtDNA/avian | Hebert et al. 2004 |
| B3GNT9F-CS1 | ACACTGACGACATGGTTCTACACCATTGGAAAGGCTTCTTCAC | betaGal beta-1,3-N-acetylglucosaminyltransferase 9 | chromosome 11/*Taeniopygia guttata* | Liu et al. 2018 |
| B3GNT9R-CS2 | TACGGTAGCAGAGACTTGGTCTTTTCTGCTCATTGCTATCAAATC | betaGal beta-1,3-N-acetylglucosaminyltransferase 9 | chromosome 11/*Taeniopygia guttata* | Liu et al. 2018 |
| BIRC2F-CS1 | ACACTGACGACATGGTTCTACAATCTGCAAGCTGTTCAGGC | baculoviral IAP repeat containing 2 | chromosome 1/*Taeniopygia guttata* | Liu et al. 2018 |
| BIRC2R-CS2 | TACGGTAGCAGAGACTTGGTCTCCGGGCTGGGTTTTATTAC | baculoviral IAP repeat containing 2 | chromosome 1/*Taeniopygia guttata* | Liu et al. 2018 |
| CCDCF-CS1 | ACACTGACGACATGGTTCTACACCAAAGTTACTGGAGAACCTTCC | coiled-coil domain containing 141 | chromosome 7/*Taeniopygia guttata* | Liu et al. 2018 |
| CCDCR-CS2 | TACGGTAGCAGAGACTTGGTCTATACAGAGCTGTCATCTTCAGCC | coiled-coil domain containing 141 | chromosome 7/*Taeniopygia guttata* | Liu et al. 2018 |
| GPRC6F-CS1 | ACACTGACGACATGGTTCTACAACAACAAAACCTACTGGGC | G protein-coupled receptor, class C, group 6, member A | chromosome 3/*Taeniopygia guttata* | Liu et al. 2018 |
| GPRC6R-CS2 | TACGGTAGCAGAGACTTGGTCTCAGGGATAAATACAATCCATGCT | G protein-coupled receptor, class C, group 6, member A | chromosome 3/*Taeniopygia guttata* | Liu et al. 2018 |
| LACTBLF-CS1 | ACACTGACGACATGGTTCTACACCAGTGAGATGGAGGCAGAG | lactamase, beta-like 1 | chromosome 21/*Taeniopygia guttata* | Liu et al. 2018 |
| LACTBLR-CS2 | TACGGTAGCAGAGACTTGGTCTATCCGCTCCCAGATGGC | lactamase, beta-like 1 | chromosome 21/*Taeniopygia guttata* | Liu et al. 2018 |
| UBP38F-CS1 | ACACTGACGACATGGTTCTACACCTGAGATCCTTAGTGGAGACAA | ubiquitin carboxyl-terminal hydrolase 38 | chromosome 4/*Taeniopygia guttata* | Liu et al. 2018 |
| UBP38R-CS2 | TACGGTAGCAGAGACTTGGTCTGCATCCATGAGGTCTTTTTGTAG | ubiquitin carboxyl-terminal hydrolase 38 | chromosome 4/*Taeniopygia guttata* | Liu et al. 2018 |

PCRs were performed in 10μL reactions in a 96-well thermal cycler (Life Technologies) with commercial polymerase master mixes (either Fidelitaq™ Master Mix, Affymetrix/USB or AmpliTaq Gold™ 360 Master Mix, Applied Biosystems™) and forward and reverse primers (0.5 μmol L^−1^ each; Integrated DNA Technology) using a thermal profile of 94°C for 4 min followed by 30 cycles of 1 min at 94°C, 1 min at 50°C, and 2 min at 72°C, and then a final extension cycle for 10 min at 72°C. The same thermal profile was utilized for all PCRs across all loci. Primers and excess dNTPs were inactivated with ExoSAP-IT™ (Applied Biosystems™) following the manufacturer’s instructions. Sequencing reactions were performed on ExoSAP-IT^®^ treated PCR amplicons either in-house (Cincinnati Museum Center or Marshall University) using BigDye^®^ v3.1 Cycle Sequencing and BigDye Xterminator^®^ kits (Life Technologies) and separating on a 3130 Genetic Analyzer (Life Technologies) or by submitting for Sanger sequencing to Eurofins Genomics using their Prepaid Plate DNA kits. PCR amplicons were sequenced in both directions with primers based on the common sequence tags incorporated into all amplicons (CS1 and CS2).

Complementary strands were aligned and edited using *Geneious* v8.1.4 (Biomatters, available from http://www.geneious.com) [@kearse2012]. Chromatograms were inspected and every variation was checked for authenticity. The phase of nuclear alleles was determined statistically using *Seqphase* [@flot2009] and *Phase* v2.1 [@stephens2001] and phased nuclear haplotypes aligned using *Geneious* (Biomatters). General descriptive statistics for Sanger sequenced loci including nucleotide diversity, haplotype diversity, segregating sites, q, Tajima’s D [@tajima1989], and the R~2~ test [@ramos-onsins2002] as well as substitution model tests and trees and networks (distance-based and maximum likelihood) for individual loci were generated using the R packages *ape* [@ape-2], *pegas* [@pegas-2], and *phangorn* [@phangorn-2]. R v4.4.2 was used throughout various analyses in this study (R Core Team, 2024).

## Data files

This dataset includes Sanger sequenced polymerase chain reaction (PCR) amplicon alignments for two mitochondrial DNA (mtDNA) and six protein coding nuclear loci (nuDNA). Prior to this analysis Sanger sequencing `ab1` files were uploaded to Geneious^Ⓡ^ 8.1.9 for editing, alignment, and quality control. Multiple sequencing runs for each amplicon were assembled and assembled loci were aligned across individual samples using Geneious Alignment. Alignments for each locus were trimmed and edited in Geneious and the identity of each locus confirmed with a megablast search. The reading frame was set by eye by avoiding frames with stop codons and a blastp search was done on the translation consensus for each alignment to confirm the locus identity. Alignments were exported as `fasta` files.

Alignments of diploid cdDNA loci were phased using PHASE 2.1.1 [@stephens2001]. First, convert the alignment `fasta` files from Geneious to the `inp` format for PHASE 2.1.1 using the online [SeqPHASE tool](https://eeg-ebe.github.io/SeqPHASE/step1.html) [@flot2009]. At the 'Genotypes to be phased' prompt click the 'Browse' button and select the appropriate `fasta` alignment. Click the 'Run SeqPHASE' button. Once step 1 of the SeqPHASE process is completed retain the `seqphase.const` file in addition to the `seqphase.inp` file and rename each file appropriately to match the locus name (e.g. `B3GNT9_seqphase.inp` and `B3GNT9_seqphase.const` for locus B3GNT9).

Move the `inp` file from SeqPHASE to the directory with PHASE 2.1.1 and `cd` to that directory in the terminal. The `inp` file is used to run the following PHASE 2.1.1 command in the terminal. For each locus `<numberseed>` should be replaced with an arbitrary number selected by the user and `LOCUSNAME` replaced with the designated name for that particular locus. Because each locus is relatively short chose the `S` option at the modeling flag (`-M`) as it assumes no within-locus recombination. Additional parameters are explained in the documentation for PHASE 2.1.1 available [here](https://stephenslab.uchicago.edu/assets/software/phase/instruct2.1.pdf).

```         
./PHASE -MS -d1 -S<numberseed> -p0.85 -q0.85 -x5 -X10 LOCUSNAME_seqphase.inp LOCUSNAME_seqphase.out 5000 2 500 
```

Further analysis will require `fasta` alignments for each locus where each diploid individual is associated with two sequences corresponding to each of the two phased alleles determined in PHASE 2.1.1. Return to the online [SeqPHASE tool](https://eeg-ebe.github.io/SeqPHASE/step1.html) and click the link to '[step 2](https://eeg-ebe.github.io/SeqPHASE/step2.html)'. Select the appropriate `out` file generated by the PHASE 2.1.1 run along with the matching `const` file generated from step 1 in the SeqPHASE tool. The 'Sort sequence' box should be checked but the 'Reduce sequences' box should not be checked. Click the 'Run SeqPHASE' button and SeqPHASE will return a `fasta` file with the complete sequence alignment and two phased haplotypes designated by the project number for that individual sample followed by a and b respectively. Click the 'Click here to download the generated Fasta file' link at the bottom of the page and rename the file to match the locus name. Including the `const` file and unchecking the 'Reduced sequences' box will ensure the resulting `fasta` file includes the full length of the alignment, not just the variable sites.

The phasing procedure was repeated for the following nuDNA `fasta` alignments.

- `B3GNT9.fasta`

- `BIRC2.fasta`

- `CCDC.fasta`

- `GPRC6.fasta`

- `LACTBL.fasta`

- `UBP38.fasta`

Subsequent analysis requires the following phased nuDNA alignments from the procedure described above.

- `COI.fasta`

- `ND2.fasta`

- `B3GNT9_phased.fasta`

- `BIRC2_phased.fasta`

- `CCDC_phased.fasta`

- `GPRC6_phased.fasta`

- `LACTBL_phased.fasta`

- `UBP38_phased.fasta`

## Description of summary statistics and tree building procedure for a single locus

### Packages

Begin the analysis by restarting the R session (Session/Restart R in RStudio), loading the initial package libraries, and starting a log of the analysis. Other package libraries will be loaded as needed in later blocks. The following R packages should be installed using the Tools/Install Packages feature in RStudio.

- logr [@logr] Optional if you wish to create a log file for the analysis.

- ape [@ape]

- pegas [@pegas]

- phangorn [@phangorn]

- gt [@gt]

- webshot2 [@webshot2]

- dplyr [@dplyr]

```{r}
#| label: Load libraries
library(logr)
log_open(file_name = "ZosteropsSangerSeqs.log")
library(ape)
library(pegas)
library(phangorn)
```

### Cytochrome c oxidase subunit I (COI)

### Load and visualize the alignment

First is a step-by-step protocol for the first locus (cytochrome c oxidase subunit I, COI). The same procedure will also be implemented for the other loci.

Load and check the `fasta` file and generate a standard color-coded visualization of the alignment.

```{r}
#| label: COI load FASTA alignment
#| fig-width: 30
#| fig-height: 17
#COI
#Load FASTA alignment and check
ZosteropsCOI <- read.dna("Data/Alignments/COI.fasta", format = "fasta")
ZosteropsCOI
checkAlignment(ZosteropsCOI, check.gaps = FALSE, plot = TRUE, what = 1)
```

### Pairwise divergence

Determine the pairwise divergences among the sequences and display the 40 least divergent pairwise comparisons and the 40 greatest pairwise comparisons. Also focus on a particular pair of samples between *Z. japonicus* and *Z. montanus* (i.e. ZJal003a and ZMOvu001a) to get an idea of the mtDNA divergence within this focal group.

```{r}
#| label: COI pairwise divergence

# Calculate pairwise divergences
pw_div_COI <- dist.dna(ZosteropsCOI, model = "raw")
# Convert the distance matrix to a data frame
pw_div_df_COI <- as.data.frame(as.matrix(pw_div_COI))
# Add row names as a column
pw_div_df_COI$Sequence <- rownames(pw_div_df_COI)
# Create a long format data frame
pw_div_long_format_COI <- data.frame(Sequence = rep(pw_div_df_COI$Sequence, each = nrow(pw_div_df_COI)),
                               Comparison = rep(pw_div_df_COI$Sequence, times = nrow(pw_div_df_COI)),
                               Divergence = as.vector(as.matrix(pw_div_COI)))
# Remove self-comparisons
pw_div_long_format_COI <- pw_div_long_format_COI[pw_div_long_format_COI$Sequence != pw_div_long_format_COI$Comparison, ]
# Find the least 40 divergences
pw_div_least_COI <- head(pw_div_long_format_COI[order(pw_div_long_format_COI$Divergence), ], 40)
# Find the greatest 40 divergences
pw_div_greatest_COI <- head(pw_div_long_format_COI[order(-pw_div_long_format_COI$Divergence), ], 40)
# Combine least and greatest divergences
pw_div_combined_results_COI <- rbind(pw_div_least_COI, pw_div_greatest_COI)
# View the results
print(pw_div_combined_results_COI)

# Filter for a specific pair of samples
specific_pair_divergence <- pw_div_long_format_COI[
  (pw_div_long_format_COI$Sequence == "ZJal003a" & pw_div_long_format_COI$Comparison == "ZMOvu001a") |
  (pw_div_long_format_COI$Sequence == "ZMOvu001a" & pw_div_long_format_COI$Comparison == "ZJal003a"), 
]

# View the results
print(specific_pair_divergence)

```

Graph the Shannon index (H) per base and the number of bases in the alignment per site and show a summary.

```{r}
#| label: COI Shannon index, number of bases per site, and summary. 
checkAlignment(ZosteropsCOI, check.gaps = FALSE, plot = TRUE, what = 3)
checkAlignment(ZosteropsCOI, check.gaps = FALSE, plot = TRUE, what = 4)
summaryCOI <- summary(ZosteropsCOI)
```

### Estimate population genetic parameters

#### Nucleotide diversity

Calculate the nucleotide diversity for the locus.

```{r}
#| label: COI nucleotide diversity
# Nucleotide diversity
ndCOI <- nuc.div(ZosteropsCOI)
ndCOI
```

#### Haplotypes

Estimate the number of haplotypes and haplotype diversity.

```{r}
#| label: COI haplotypes
# Number of haplotypes from phased DNA sequences (unique sequences) 
# and hapotype diversity
hapCOI <- haplotype(ZosteropsCOI)
NumberHapsCOI <- nrow(hapCOI)
hapdivCOI <- hap.div(ZosteropsCOI)
hapCOI
NumberHapsCOI
hapdivCOI
```

#### Segregating sites

Estimate the number of segregating sites.

```{r}
#| label: COI segregating sites
# An index of segregating sites and number of segregating sites 
seg.sitesCOI <- seg.sites(ZosteropsCOI)
sCOI <- length(seg.sitesCOI)
seg.sitesCOI
sCOI
```

#### Sequence length

Calculate sequence length.

```{r}
#| label: COI sequence length
# Total length of all sequences (number of sequences x sequence length) 
# and sequence length
lengthCOItotal <- length(ZosteropsCOI)
seqlengthCOI <- ncol(hapCOI)
lengthCOItotal
seqlengthCOI
```

#### Number of sequences

Calculate the number of sequences in the alignment. Diploid phased nuDNA loci will have two sequences per individual sample and haploid mtDNA loci will have a single sequence per locus.

```{r}
#| label: COI number of sequences
# Number of sequences calulated two ways, by divdiding the total 
# length by the sequence length (calculation 1) and by extracting 
# the length value from the summary (caclulation 2)
# Confirm both calculations for number of sequences are the same
NumberSequencesCOI1 <- lengthCOItotal/seqlengthCOI
SummaryCOIlength <- unlist(strsplit(summaryCOI[1,1], 
                                    split = ":", fixed = TRUE))[2]
NumberSequencesCOI2 <- as.numeric(SummaryCOIlength)
all.equal(NumberSequencesCOI1, NumberSequencesCOI2)
NumberSequencesCOI1
NumberSequencesCOI2
```

#### 𝛉 estimates

Estimate 𝛉 per locus, 𝛉 per site, and 𝛉 variance based on the the number of segregating sites (`theta.s`) and also estimate 𝛉 per locus based on the number of alleles (`theta.k`).

```{r}
#| label: COI theta estimates
# Calculate theta per sequence, theta per site, theta variance 
# from the Watterson estimator based on number of segregating sites.
ThetaCOIperLocusW <- theta.s(sCOI, NumberSequencesCOI1)
ThetaCOIperSiteW <- ThetaCOIperLocusW/seqlengthCOI
ThetaCOIperLocusVarW <- theta.s(sCOI, NumberSequencesCOI1, 
                                variance = TRUE)
ThetaCOIperLocusVarW <- ThetaCOIperLocusVarW[2]
ThetaCOIk <- theta.k(n = 132, k = 52)
ThetaCOIperLocusW
ThetaCOIperSiteW
ThetaCOIperLocusVarW
ThetaCOIk
```

#### Tajima's D and R2 tests

Test for potential deviations from neutrality using Tajima's D [@tajima1989] and population growth from the Ramos-Onsins and Roza's test [@ramos-onsins2002].

```{r}
#| label: COI Tajima's D and Ramos-Onsins and Roza's Test (R2 Test)
# Tests for neutrality including Tajima's D and 
# Ramos-Onsins and Roza's test
tajimaCOI <- tajima.test(ZosteropsCOI)
R2COI <- R2.test(ZosteropsCOI,B = 10000, 
                 theta = ThetaCOIperLocusW, plot = TRUE, quiet = FALSE)
tajimaCOI
R2COI
```

### Trees

#### Distance-based trees

Generate trees using the unweighted pair group method with arithmetic mean (UPGMA) and neighbor joining (NJ) distance-based methods.

```{r}
#| label: COI UPGMA and NJ trees
#| fig-width: 60
#| fig-height: 55
#UPGMA and NJ trees
dmCOI <- dist.ml(ZosteropsCOI)
UPGMAtreeCOI <- upgma(dmCOI)
NJtreeCOI <- NJ(dmCOI)
plot(UPGMAtreeCOI, main = "UPGMA COI", cex = 3)
plot(NJtreeCOI, "unrooted", main = "NJ COI", cex = 3)
```

#### Maximum likelihood trees

Generate maximum likelihood (ML) trees rooted either on the midpoint or an outgroup.

```{r}
#| label: COI ML trees
#| fig-width: 60
#| fig-height: 55
#ML tree midpoint and outgroup rooted trees
mtCOI <- modelTest(ZosteropsCOI)
fitCOI <- as.pml(mtCOI, "AIC")
fit_mtCOI <- pml_bb(mtCOI, control = pml.control(trace = 0))
fit_mtCOI
bsCOI <- bootstrap.pml(fit_mtCOI, bs = 100, 
                       optNni = TRUE, control = pml.control(trace = 0))
mlCOI <- plotBS(midpoint(fit_mtCOI$tree), bsCOI, 
                show.node.label = TRUE, p = 50, type="p", 
                main="Maximum likelihood midpoint rooted standard bootstrap", 
                cex = 3)
outgroupCOI <- c("ZAsu001a", "ZAsu002a")
mlCOIroot <- root(mlCOI, outgroup = outgroupCOI)
plot(mlCOIroot, show.node.label = TRUE, p = 50, type="p", 
     main="Maximum Likelihood rooted on outgroup", cex = 3)
```

## Summary statistics and trees for the remaining loci

Below are the code blocks for the remaining loci grouped into loading and summarizing alignments, summary population statistics, and distance-based and ML trees.

### NADH dehydrogenase 2 (ND2)

#### Load and visualize alignment

```{r}
#| label: ND2 load alignment
#| fig-width: 30
#| fig-height: 17
#ND2
#Load FASTA alignment and check
ZosteropsND2 <- read.dna("Data/Alignments/ND2.fasta", format = "fasta")
ZosteropsND2
checkAlignment(ZosteropsND2, check.gaps = FALSE, plot = TRUE, what = 1)
```

Determine the pairwise divergences among the sequences and display the 40 least divergent pairwise comparisons and the 40 greatest pairwise comparisons.

```{r}
#| label: ND2 pairwise divergence

# Calculate pairwise divergences
pw_div_ND2 <- dist.dna(ZosteropsND2, model = "raw")
# Convert the distance matrix to a data frame
pw_div_df_ND2 <- as.data.frame(as.matrix(pw_div_ND2))
# Add row names as a column
pw_div_df_ND2$Sequence <- rownames(pw_div_df_ND2)
# Create a long format data frame
pw_div_long_format_ND2 <- data.frame(Sequence = rep(pw_div_df_ND2$Sequence, each = nrow(pw_div_df_ND2)),
                                     Comparison = rep(pw_div_df_ND2$Sequence, times = nrow(pw_div_df_ND2)),
                                     Divergence = as.vector(as.matrix(pw_div_ND2)))
# Remove self-comparisons
pw_div_long_format_ND2 <- pw_div_long_format_ND2[pw_div_long_format_ND2$Sequence != pw_div_long_format_ND2$Comparison, ]
# Find the least 40 divergences
pw_div_least_ND2 <- head(pw_div_long_format_ND2[order(pw_div_long_format_ND2$Divergence), ], 40)
# Find the greatest 40 divergences
pw_div_greatest_ND2 <- head(pw_div_long_format_ND2[order(-pw_div_long_format_ND2$Divergence), ], 40)
# Combine least and greatest divergences
pw_div_combined_results_ND2 <- rbind(pw_div_least_ND2, pw_div_greatest_ND2)
# View the results
print(pw_div_combined_results_ND2)

```

Graph the Shannon index (H) per base and the number of bases in the alignment per site and show summary.

```{r}
#| label: ND2 Shannon index, number of bases per site, and summary. 
checkAlignment(ZosteropsND2, check.gaps = FALSE, plot = TRUE, what = 3)
checkAlignment(ZosteropsND2, check.gaps = FALSE, plot = TRUE, what = 4)
summaryND2 <- summary(ZosteropsND2)
```

#### Summary statistics

```{r}
#| label: ND2 summary statistics
# Nucleotide diversity
ndND2 <- nuc.div(ZosteropsND2)
# Number of haplotypes from phased DNA sequences (unique sequences) 
# and hapotype diversity
hapND2 <- haplotype(ZosteropsND2)
NumberHapsND2 <- nrow(hapND2)
hapdivND2 <- hap.div(ZosteropsND2)
# An index of segregating sites and number of segregating sites 
seg.sitesND2 <- seg.sites(ZosteropsND2)
sND2 <- length(seg.sitesND2)
# Total length of all sequences (number of sequences x sequence length) 
# and sequence length
lengthND2total <- length(ZosteropsND2)
seqlengthND2 <- ncol(hapND2)
# Number of sequences calulated two ways, by divdiding the total length 
# by the sequence length (calculation 1) and by extracting 
# the length value from the summary (caclulation 2)
# Confirm both calculations for number of sequences are the same
NumberSequencesND21 <- lengthND2total/seqlengthND2
SummaryND2length <- unlist(strsplit(summaryND2[1,1], split = ":", 
                                    fixed = TRUE))[2]
NumberSequencesND22 <- as.numeric(SummaryND2length)
all.equal(NumberSequencesND21, NumberSequencesND22)
# Calculate theta per sequence, theta per site, theta variance 
# from the Watterson estimator based on number of segregating sites.
ThetaND2perLocusW <- theta.s(sND2, NumberSequencesND21)
ThetaND2perSiteW <- ThetaND2perLocusW/seqlengthND2
ThetaND2perLocusVarW <- theta.s(sND2, NumberSequencesND21, variance = TRUE)
ThetaND2perLocusVarW <- ThetaND2perLocusVarW[2]
# Tests for neutrality including Tajima's D and Ramos-Onsins and Roza's test
tajimaND2 <- tajima.test(ZosteropsND2)
R2ND2 <- R2.test(ZosteropsND2,B = 10000, theta = ThetaND2perLocusW, 
                 plot = TRUE, quiet = FALSE)
```

#### Trees

```{r}
#| label: ND2 trees
#| fig-width: 60
#| fig-height: 55
#UPGMA and NJ trees
dmND2 <- dist.ml(ZosteropsND2)
UPGMAtreeND2 <- upgma(dmND2)
NJtreeND2 <- NJ(dmND2)
plot(UPGMAtreeND2, main = "UPGMA ND2", cex = 3)
plot(NJtreeND2, "unrooted", main = "NJ ND2", cex = 3)
#ML tree midpoint and outgroup rooted trees
mtND2 <- modelTest(ZosteropsND2)
fitND2 <- as.pml(mtND2, "AIC")
fit_mtND2 <- pml_bb(mtND2, control = pml.control(trace = 0))
fit_mtND2
bsND2 <- bootstrap.pml(fit_mtND2, bs = 100, optNni = TRUE, 
                       control = pml.control(trace = 0))
mlND2 <- plotBS(midpoint(fit_mtND2$tree), bsND2, 
                show.node.label = TRUE, p = 50, type="p", 
                main="Maximum likelihood midpoint rooted standard bootstrap", 
                cex = 3)
outgroupND2 <- c("ZAsu002a")
mlND2root <- root(mlND2, outgroup = outgroupND2)
plot(mlND2root, show.node.label = TRUE, p = 50, type="p", 
     main="Maximum Likelihood rooted on outgroup", cex = 3)
```

### BetaGal beta-1,3-N-acetylglucosaminyltransferase 9 (B3GNT9)

#### Load and visualize phased alignment

```{r}
#| label: B3GNT9 load phased alignment
#| fig-width: 30
#| fig-height: 17
#B3GNT
#Load FASTA alignment and check
ZosteropsB3GNT9 <- read.dna("Data/Alignments/B3GNT9_phased.fasta", 
                            format = "fasta")
ZosteropsB3GNT9
checkAlignment(ZosteropsB3GNT9, check.gaps = FALSE, plot = TRUE, what = 1)
```

Graph the Shannon index (H) per base and the number of bases in the alignment per site and show summary.

```{r}
#| label: B3GNT9 Shannon index, number of bases per site, and summary. 
checkAlignment(ZosteropsB3GNT9, check.gaps = FALSE, plot = TRUE, what = 3)
checkAlignment(ZosteropsB3GNT9, check.gaps = FALSE, plot = TRUE, what = 4)
summaryB3GNT9 <- summary(ZosteropsB3GNT9)
```

#### Summary statistics

```{r}
#| label: B3GNT9 summary statistics
# Nucleotide diversity
ndB3GNT9 <- nuc.div(ZosteropsB3GNT9)
# Number of haplotypes from phased DNA sequences (unique sequences) 
# and hapotype diversity
hapB3GNT9 <- haplotype(ZosteropsB3GNT9)
NumberHapsB3GNT9 <- nrow(hapB3GNT9)
hapdivB3GNT9 <- hap.div(ZosteropsB3GNT9)
# An index of segregating sites and number of segregating sites 
seg.sitesB3GNT9 <- seg.sites(ZosteropsB3GNT9)
sB3GNT9 <- length(seg.sitesB3GNT9)
# Total length of all sequences (number of sequences x sequence length) 
# and sequence length
lengthB3GNT9total <- length(ZosteropsB3GNT9)
seqlengthB3GNT9 <- ncol(hapB3GNT9)
# Number of sequences calulated two ways, by divdiding the total length 
# by the sequence length (calculation 1) and by extracting the length value 
# from the summary (caclulation 2)
# Confirm both calculations for number of sequences are the same
NumberSequencesB3GNT91 <- lengthB3GNT9total/seqlengthB3GNT9
SummaryB3GNT9length <- unlist(strsplit(summaryB3GNT9[1,1], 
                                       split = ":", fixed = TRUE))[2]
NumberSequencesB3GNT92 <- as.numeric(SummaryB3GNT9length)
all.equal(NumberSequencesB3GNT91, NumberSequencesB3GNT92)
# Calculate theta per sequence, theta per site, theta variance 
# from the Watterson estimator based on number of segregating sites.
ThetaB3GNT9perLocusW <- theta.s(sB3GNT9, NumberSequencesB3GNT91)
ThetaB3GNT9perSiteW <- ThetaB3GNT9perLocusW/seqlengthB3GNT9
ThetaB3GNT9perLocusVarW <- theta.s(sB3GNT9, NumberSequencesB3GNT91, 
                                   variance = TRUE)
ThetaB3GNT9perLocusVarW <- ThetaB3GNT9perLocusVarW[2]
# Tests for neutrality including Tajima's D and Ramos-Onsins and Roza's test
tajimaB3GNT9 <- tajima.test(ZosteropsB3GNT9)
R2B3GNT9 <- R2.test(ZosteropsB3GNT9,B = 10000, 
                    theta = ThetaB3GNT9perLocusW, plot = TRUE, quiet = FALSE)
```

#### Trees

```{r}
#| label: B3GNT9 trees
#| fig-width: 60
#| fig-height: 55
#UPGMA and NJ trees
dmB3GNT9 <- dist.ml(ZosteropsB3GNT9)
UPGMAtreeB3GNT9 <- upgma(dmB3GNT9)
NJtreeB3GNT9 <- NJ(dmB3GNT9)
plot(UPGMAtreeB3GNT9, main = "UPGMA B3GNT9", cex = 3)
plot(NJtreeB3GNT9, "unrooted", main = "NJ B3GNT9", cex = 3)
#ML tree midpoint and outgroup rooted trees
mtB3GNT9 <- modelTest(ZosteropsB3GNT9)
fitB3GNT9 <- as.pml(mtB3GNT9, "AIC")
fit_mtB3GNT9 <- pml_bb(mtB3GNT9, control = pml.control(trace = 0))
fit_mtB3GNT9
bsB3GNT9 <- bootstrap.pml(fit_mtB3GNT9, bs = 100, optNni = TRUE, 
                          control = pml.control(trace = 0))
mlB3GNT9 <- plotBS(midpoint(fit_mtB3GNT9$tree), bsB3GNT9, 
                   show.node.label = TRUE, p = 50, type="p", 
                   main="Maximum likelihood midpoint rooted standard bootstrap", 
                   cex = 3)
outgroupB3GNT9 <- c("ZAsu001a","ZAsu001b","ZAsu002a", "ZAsu002b")
mlB3GNT9root <- root(mlB3GNT9, outgroup = outgroupB3GNT9)
plot(mlB3GNT9root, show.node.label = TRUE, p = 50, type="p", 
     main="Maximum Likelihood rooted on outgroup", cex = 3)
```

### Baculoviral IAP repeat containing 2 (BIRC2)

#### Load and visualize phased alignment

```{r}
#| label: BIRC2 load phased alignment
#| fig-width: 30
#| fig-height: 17
#BIRC2
#Load FASTA alignment and check
ZosteropsBIRC2 <- read.dna("Data/Alignments/BIRC2_phased.fasta", 
                           format = "fasta")
ZosteropsBIRC2
checkAlignment(ZosteropsBIRC2, check.gaps = FALSE, plot = TRUE, what = 1)
```

Graph the Shannon index (H) per base and the number of bases in the alignment per site and show summary.

```{r}
#| label: BIRC2 Shannon index, number of bases per site, and summary. 
checkAlignment(ZosteropsBIRC2, check.gaps = FALSE, plot = TRUE, what = 3)
checkAlignment(ZosteropsBIRC2, check.gaps = FALSE, plot = TRUE, what = 4)
summaryBIRC2 <- summary(ZosteropsBIRC2)
```

#### Summary statistics

```{r}
#| label: BIRC2 summary statistics
# Nucleotide diversity
ndBIRC2 <- nuc.div(ZosteropsBIRC2)
# Number of haplotypes from phased DNA sequences (unique sequences) 
# and hapotype diversity
hapBIRC2 <- haplotype(ZosteropsBIRC2)
NumberHapsBIRC2 <- nrow(hapBIRC2)
hapdivBIRC2 <- hap.div(ZosteropsBIRC2)
# An index of segregating sites and number of segregating sites 
seg.sitesBIRC2 <- seg.sites(ZosteropsBIRC2)
sBIRC2 <- length(seg.sitesBIRC2)
# Total length of all sequences (number of sequences x sequence length) 
# and sequence length
lengthBIRC2total <- length(ZosteropsBIRC2)
seqlengthBIRC2 <- ncol(hapBIRC2)
# Number of sequences calulated two ways, by divdiding the total length 
# by the sequence length (calculation 1) and by extracting 
# the length value from the summary (caclulation 2)
# Confirm both calculations for number of sequences are the same
NumberSequencesBIRC21 <- lengthBIRC2total/seqlengthBIRC2
SummaryBIRC2length <- unlist(strsplit(summaryBIRC2[1,1], split = ":", 
                                      fixed = TRUE))[2]
NumberSequencesBIRC22 <- as.numeric(SummaryBIRC2length)
all.equal(NumberSequencesBIRC21, NumberSequencesBIRC22)
# Calculate theta per sequence, theta per site, theta variance 
# from the Watterson estimator based on number of segregating sites.
ThetaBIRC2perLocusW <- theta.s(sBIRC2, NumberSequencesBIRC21)
ThetaBIRC2perSiteW <- ThetaBIRC2perLocusW/seqlengthBIRC2
ThetaBIRC2perLocusVarW <- theta.s(sBIRC2, NumberSequencesBIRC21, 
                                  variance = TRUE)
ThetaBIRC2perLocusVarW <- ThetaBIRC2perLocusVarW[2]
# Tests for neutrality including Tajima's D and Ramos-Onsins and Roza's test
tajimaBIRC2 <- tajima.test(ZosteropsBIRC2)
R2BIRC2 <- R2.test(ZosteropsBIRC2,B = 10000, theta = ThetaBIRC2perLocusW, 
                   plot = TRUE, quiet = FALSE)
```

#### Trees

```{r}
#| label: BIRC2 trees
#| fig-width: 60
#| fig-height: 55
#UPGMA and NJ trees
dmBIRC2 <- dist.ml(ZosteropsBIRC2)
UPGMAtreeBIRC2 <- upgma(dmBIRC2)
NJtreeBIRC2 <- NJ(dmBIRC2)
plot(UPGMAtreeBIRC2, main = "UPGMA BIRC2", cex = 3)
plot(NJtreeBIRC2, "unrooted", main = "NJ BIRC2", cex = 3)
#ML tree midpoint and outgroup rooted trees
mtBIRC2 <- modelTest(ZosteropsBIRC2)
fitBIRC2 <- as.pml(mtBIRC2, "AIC")
fit_mtBIRC2 <- pml_bb(mtBIRC2, control = pml.control(trace = 0))
fit_mtBIRC2
bsBIRC2 <- bootstrap.pml(fit_mtBIRC2, bs = 100, optNni = TRUE, 
                         control = pml.control(trace = 0))
mlBIRC2 <- plotBS(midpoint(fit_mtBIRC2$tree), bsBIRC2, 
                  show.node.label = TRUE, p = 50, type="p", 
                  main="Maximum likelihood midpoint rooted standard bootstrap", 
                  cex = 3)
outgroupBIRC2 <- c("ZAsu001a","ZAsu002a")
mlBIRC2root <- root(mlBIRC2, outgroup = outgroupBIRC2)
plot(mlBIRC2root, show.node.label = TRUE, p = 50, type="p", 
     main="Maximum Likelihood rooted on outgroup", cex = 3)
```

### Coiled-coil domain containing 141 (CCDC)

#### Load and visualize phased alignment

```{r}
#| label: CCDC load phased alignment
#| fig-width: 30
#| fig-height: 17
#CCDC
#Load FASTA alignment and check
ZosteropsCCDC <- read.dna("Data/Alignments/CCDC_phased.fasta", 
                          format = "fasta")
ZosteropsCCDC
checkAlignment(ZosteropsCCDC, check.gaps = FALSE, plot = TRUE, what = 1)
```

Graph the Shannon index (H) per base and the number of bases in the alignment per site and show summary.

```{r}
#| label: CCDC Shannon index, number of bases per site, and summary. 
checkAlignment(ZosteropsCCDC, check.gaps = FALSE, plot = TRUE, what = 3)
checkAlignment(ZosteropsCCDC, check.gaps = FALSE, plot = TRUE, what = 4)
summaryCCDC <- summary(ZosteropsCCDC)
```

#### Summary statistics

```{r}
#| label: CCDC summary statistics
# Nucleotide diversity
ndCCDC <- nuc.div(ZosteropsCCDC)
# Number of haplotypes from phased DNA sequences (unique sequences) 
# and hapotype diversity
hapCCDC <- haplotype(ZosteropsCCDC)
NumberHapsCCDC <- nrow(hapCCDC)
hapdivCCDC <- hap.div(ZosteropsCCDC)
# An index of segregating sites and number of segregating sites 
seg.sitesCCDC <- seg.sites(ZosteropsCCDC)
sCCDC <- length(seg.sitesCCDC)
# Total length of all sequences (number of sequences x sequence length) 
# and sequence length
lengthCCDCtotal <- length(ZosteropsCCDC)
seqlengthCCDC <- ncol(hapCCDC)
# Number of sequences calulated two ways, by divdiding 
# the total length by the sequence length (calculation 1) 
# and by extracting the length value from the summary (caclulation 2)
# Confirm both calculations for number of sequences are the same
NumberSequencesCCDC1 <- lengthCCDCtotal/seqlengthCCDC
SummaryCCDClength <- unlist(strsplit(summaryCCDC[1,1], split = ":", 
                                     fixed = TRUE))[2]
NumberSequencesCCDC2 <- as.numeric(SummaryCCDClength)
all.equal(NumberSequencesCCDC1, NumberSequencesCCDC2)
# Calculate theta per sequence, theta per site, theta variance 
# from the Watterson estimator based on number of segregating sites.
ThetaCCDCperLocusW <- theta.s(sCCDC, NumberSequencesCCDC1)
ThetaCCDCperSiteW <- ThetaCCDCperLocusW/seqlengthCCDC
ThetaCCDCperLocusVarW <- theta.s(sCCDC, NumberSequencesCCDC1, variance = TRUE)
ThetaCCDCperLocusVarW <- ThetaCCDCperLocusVarW[2]
# Tests for neutrality including Tajima's D and Ramos-Onsins and Roza's test
tajimaCCDC <- tajima.test(ZosteropsCCDC)
R2CCDC <- R2.test(ZosteropsCCDC,B = 10000, theta = ThetaCCDCperLocusW, 
                  plot = TRUE, quiet = FALSE)
```

#### Trees

```{r}
#| label: CCDC trees
#| fig-width: 60
#| fig-height: 55
#UPGMA and NJ trees
dmCCDC <- dist.ml(ZosteropsCCDC)
UPGMAtreeCCDC <- upgma(dmCCDC)
NJtreeCCDC <- NJ(dmCCDC)
plot(UPGMAtreeCCDC, main = "UPGMA CCDC", cex = 3)
plot(NJtreeCCDC, "unrooted", main = "NJ CCDC", cex = 3)
#ML tree midpoint and outgroup rooted trees
mtCCDC <- modelTest(ZosteropsCCDC)
fitCCDC <- as.pml(mtCCDC, "AIC")
fit_mtCCDC <- pml_bb(mtCCDC, control = pml.control(trace = 0))
fit_mtCCDC
bsCCDC <- bootstrap.pml(fit_mtCCDC, bs = 100, optNni = TRUE, 
                        control = pml.control(trace = 0))
mlCCDC <- plotBS(midpoint(fit_mtCCDC$tree), bsCCDC, 
                 show.node.label = TRUE, p = 50, type="p", 
                 main="Maximum likelihood midpoint rooted standard bootstrap", 
                 cex = 3)
outgroupCCDC <- c("ZAsu001b","ZAsu002b")
mlCCDCroot <- root(mlCCDC, outgroup = outgroupCCDC)
plot(mlCCDCroot, show.node.label = TRUE, p = 50, type="p", 
     main="Maximum Likelihood rooted on outgroup", cex = 3)
```

### G-protein coupled receptor family C group 6 member A (GPRC6)

#### Load and visualize phased alignment

```{r}
#| label: GPRC6 load phased alignment
#| fig-width: 30
#| fig-height: 17
#GPRC6
#Load FASTA alignment and check
ZosteropsGPRC6 <- read.dna("Data/Alignments/GPRC6_phased.fasta", 
                           format = "fasta")
ZosteropsGPRC6
checkAlignment(ZosteropsGPRC6, check.gaps = FALSE, plot = TRUE, what = 1)
```

Graph the Shannon index (H) per base and the number of bases in the alignment per site and show summary.

```{r}
#| label: GPRC6 Shannon index, number of bases per site, and summary. 
checkAlignment(ZosteropsGPRC6, check.gaps = FALSE, plot = TRUE, what = 3)
checkAlignment(ZosteropsGPRC6, check.gaps = FALSE, plot = TRUE, what = 4)
summaryGPRC6 <- summary(ZosteropsGPRC6)
```

#### Summary statistics

```{r}
#| label: GPRC6 summary statistics
# Nucleotide diversity
ndGPRC6 <- nuc.div(ZosteropsGPRC6)
# Number of haplotypes from phased DNA sequences (unique sequences) 
# and hapotype diversity
hapGPRC6 <- haplotype(ZosteropsGPRC6)
NumberHapsGPRC6 <- nrow(hapGPRC6)
hapdivGPRC6 <- hap.div(ZosteropsGPRC6)
# An index of segregating sites and number of segregating sites 
seg.sitesGPRC6 <- seg.sites(ZosteropsGPRC6)
sGPRC6 <- length(seg.sitesGPRC6)
# Total length of all sequences (number of sequences x sequence length) 
# and sequence length
lengthGPRC6total <- length(ZosteropsGPRC6)
seqlengthGPRC6 <- ncol(hapGPRC6)
# Number of sequences calulated two ways, by divdiding the total length 
# by the sequence length (calculation 1) and by extracting 
# the length value from the summary (caclulation 2)
# Confirm both calculations for number of sequences are the same
NumberSequencesGPRC61 <- lengthGPRC6total/seqlengthGPRC6
SummaryGPRC6length <- unlist(strsplit(summaryGPRC6[1,1], 
                                      split = ":", fixed = TRUE))[2]
NumberSequencesGPRC62 <- as.numeric(SummaryGPRC6length)
all.equal(NumberSequencesGPRC61, NumberSequencesGPRC62)
# Calculate theta per sequence, theta per site, theta variance 
# from the Watterson estimator based on number of segregating sites.
ThetaGPRC6perLocusW <- theta.s(sGPRC6, NumberSequencesGPRC61)
ThetaGPRC6perSiteW <- ThetaGPRC6perLocusW/seqlengthGPRC6
ThetaGPRC6perLocusVarW <- theta.s(sGPRC6, NumberSequencesGPRC61, 
                                  variance = TRUE)
ThetaGPRC6perLocusVarW <- ThetaGPRC6perLocusVarW[2]
# Tests for neutrality including Tajima's D and Ramos-Onsins and Roza's test
tajimaGPRC6 <- tajima.test(ZosteropsGPRC6)
R2GPRC6 <- R2.test(ZosteropsGPRC6,B = 10000, 
                   theta = ThetaGPRC6perLocusW, plot = TRUE, quiet = FALSE)
```

#### Trees

```{r}
#| label: GPRC6 trees
#| fig-width: 60
#| fig-height: 55
#UPGMA and NJ trees
dmGPRC6 <- dist.ml(ZosteropsGPRC6)
UPGMAtreeGPRC6 <- upgma(dmGPRC6)
NJtreeGPRC6 <- NJ(dmGPRC6)
plot(UPGMAtreeGPRC6, main = "UPGMA GPRC6", cex = 3)
plot(NJtreeGPRC6, "unrooted", main = "NJ GPRC6", cex = 3)
#ML tree midpoint and outgroup rooted trees
mtGPRC6 <- modelTest(ZosteropsGPRC6)
fitGPRC6 <- as.pml(mtGPRC6, "AIC")
fit_mtGPRC6 <- pml_bb(mtGPRC6, control = pml.control(trace = 0))
fit_mtGPRC6
bsGPRC6 <- bootstrap.pml(fit_mtGPRC6, bs = 100, optNni = TRUE, 
                         control = pml.control(trace = 0))
mlGPRC6 <- plotBS(midpoint(fit_mtGPRC6$tree), bsGPRC6, 
                  show.node.label = TRUE, p = 50, type="p", 
                  main="Maximum likelihood midpoint rooted standard bootstrap", 
                  cex = 3)
outgroupGPRC6 <- c("ZAsu001b","ZAsu002b")
mlGPRC6root <- root(mlGPRC6, outgroup = outgroupGPRC6)
plot(mlGPRC6root, show.node.label = TRUE, p = 50, type="p", 
     main="Maximum Likelihood rooted on outgroup", cex = 3)
```

### Lactamase beta like 1 (LACTBL)

#### Load and visualize phased alignment

```{r}
#| label: LACTBL load phased alignment
#| fig-width: 30
#| fig-height: 17
#LACTBL
#Load FASTA alignment and check
ZosteropsLACTBL <- read.dna("Data/Alignments/LACTBL_phased.fasta", 
                            format = "fasta")
ZosteropsLACTBL
checkAlignment(ZosteropsLACTBL, check.gaps = FALSE, plot = TRUE, what = 1)
```

Graph the Shannon index (H) per base and the number of bases in the alignment per site and show summary.

```{r}
#| label: LACTBL Shannon index, number of bases per site, and summary. 
checkAlignment(ZosteropsLACTBL, check.gaps = FALSE, plot = TRUE, what = 3)
checkAlignment(ZosteropsLACTBL, check.gaps = FALSE, plot = TRUE, what = 4)
summaryLACTBL <- summary(ZosteropsLACTBL)
```

#### Summary statistics

```{r}
#| label: LACTBL summary statistics
# Nucleotide diversity
ndLACTBL <- nuc.div(ZosteropsLACTBL)
# Number of haplotypes from phased DNA sequences (unique sequences) 
# and hapotype diversity
hapLACTBL <- haplotype(ZosteropsLACTBL)
NumberHapsLACTBL <- nrow(hapLACTBL)
hapdivLACTBL <- hap.div(ZosteropsLACTBL)
# An index of segregating sites and number of segregating sites 
seg.sitesLACTBL <- seg.sites(ZosteropsLACTBL)
sLACTBL <- length(seg.sitesLACTBL)
# Total length of all sequences (number of sequences x sequence length) 
# and sequence length
lengthLACTBLtotal <- length(ZosteropsLACTBL)
seqlengthLACTBL <- ncol(hapLACTBL)
# Number of sequences calulated two ways, by divdiding the total length 
# by the sequence length (calculation 1) and by extracting the length value 
# from the summary (caclulation 2)
# Confirm both calculations for number of sequences are the same
NumberSequencesLACTBL1 <- lengthLACTBLtotal/seqlengthLACTBL
SummaryLACTBLlength <- unlist(strsplit(summaryLACTBL[1,1], 
                                       split = ":", fixed = TRUE))[2]
NumberSequencesLACTBL2 <- as.numeric(SummaryLACTBLlength)
all.equal(NumberSequencesLACTBL1, NumberSequencesLACTBL2)
# Calculate theta per sequence, theta per site, theta variance 
# from the Watterson estimator based on number of segregating sites.
ThetaLACTBLperLocusW <- theta.s(sLACTBL, NumberSequencesLACTBL1)
ThetaLACTBLperSiteW <- ThetaLACTBLperLocusW/seqlengthLACTBL
ThetaLACTBLperLocusVarW <- theta.s(sLACTBL, NumberSequencesLACTBL1, 
                                   variance = TRUE)
ThetaLACTBLperLocusVarW <- ThetaLACTBLperLocusVarW[2]
# Tests for neutrality including Tajima's D and Ramos-Onsins and Roza's test
tajimaLACTBL <- tajima.test(ZosteropsLACTBL)
R2LACTBL <- R2.test(ZosteropsLACTBL,B = 10000, 
                    theta = ThetaLACTBLperLocusW, plot = TRUE, quiet = FALSE)
```

#### Trees

```{r}
#| label: LACTBL trees
#| fig-width: 60
#| fig-height: 55
#UPGMA and NJ trees
dmLACTBL <- dist.ml(ZosteropsLACTBL)
UPGMAtreeLACTBL <- upgma(dmLACTBL)
NJtreeLACTBL <- NJ(dmLACTBL)
plot(UPGMAtreeLACTBL, main = "UPGMA LACTBL", cex = 3)
plot(NJtreeLACTBL, "unrooted", main = "NJ LACTBL", cex = 3)
#ML tree midpoint and outgroup rooted trees
mtLACTBL <- modelTest(ZosteropsLACTBL)
fitLACTBL <- as.pml(mtLACTBL, "AIC")
fit_mtLACTBL <- pml_bb(mtLACTBL, control = pml.control(trace = 0))
fit_mtLACTBL
bsLACTBL <- bootstrap.pml(fit_mtLACTBL, bs = 100, optNni = TRUE, 
                          control = pml.control(trace = 0))
mlLACTBL <- plotBS(midpoint(fit_mtLACTBL$tree), bsLACTBL, 
                   show.node.label = TRUE, p = 50, type="p", 
                   main="Maximum likelihood midpoint rooted standard bootstrap", 
                   cex = 3)
outgroupLACTBL <- c("ZAsu002a","ZAsu002b")
mlLACTBLroot <- root(mlLACTBL, outgroup = outgroupLACTBL)
plot(mlLACTBLroot, show.node.label = TRUE, p = 50, type="p", 
     main="Maximum Likelihood rooted on outgroup", cex = 3)
```

### Ubiquitin carboxyl-terminal hydrolase 38 (UBP38)

#### Load and visualize phased alignment

```{r}
#| label: UBP38 load phased alignment
#| fig-width: 30
#| fig-height: 17
#UBP38
#Load FASTA alignment and check
ZosteropsUBP38 <- read.dna("Data/Alignments/UBP38_phased.fasta", 
                           format = "fasta")
ZosteropsUBP38
checkAlignment(ZosteropsUBP38, check.gaps = FALSE, plot = TRUE, what = 1)
```

Graph the Shannon index (H) per base and the number of bases in the alignment per site and show summary.

```{r}
#| label: UBP38 Shannon index, number of bases per site, and summary. 
checkAlignment(ZosteropsUBP38, check.gaps = FALSE, plot = TRUE, what = 3)
checkAlignment(ZosteropsUBP38, check.gaps = FALSE, plot = TRUE, what = 4)
summaryUBP38 <- summary(ZosteropsUBP38)
```

#### Summary statistics

```{r}
#| label: UBP38 summary statistics
# Nucleotide diversity
ndUBP38 <- nuc.div(ZosteropsUBP38)
# Number of haplotypes from phased DNA sequences (unique sequences) 
# and hapotype diversity
hapUBP38 <- haplotype(ZosteropsUBP38)
NumberHapsUBP38 <- nrow(hapUBP38)
hapdivUBP38 <- hap.div(ZosteropsUBP38)
# An index of segregating sites and number of segregating sites 
seg.sitesUBP38 <- seg.sites(ZosteropsUBP38)
sUBP38 <- length(seg.sitesUBP38)
# Total length of all sequences (number of sequences x sequence length) 
# and sequence length
lengthUBP38total <- length(ZosteropsUBP38)
seqlengthUBP38 <- ncol(hapUBP38)
# Number of sequences calulated two ways, by divdiding the total length 
# by the sequence length (calculation 1) and by extracting 
# the length value from the summary (caclulation 2)
# Confirm both calculations for number of sequences are the same
NumberSequencesUBP381 <- lengthUBP38total/seqlengthUBP38
SummaryUBP38length <- unlist(strsplit(summaryUBP38[1,1], split = ":", 
                                      fixed = TRUE))[2]
NumberSequencesUBP382 <- as.numeric(SummaryUBP38length)
all.equal(NumberSequencesUBP381, NumberSequencesUBP382)
# Calculate theta per sequence, theta per site, theta variance 
# from the Watterson estimator based on number of segregating sites.
ThetaUBP38perLocusW <- theta.s(sUBP38, NumberSequencesUBP381)
ThetaUBP38perSiteW <- ThetaUBP38perLocusW/seqlengthUBP38
ThetaUBP38perLocusVarW <- theta.s(sUBP38, NumberSequencesUBP381, 
                                  variance = TRUE)
ThetaUBP38perLocusVarW <- ThetaUBP38perLocusVarW[2]
# Tests for neutrality including Tajima's D and Ramos-Onsins and Roza's test
tajimaUBP38 <- tajima.test(ZosteropsUBP38)
R2UBP38 <- R2.test(ZosteropsUBP38,B = 10000, 
                   theta = ThetaUBP38perLocusW, plot = TRUE, quiet = FALSE)
```

#### Trees

```{r}
#| label: UBP38 trees
#| fig-width: 60
#| fig-height: 55
#UPGMA and NJ trees
dmUBP38 <- dist.ml(ZosteropsUBP38)
UPGMAtreeUBP38 <- upgma(dmUBP38)
NJtreeUBP38 <- NJ(dmUBP38)
plot(UPGMAtreeUBP38, main = "UPGMA UBP38", cex = 3)
plot(NJtreeUBP38, "unrooted", main = "NJ UBP38", cex = 3)
#ML tree midpoint and outgroup rooted trees
mtUBP38 <- modelTest(ZosteropsUBP38)
fitUBP38 <- as.pml(mtUBP38, "AIC")
fit_mtUBP38 <- pml_bb(mtUBP38, control = pml.control(trace = 0))
fit_mtUBP38
bsUBP38 <- bootstrap.pml(fit_mtUBP38, bs = 100, optNni = TRUE, 
                         control = pml.control(trace = 0))
mlUBP38 <- plotBS(midpoint(fit_mtUBP38$tree), bsUBP38, 
                  show.node.label = TRUE, p = 50, type="p", 
                  main="Maximum likelihood midpoint rooted standard bootstrap", 
                  cex = 3)
outgroupUBP38 <- c("ZAsu001a")
mlUBP38root <- root(mlUBP38, outgroup = outgroupUBP38)
plot(mlUBP38root, show.node.label = TRUE, p = 50, type="p", 
     main="Maximum Likelihood rooted on outgroup", cex = 3)
```

## Summary table for all loci

### Load libraries

Create a table that summarizes the population genetic data across all loci. First, load the `gt` and `dplyr` package libraries.

```{r}
#| label: Load package libraries
library(gt)
library(dplyr)
library(webshot2)
```

### Create summary table

Create the summary table from the population genetic parameters estimated for each locus.

```{r}
#| label: All loci summary table
#| fig-width: 40
#| fig-height: 30
#Create table from estimates across all loci.
LocusNames <- c("COI","ND2", "B3GNT9", "BIRC2", 
                "CCDC", "GPRC6", "LACTBL", "UBP38")
LocusType <- c("mt", "mt", "nu", "nu", "nu", "nu", "nu", "nu")
GenBank <- c("PZ392988-PZ393119", "PZ409047-PZ409189", "PZ409190-PZ409276", 
             "PZ409277-PZ409364", "PZ409365-PZ409451", "PZ409452-PZ409538", 
             "PZ409539-PZ409627", "PZ409628-PZ409718")
NumberIndividuals <- c(NumberSequencesCOI1, NumberSequencesND21, 
                       NumberSequencesB3GNT91/2, NumberSequencesBIRC21/2, 
                       NumberSequencesCCDC1/2, NumberSequencesGPRC61/2, 
                       NumberSequencesLACTBL1/2, NumberSequencesUBP381/2)
NumberSequences <- c(NumberSequencesCOI1, NumberSequencesND21, 
                     NumberSequencesB3GNT91, NumberSequencesBIRC21, 
                     NumberSequencesCCDC1, NumberSequencesGPRC61, 
                     NumberSequencesLACTBL1, NumberSequencesUBP381)
SequenceLength <- c(seqlengthCOI, seqlengthND2, seqlengthB3GNT9, 
                    seqlengthBIRC2, seqlengthCCDC, seqlengthGPRC6, 
                    seqlengthLACTBL, seqlengthUBP38)
SegSites <- c(sCOI, sND2, sB3GNT9, sBIRC2, sCCDC, sGPRC6, sLACTBL, sUBP38)
NucleotideDiversity <- c(ndCOI, ndND2, ndB3GNT9, ndBIRC2, 
                         ndCCDC, ndGPRC6, ndLACTBL, ndUBP38)
NucleotideDiversity <- round(NucleotideDiversity, digits = 3)
NumberHaplotypes <- c(NumberHapsCOI, NumberHapsND2, NumberHapsB3GNT9, 
                      NumberHapsBIRC2, NumberHapsCCDC, NumberHapsGPRC6, 
                      NumberHapsLACTBL, NumberHapsUBP38)
HaplotypeDiversity <- c(hapdivCOI, hapdivND2, hapdivB3GNT9, hapdivBIRC2, 
                        hapdivCCDC, hapdivGPRC6, hapdivLACTBL, hapdivUBP38)
HaplotypeDiversity <- round(HaplotypeDiversity, digits = 2)
ThetaWperLocus <- c(ThetaCOIperLocusW, ThetaND2perLocusW, 
                    ThetaB3GNT9perLocusW, ThetaBIRC2perLocusW, 
                    ThetaCCDCperLocusW, ThetaGPRC6perLocusW, 
                    ThetaLACTBLperLocusW, ThetaUBP38perLocusW)
ThetaWperLocus <- round(ThetaWperLocus, digits = 2)
ThetaWvariance <- c(ThetaND2perLocusVarW, ThetaND2perLocusVarW, 
                    ThetaB3GNT9perLocusVarW, ThetaBIRC2perLocusVarW, 
                    ThetaCCDCperLocusVarW, ThetaGPRC6perLocusVarW, 
                    ThetaLACTBLperLocusVarW, ThetaUBP38perLocusVarW)
ThetaWvariance <- round(ThetaWvariance, digits = 2)
ThetaWperSite <- c(ThetaCOIperSiteW, ThetaND2perSiteW, ThetaB3GNT9perSiteW, 
                   ThetaBIRC2perSiteW, ThetaCCDCperSiteW, ThetaGPRC6perSiteW, 
                   ThetaLACTBLperSiteW, ThetaUBP38perSiteW)
ThetaWperSite <- round(ThetaWperSite, digits = 3)
Tajima <- c(tajimaCOI$D,tajimaND2$D, tajimaB3GNT9$D, tajimaBIRC2$D, 
            tajimaCCDC$D, tajimaGPRC6$D, tajimaLACTBL$D, tajimaUBP38$D)
Tajima <- round(Tajima, digits = 2)
TajimaPValue <- c(tajimaCOI$Pval.normal,tajimaND2$Pval.normal, 
                  tajimaB3GNT9$Pval.normal, tajimaBIRC2$Pval.normal, 
                  tajimaCCDC$Pval.normal, tajimaGPRC6$Pval.normal, 
                  tajimaLACTBL$Pval.normal, tajimaUBP38$Pval.normal)
TajimaPValue <- round(TajimaPValue, digits = 3)
R2Test <- c(R2COI$R2, R2ND2$R2, R2B3GNT9$R2, R2BIRC2$R2, R2CCDC$R2, 
            R2GPRC6$R2, R2LACTBL$R2, R2UBP38$R2)
R2Test <- round(R2Test, digits = 2)
R2TestPValue <- c(R2COI$P.val, R2ND2$P.val, R2B3GNT9$P.val, 
                  R2BIRC2$P.val, R2CCDC$P.val, R2GPRC6$P.val, 
                  R2LACTBL$P.val, R2UBP38$P.val)
R2TestPValue <- round(R2TestPValue, digits = 3)
AllLociSummary <- data.frame(LocusNames, LocusType, GenBank, NumberIndividuals, 
                             NumberSequences, SequenceLength, SegSites, 
                             NucleotideDiversity, NumberHaplotypes, 
                             HaplotypeDiversity, ThetaWperLocus, 
                             ThetaWvariance, ThetaWperSite, Tajima, 
                             TajimaPValue, R2Test, R2TestPValue)
colnames(AllLociSummary) <- c("Locus", "Locus type", "GenBank", "Individuals", 
                              "Number of sequences", "Length", 
                              "Segregating sites", "\u03C0", 
                              "Number of haplotypes", "Haplotype diversity", 
                              "\u03B8 per locus", "\u03B8 per locus variance", 
                              "\u03B8 per site", "Tajima's D", 
                              "Tajima's D p-value", "R2 test", 
                              "R2 test p-value")
AllLociSummary
# AllLociSummary %>%
AllLociSummaryTable <- gt(AllLociSummary, auto_align = FALSE) %>%
  tab_footnote(footnote = md("Locus type denoted as mitochondrial (mt) 
                             or nuclear protein coding (nu). Sequential 
                             GenBank accession numbers denoted for each 
                             locus. For phased nuclear DNA alignments the 
                             number of sequences is 2x the number of 
                             individuals sequenced. \u03C0 is per site 
                             nucleotide diversity. Haplotype diversity 
                             calculated according to Nei and Tajima (1981). 
                             \u03B8 per locus and per locus variance in 
                             \u03B8 are estimated from the number of 
                             segregating sites (Watterson estimator). 
                             \u03B8 per site is calculated by dividing the 
                             \u03B8 per locus by the sequence length. Tajima's 
                             D and the Ramos-Onsins and Rozas (R2) tests are 
                             used to test for neutralilty and population growth, 
                             repsectively, at each locus. Population genetic 
                             parameters for each locus were estimated in 
                             the pegas and ape R packages"))
AllLociSummaryTable
gtsave(AllLociSummaryTable, filename = "SangerSeq_AllLociSummaryTable.png", 
       vheight = 600, vwidth = 1300)
```

## Close log

Close the log file.

```{r}
#| label: Close log file
log_close()
```
