---
title: "Zosterops_StarBEAST2_extra"
---

### Extra code showing how to add branch support using ape

Don't add this to the yml fie and index. It is just saving some extra code for possible use later.

This is an ancillary bit of code showing how to use ape to add branch support values to a tree. It is pretty tedious. There must be a better way to do this. Surely ape can read support values from a newick formatted tree?

Likely can use this code to make a nicer looking tree that could be used as a figure in the manuscript?

This example is from the \*BEAST2 consensus species tree from the A runs series of 20 replicate runs using all loci and the `Zosterops taxon sets A.txt` taxon set file. Tree edited in FigTree with posterior probability displayed at the nodes and an axis in Ma.

```{r}
#| label: Load libraries
library(logr)
log_open(file_name = "ZosteropsStarBEAST2trees.log")
library(ape)
```

Load nexus tree

```{r}
#| label: Read and plot species tree A
#| fig-width: 30
#| fig-height: 17
# Load species tree
ZosteropsSpeciestreeA <- read.tree(file = "Data/BEAST trees/ZosteropsBEAST2_A.tre")
# Root the tree
ZosteropsSpeciestreeA <- root(ZosteropsSpeciestreeA, outgroup = "'Zosterops atrifrons surdus'", resolve.root = TRUE)
# Remove quotes from tip labels
ZosteropsSpeciestreeA$tip.label <- gsub("'", "", ZosteropsSpeciestreeA$tip.label)
# Count the number of internal nodes on the tree and replace node labels with posterior support when support is greater than 50. * 50 - 90. ** 90 - 100
Nnode(ZosteropsSpeciestreeA)
getMRCA(ZosteropsSpeciestreeA, c("Zosterops nigrorum nigrorum PANAY", "Zosterops nigrorum luzonicus LUZON"))
ZosteropsSpeciestreeA$node.label[103] <- c("*")
ZosteropsSpeciestreeA$node.label[104] <- c("*")
ZosteropsSpeciestreeA$node.label[105] <- c("*")
getMRCA(ZosteropsSpeciestreeA, c("Zosterops simplex TAIWAN", "Zosterops simplex SINGAPORE"))
ZosteropsSpeciestreeA$node.label[99] <- c("**")
ZosteropsSpeciestreeA$node.label[100] <- c("**")
getMRCA(ZosteropsSpeciestreeA, c("Zosterops nigrorum nigrorum PANAY", "Zosterops everetti basilanicus"))
ZosteropsSpeciestreeA$node.label[55] <- c("*")
getMRCA(ZosteropsSpeciestreeA, c("Zosterops montanus montanus SULAWESI", "Zosterops everetti basilanicus"))
ZosteropsSpeciestreeA$node.label[57] <- c("*")
getMRCA(ZosteropsSpeciestreeA, c("Zosterops montanus vulcani MINDANAO", "Zosterops everetti basilanicus"))
ZosteropsSpeciestreeA$node.label[59] <- c("*")
ZosteropsSpeciestreeA$node.label[60] <- c("*")
getMRCA(ZosteropsSpeciestreeA, c("Zosterops meyeni LANYU", "Zosterops meyeni BATANES"))
ZosteropsSpeciestreeA$node.label[96] <- c("*")
getMRCA(ZosteropsSpeciestreeA, c("Zosterops montanus whiteheadi LUZON", "Zosterops montanus halconensis MINDORO"))
ZosteropsSpeciestreeA$node.label[91] <- c("**")
# Plot species tree
plot(ZosteropsSpeciestreeA,
     main = "*BEAST2 Zosterops tree A", 
     cex = 1.35,  # Adjust text size
     label.offset = 0.01,
     show.node.label = TRUE)
```

Load newick format tree.

```{r}
#| label: Read and plot species tree B
#| fig-width: 30
#| fig-height: 17
# Load species tree
ZosteropsSpeciestreeB <- read.tree(file = "Data/BEAST trees/ZosteropsBEAST2_B.tre")
# Root the tree
ZosteropsSpeciestreeB <- root(ZosteropsSpeciestreeB, outgroup = "'Zosterops atrifrons surdus'", resolve.root = TRUE)
# Remove quotes from tip labels
ZosteropsSpeciestreeB$tip.label <- gsub("'", "", ZosteropsSpeciestreeB$tip.label)
# Count the number of internal nodes on the tree and replace node labels with posterior support when support is greater than 50. * 50 - 90. ** 90 - 100
Nnode(ZosteropsSpeciestreeB)
getMRCA(ZosteropsSpeciestreeB, c("Zosterops nigrorum nigrorum PANAY", "Zosterops nigrorum luzonicus LUZON"))
ZosteropsSpeciestreeB$node.label[96] <- c("**")
ZosteropsSpeciestreeB$node.label[97] <- c("*")
ZosteropsSpeciestreeB$node.label[98] <- c("**")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops nigrorum nigrorum PANAY", "Zosterops atrifrons surdus"))
ZosteropsSpeciestreeB$node.label[51] <- c("*")
ZosteropsSpeciestreeB$node.label[52] <- c("*")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops meyeni LANYU", "Zosterops everetti basilanicus"))
ZosteropsSpeciestreeB$node.label[57] <- c("**")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops meyeni LANYU", "Zosterops meyeni BATANES"))
ZosteropsSpeciestreeB$node.label[95] <- c("**")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops meyeni LANYU", "Zosterops montanus diuatae MINDANAO"))
ZosteropsSpeciestreeB$node.label[68] <- c("*")
ZosteropsSpeciestreeB$node.label[69] <- c("**")
ZosteropsSpeciestreeB$node.label[70] <- c("**")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops japonicus stejnegeri MIYAKE", "Zosterops japonicus alani ANEJIMA"))
ZosteropsSpeciestreeB$node.label[72] <- c("*")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops japonicus stejnegeri MIYAKE", "Zosterops japonicus stejnegeri NIIJIMA"))
ZosteropsSpeciestreeB$node.label[77] <- c("*")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops japonicus japonicus JEJU", "Zosterops japonicus japonicus TSUSHIMA"))
ZosteropsSpeciestreeB$node.label[86] <- c("*")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops japonicus loochoensis IRIOMOTE", "Zosterops japonicus loochoensis YONAGUNI"))
ZosteropsSpeciestreeB$node.label[93] <- c("*")
ZosteropsSpeciestreeB$node.label[92] <- c("*")
ZosteropsSpeciestreeB$node.label[91] <- c("*")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops montanus parkesi PALAWAN", "Zosterops montanus halconensis MINDORO"))
ZosteropsSpeciestreeB$node.label[66] <- c("**")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops montanus whiteheadi LUZON", "Zosterops montanus halconensis MINDORO"))
ZosteropsSpeciestreeB$node.label[67] <- c("**")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops montanus PANAY", "Zosterops montanus pectoralis NEGROS"))
ZosteropsSpeciestreeB$node.label[62] <- c("*")
ZosteropsSpeciestreeB$node.label[63] <- c("**")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops montanus montanus SULAWESI", "Zosterops montanus montanus JAVA"))
ZosteropsSpeciestreeB$node.label[65] <- c("*")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops montanus vulcani MINDANAO", "Zosterops everetti basilanicus"))
ZosteropsSpeciestreeB$node.label[61] <- c("**")
getMRCA(ZosteropsSpeciestreeB, c("Zosterops simplex GUANGDONG", "Zosterops simplex SINGAPORE"))
ZosteropsSpeciestreeB$node.label[53] <- c("**")
ZosteropsSpeciestreeB$node.label[54] <- c("**")
ZosteropsSpeciestreeB$node.label[55] <- c("*")
ZosteropsSpeciestreeB$node.label[56] <- c("*")
# Plot species tree
plot(ZosteropsSpeciestreeB,
     main = "*BEAST2 Zosterops tree B", 
     cex = 1.35,  # Adjust text size
     label.offset = 0.01,
     show.node.label = TRUE)  # Adjust label offset)
```
