---
title: "Genetic differentiation and biogeographic breaks"

editor: visual
knitr: true
bibliography: ZosteropsDistanceFst.bib
---

## Goals

What follows is our workflow for estimating genetic distance and differentiation among East Asian and Philippine Zosterops populations and across biogeographic breaks based on single nucleotide polymorphism (SNP) data from our double digest restriction-site associated DNA (ddRAD) library.

## Getting started

We are using R version 4.4.2, RStudio version 2024.09.1 for the analysis. Quarto version 1.6.39 for Mac OS was used to document the workflow. Set the working directory in RStudio to one where the `vcf` file for the SNPs and a text file defining the population for each sample are available.

## General methods summary

The following analyses used the SNP dataset converted to the genlight format. Pairwise fixed differences were measured from the SNP dataset in R using the `gl.fixed.diff` function in the *dartR* v2 package. Samples were grouped according to a combination of species and geography with *Z. japonicus* split into an Ogasawara and Volcano Islands group containing *Z. japonicus alani* and a ‘continental’ island group containing the remaining subspecies. *Z. montanus* was divided into four groups: the nominate subspecies in Indonesia, a northern Philippines group, a central/southern Philippines group, and the *Z. montanus parkesi* subspecies on Palawan. In every tree produced, and the PCA and DAPC analyses, our single sample from *Z. nigrorum catarmanensis* from the island of Camiguin South (Zni14341) fell within *Z. montanus* with consistent and strong support and was considered a central/southern Philippine *Z. montanus* for these analyses. The remaining samples were assigned according to species.

Pairwise F~ST~ was determined using the `stamppFst` function and pairwise Nei’s distances were measured using the `stamppNeisD` function in *StAMPP* and displayed as heatmaps using the R package *ggplot2*.  Additional population grouping was done for these analyses with *Z. japonicus* divided into populations in the northern, central, and southern Ryukyu Islands, the Korean Peninsula and adjacent islands, the main islands of Japan, the Ogasawara and Volcano Islands, the Izu Islands, and the introduced population in Hawaii. *Z. montanus* was divided into the Indonesian, northern Philippine, Visayas, Mindanao, and Palawan populations. *Z. simplex* was divided into those samples from the native range in Asia and introduced populations in North America. The rest of the samples were grouped only by species.

### Packages

Set the working directory to one where the SNP data file `zost.filtered.snps.vcf` and the taxon/population designation file `ZosteropsPOPS_distance.txt` are both accessible. Conducting the distance and Fst analysis on these data and generating the associated figures and maps requires the following R packages.

- logr [@logr] Optional if you wish to create a log file for the analysis.

- vcfR [@vcfR]

- SNPfiltR [@SNPfiltR]

- ggplot2 [@ggplot2]

- adegenet [@adegenet]

- StAMPP [@StAMPP]

- maps [@maps]

- sf [@sf]

- dartR.base [@dartR.base]

Restart R in the Session/Restart R drop-down in R Studio and create a `log` file for the analysis (optional).

```{r}
#| label: Create a log file.
library(logr)
log_open(file_name = "ZosteropsDistanceFST.log")
```

After making sure they are installed load the libraries for each required package.

```{r}
#| label: Load packages
library(vcfR)
library(SNPfiltR)
library(ggplot2)
library(adegenet)
library(StAMPP)
library(maps)
library(sf)
```

### Data files

This analysis requires the filtered variant call format (VCF) file titled `zost.filtered.snps.vcf`. Additionally the population of origin and/or taxon assignment will need to be defined for each sample. Taxon and population data for this analysis is in the file titled `ZosteropsPOPS_distance.txt`. In this file populations are defined in reference to the following combinations of (mostly) allopatric taxa and biogeographic regions. A second taxon assignment file was created assigning individual samples to mostly species designation or more broad allopatric groups within species. That file is titled `ZosteropsSPECIES.txt`. A detailed description of the assignments in each file is shown below. For discussions of biogeographic breaks in Japan and the Ryukyu Islands see Komaki 2021 [@komaki2021] and McKay 2012 [@mckay2011].

#### Population (island and island group) assignments.

Assignment file `ZosteropsPOPS_distance.txt`

- *Zosterops japonicus*

  - `japonicusKOREA` The Korean Peninsula and adjacent islands of the South China Sea and Korean and Tsushima Straits.

  - `japonicusJAPAN` The main islands of the Japanese archipelago including Hokkaido, Honshu, Shikoku, and Kyushu.

  - `japonicusRYUKYUn` The northern Ryukyu islands of the Osumi group including Yakushima and Tanegashima. Populations with this designation are those in the broader Ryukyu archipelago but north of the biogeographic break known as Watse's Line.

  - `japonicusRYUKYUc` The central Ryukyu islands including Amami and Okinawa and their adjacent islands.

  - `japonicusRYUKYUs` The southern-most Ryukyu islands of the Sakishima group including the Miyako Islands, Yaeyama Islands, and Yonaguni island.

  - `japonicusIZU` Populations of the *Z. japonicus stejnegeri* subspecies from the Izu Islands (including Niijima, Aogashima, Kozushima, and Miyake).

  - `japonicusOGASAWARA` Populations of the *Z. japonicus alani* subspecies in the Ogasawara (including Chichijima, Hahajima, Anejima, and Imotojima islands) and Volcano islands (Iwoto).

  - `japonicusHAWAII` Introduced population in the Hawaiian archipelago.

- *Zosterops meyeni*

  - `meyeniLANYU` Lanyu island off the east coast of Taiwan.

  - `meyeniBATAN` The Batan Islands of the Philippines including Batan and Sabtang islands.

- *Zosterops montanus*

  - `montanusNORTH` The Northern Philippine islands of Luzon and Mindoro.

  - `montanusVISAYAS` The Visayas region of the central Philippines including the main islands of Negros, Panay, Cebu, Bohol, Leyte, and Samar.

  - `montanusMINDANAO` The southern most islands of the Philippines including the island of Mindanao. Included with *Z. montanus* samples from Mindanao in this analysis is a *Z. nigrorum* sample (`Zni14341`). This sample originating from the island of Camiguin South off the northern coast of Mindanao was initially assigned to the *Z. nigrorum* subspecies catarmanensis but in this study was found to cluster strongly with the central and southern Philippine populations of *Z. montanus* in both Sanger sequencing and SNP datasets.

  - `montanusPALAWAN` Philippine *Z. montanus* population on the island of Palawan.

  - `montanusINDONESIA` *Z. montanus* populations of the Indonesian archipelago.

- *Zosterops simplex*

  - `simplexASIA` All *Z. simplex* populations on either the island of Taiwan or on mainland continental Asia.

  - `simplexNA` Introduced *Z. simplex* populations in the southern California region of North America.

- Additional outgroup taxa

  - `erythropleurus` *Z. erythropleurus* including samples originating in the breeding range in the Russian Far East and wintering individuals in Southeast Asia.

  - `everetti` *Z. everetti* in the Visayas and Southern Philippines.

  - `nigrorum` *Z. nigrorum* in the Visayas and Northern Philippines.

  - `palpebrosus` *Z. palpebrosus* samples from Vietnam.

Below are the taxon and population assignments in the file `ZosteropsPOPS_distance.txt` with the column headers `ID` and `POP`.

```         
ID  POP
ZJlo012 japonicusRYUKYUc
ZJlo015 japonicusRYUKYUc
ZJlo016 japonicusRYUKYUs
ZJlo017 japonicusRYUKYUs
ZJlo021 japonicusRYUKYUc
ZJlo022 japonicusRYUKYUc
ZJlo024 japonicusRYUKYUs
ZJlo031 japonicusRYUKYUc
ZJlo045 japonicusRYUKYUc
ZJlo046 japonicusRYUKYUs
ZJlo050 japonicusRYUKYUc
ZJlo052 japonicusRYUKYUc
ZJlo055 japonicusRYUKYUc
ZJja001 japonicusKOREA
ZJja002 japonicusKOREA
ZJja003 japonicusKOREA
ZJja004 japonicusKOREA
ZJja005 japonicusKOREA
ZJja009 japonicusKOREA
ZJja010 japonicusKOREA
ZJja012 japonicusJAPAN
ZJja013 japonicusJAPAN
ZJja014 japonicusJAPAN
ZJja016 japonicusJAPAN
ZJja030 japonicusKOREA
ZJal003 japonicusOGASAWARA
Zjal004 japonicusOGASAWARA
ZJal005 japonicusOGASAWARA
ZJal010 japonicusOGASAWARA
ZJal011 japonicusOGASAWARA
ZJal012 japonicusOGASAWARA
ZJal020 japonicusOGASAWARA
ZJst007 japonicusIZU
ZJst009 japonicusIZU
ZJst010 japonicusIZU
ZJst012 japonicusIZU
ZJst014 japonicusIZU
ZJst015 japonicusIZU
ZJin001 japonicusRYUKYUn
ZJin003 japonicusRYUKYUn
ZJsi002 simplexASIA
ZJsi017 simplexASIA
ZJsi032 simplexASIA
ZJsi023 simplexASIA
ZJsi024 simplexASIA
ZJsi027 simplexASIA
ZJsi028 simplexASIA
ZJsi029 simplexASIA
ZJsi031 simplexASIA
ZPxx001 simplexASIA
ZPxx002 simplexASIA
ZMxx002 meyeniBATAN
ZMxx003 meyeniLANYU
ZMxx004 meyeniLANYU
ZMxx005 meyeniLANYU
ZMxx006 meyeniLANYU
ZMOxx001    montanusVISAYAS
ZMOxx002    montanusVISAYAS
ZMOxx003    montanusVISAYAS
ZMOxx005    montanusVISAYAS
ZMOha001    montanusNORTH
ZMOwh002    montanusNORTH
ZMOwh003    montanusNORTH
ZMOwh004    montanusNORTH
ZMOwh005    montanusNORTH
ZMOwh006    montanusNORTH
ZMOvu002    montanusMINDANAO
ZMOvu003    montanusMINDANAO
ZMOvu004    montanusMINDANAO
ZMOvu005    montanusMINDANAO
ZMOmo001    montanusINDONESIA
ZMOpa001    montanusPALAWAN
ZMOpa002    montanusPALAWAN
ZNni001 nigrorum
ZERxx002    erythropleurus
ZERxx004    erythropleurus
ZERxx005    erythropleurus
Zsim23588   simplexASIA
Zsim28142   simplexASIA
Zsim30897   palpebrosus
Zpal23498   palpebrosus
Zpal23522   palpebrosus
Zmon20893   montanusVISAYAS
Zmon20892   montanusVISAYAS
Zmon20902   montanusVISAYAS
Zmon20909   montanusVISAYAS
Zsim31166   montanusINDONESIA
Zsim31159   montanusINDONESIA
Zmey17876   meyeniBATAN
Zmey17877   meyeniBATAN
Zmey17852   meyeniBATAN
Zmey17922   meyeniBATAN
Zmey17920   meyeniBATAN 
Zmey17925   meyeniBATAN
Zmey17923   meyeniBATAN
Zery28090   erythropleurus
Zery28091   erythropleurus
Zery28088   erythropleurus
Zni10863    nigrorum
Zni14341    montanusMINDANAO    
Zni19650    nigrorum
Zni17984    nigrorum
ZjaDOT-10981    palpebrosus
ZjaDOT-5235 simplexASIA
Z_LA_122866_2   simplexNA
Z_LA_122577_2   simplexNA
Z_LA_122188_2   simplexNA
Zsim13809   simplexASIA
Zsim13773   simplexASIA
Zsim6797    simplexASIA
Zsim10336   simplexASIA
Zsim11362   simplexASIA
Zsim11102   simplexASIA
Zsim11220   simplexASIA
Zmon20899   montanusVISAYAS
Zmon28375   montanusMINDANAO
Zev13949    everetti
Zev31650    everetti
Zev28451    everetti
Z_HI_BRY431 japonicusHAWAII
Z_HI_NAN290 japonicusHAWAII
Z_HI_WAI087 japonicusHAWAII
Z_HI_WAI078 japonicusHAWAII
Z_HI_SOL783 japonicusHAWAII
```

#### Species assignments.

Assignment file `ZosteropsSPECIES.txt`

- *Zosterops japonicus*

  - `japonicus` All populations of *Z. japonicus* in the main islands of Japan, Korean Peninsula, Ryukyu Islands, and Izu islands to Chichijima in the Ogasawara islands and the introduced *Z. japonicus* population in Hawaii.

  - `japonicusOGASAWARA` Populations of the *Z. japonicus alani* subspecies in the Ogasawara (including Chichijima, Hahajima, Anejima, and Imotojima islands) and Volcano islands (Iwoto) are the most divergent within the *Z. japonicus* taxon and are for exploratory purposes considered a separate taxon.

- *Zosterops meyeni*

  - `meyeni` Populations on Lanyu island off the east coast fo Taiwan and the Batan Islands of the Philippines, including Batan and Sabtang islands.

- *Zosterops montanus*

  - `montanusNORTH` The Northern Philippine islands of Luzon and Mindoro.

  - `montanusSOUTH` Philippine *Z. montanus* populations in the Visayas region of the central Philippines including the main islands of Negros, Panay, Cebu, Bohol, Leyte, and Samar and the southern most islands of the Philippines including the island of Mindanao. Included with *Z. montanus* samples from Mindanao in this analysis is a *Z. nigrorum* sample (`Zni14341`), see description in the previous section above.

  - `montanusPALAWAN` Philippine *Z. montanus* population on the island of Palawan.

  - `montanusINDONESIA` *Z. montanus* populations of the Indonesian archipelago.

- *Zosterops simplex*

  - `simplex` All *Z. simplex* populations on either the island of Taiwan or on mainland continental Asia and introduced *Z. simplex* populations in the southern California region of North America.

- Additional outgroup taxa

  - `erythropleurus` *Z. erythropleurus* including samples originating in the breeding range in the Russian Far East and wintering individuals in Southeast Asia.

  - `everetti` *Z. everetti* in the Visayas and Southern Philippines.

  - `nigrorum` *Z. nigrorum* in the Visayas and Northern Philippines.

  - `palpebrosus` *Z. palpebrosus* samples from Vietnam.

Below are the species assignments in the file `ZosteropsSPECIES.txt` with the column headers `ID` and `POP`.

```         
ID  POP
ZJlo012 japonicus
ZJlo015 japonicus
ZJlo016 japonicus
ZJlo017 japonicus
ZJlo021 japonicus
ZJlo022 japonicus
ZJlo024 japonicus
ZJlo031 japonicus
ZJlo045 japonicus
ZJlo046 japonicus
ZJlo050 japonicus
ZJlo052 japonicus
ZJlo055 japonicus
ZJja001 japonicus
ZJja002 japonicus
ZJja003 japonicus
ZJja004 japonicus
ZJja005 japonicus
ZJja009 japonicus
ZJja010 japonicus
ZJja012 japonicus
ZJja013 japonicus
ZJja014 japonicus
ZJja016 japonicus
ZJja030 japonicus
ZJal003 japonicusOGASAWARA
Zjal004 japonicusOGASAWARA
ZJal005 japonicusOGASAWARA
ZJal010 japonicusOGASAWARA
ZJal011 japonicusOGASAWARA
ZJal012 japonicusOGASAWARA
ZJal020 japonicusOGASAWARA
ZJst007 japonicus
ZJst009 japonicus
ZJst010 japonicus
ZJst012 japonicus
ZJst014 japonicus
ZJst015 japonicus
ZJin001 japonicus
ZJin003 japonicus
ZJsi002 simplex
ZJsi017 simplex
ZJsi032 simplex
ZJsi023 simplex
ZJsi024 simplex
ZJsi027 simplex
ZJsi028 simplex
ZJsi029 simplex
ZJsi031 simplex
ZPxx001 simplex
ZPxx002 simplex
ZMxx002 meyeni
ZMxx003 meyeni
ZMxx004 meyeni
ZMxx005 meyeni
ZMxx006 meyeni
ZMOxx001    montanusSOUTH
ZMOxx002    montanusSOUTH
ZMOxx003    montanusSOUTH
ZMOxx005    montanusSOUTH
ZMOha001    montanusNORTH
ZMOwh002    montanusNORTH
ZMOwh003    montanusNORTH
ZMOwh004    montanusNORTH
ZMOwh005    montanusNORTH
ZMOwh006    montanusNORTH
ZMOvu002    montanusSOUTH
ZMOvu003    montanusSOUTH
ZMOvu004    montanusSOUTH
ZMOvu005    montanusSOUTH
ZMOmo001    montanusINDONESIA
ZMOpa001    montanusPALAWAN
ZMOpa002    montanusPALAWAN
ZNni001 nigrorum
ZERxx002    erythropleurus
ZERxx004    erythropleurus
ZERxx005    erythropleurus
Zsim23588   simplex
Zsim28142   simplex
Zsim30897   palpebrosus
Zpal23498   palpebrosus
Zpal23522   palpebrosus
Zmon20893   montanusSOUTH
Zmon20892   montanusSOUTH
Zmon20902   montanusSOUTH
Zmon20909   montanusSOUTH
Zsim31166   montanusINDONESIA
Zsim31159   montanusINDONESIA
Zmey17876   meyeni
Zmey17877   meyeni
Zmey17852   meyeni
Zmey17922   meyeni
Zmey17920   meyeni  
Zmey17925   meyeni
Zmey17923   meyeni
Zery28090   erythropleurus
Zery28091   erythropleurus
Zery28088   erythropleurus
Zni10863    nigrorum
Zni14341    montanusSOUTH   
Zni19650    nigrorum
Zni17984    nigrorum
ZjaDOT-10981    palpebrosus
ZjaDOT-5235 simplex
Z_LA_122866_2   simplex
Z_LA_122577_2   simplex
Z_LA_122188_2   simplex
Zsim13809   simplex
Zsim13773   simplex
Zsim6797    simplex
Zsim10336   simplex
Zsim11362   simplex
Zsim11102   simplex
Zsim11220   simplex
Zmon20899   montanusSOUTH
Zmon28375   montanusSOUTH
Zev13949    everetti
Zev31650    everetti
Zev28451    everetti
Z_HI_BRY431 japonicus
Z_HI_NAN290 japonicus
Z_HI_WAI087 japonicus
Z_HI_WAI078 japonicus
Z_HI_SOL783 japonicus
```

### VCF file

Load the SNP data from the `vcf` file. Recall the `ZosteropsVCF` object to summarize the dataset. The `ZosteropsVCF` object should have 124 samples and 15,704 variants.

```{r}
#| label: Load and summarize the SNP data
ZosteropsVCF <- read.vcfR("Data/VCF files/zost.filtered.snps.vcf")
ZosteropsVCF
```

### Taxon and population assignments

#### Population (island and island group) assignments

Assign the taxon and population designations to the individual samples in the SNP dataset from the `ZosteropsPOPS_distance.txt` file. Confirm that the IDs in the taxon/population file match the sample IDs in the VCF file.

```{r}
#| label: Assign taxon and population designations to individual samples
sample.info <- read.table("Data/Taxon files/ZosteropsPOPS_distance.txt", sep = "\t", header = TRUE)
sample.info <- sample.info[sample.info$ID %in% colnames(ZosteropsVCF@gt),]
table(sample.info$POP)
colnames(ZosteropsVCF@gt)
sample.info$ID == colnames(ZosteropsVCF@gt) [-1]
rownames(sample.info) <- c(1:nrow(sample.info))
```

#### Species assignments

Create another set of assignments for mostly species level designations to the individual samples in the SNP dataset from the `ZosteropsSPECIES.txt` file. Confirm that the IDs in the taxon/population file match the sample IDs in the VCF file.

```{r}
#| label: Assign species designations to individual samples
sample.species.info <- read.table("Data/Taxon files/ZosteropsSPECIES.txt", sep = "\t", header = TRUE) 
sample.species.info <- sample.species.info[sample.species.info$ID %in% colnames(ZosteropsVCF@gt),]
table(sample.species.info$POP) 
colnames(ZosteropsVCF@gt) 
sample.species.info$ID == colnames(ZosteropsVCF@gt) [-1] 
rownames(sample.species.info) <- c(1:nrow(sample.species.info))
```

### Convert data to Genlight format

Convert the SNP dataset `ZosteropsVCF` object into a genlight object required for estimating distance and Fst in StAMMP and assign the correct taxa and populations to that object from the `ZosteropsPOPS_distance.txt` file (`sample.info`).

```{r}
#| label: Convert to a genlight object and assign taxa and populations for distance and Fst
ZosteropsGEN <- vcfR2genlight(ZosteropsVCF)
ZosteropsGEN@pop <- as.factor(sample.info$POP)
```

Convert the SNP dataset `ZosteropsVCF` object into a genlight object and assign the correct species to that object from the `ZosteropsSPECIES.txt` assignments (`sample.species.info`). This genlight object (`ZosteropsGENfdiff`) will be used for an additional analysis of fixed differences at the species level.

```{r}
#| label: Convert to a genlight object and assign species for fixed differences analysis
ZosteropsGENfdiff <- vcfR2genlight(ZosteropsVCF)
ZosteropsGENfdiff@pop <- as.factor(sample.species.info$POP)
```

## Fixed differences

Conduct a compliance check using the `gl.compliance.check` function in `dartR` on the `ZosteropsGENfdiff` genlight object. Use the `gl.fixed.diff` function to then count pairwise fixed SNP differences between the assigned species groups. Conduct the pairwise fixed differences analysis under the default parameters including `alpha = 0.05` and a threshold value of 0.02 (`delta = 0.02`). Because these are the default parameters they do not appear as delineated in the command. Include in the fixed differences analysis a statistical test to estimate p-values associated with each pairwise comparison (`test = TRUE`).

```{r}
#| label: Estimate fixed differences among groups
library(dartR.base)
ZosteropsGENfdiff <- gl.compliance.check(ZosteropsGENfdiff)
gl.fixed.diff(ZosteropsGENfdiff, test = TRUE)
```

### Estimate Fst and Nei's distance

Generate the Fst and Nei's distance estimates from the Zosterops genlight object (`ZosteropsGEN`).

```{r}
#| label: Fst and distance estimates
ZosteropsFST <- stamppFst(ZosteropsGEN, nboots = 100, percent = 95, nclusters = 4)
ZosteropsFSTmatrix <- ZosteropsFST$Fsts
ZosteropsFSTmatrix[upper.tri(ZosteropsFSTmatrix)] <- t(ZosteropsFSTmatrix)[upper.tri(ZosteropsFSTmatrix)]
ZosteropsFSTheat <- reshape::melt(ZosteropsFSTmatrix)
ZosteropsNeiDmatrix <- stamppNeisD(ZosteropsGEN, pop = TRUE)
ZosteropsNeiDmatrix[upper.tri(ZosteropsNeiDmatrix)] <- t(ZosteropsNeiDmatrix)[upper.tri(ZosteropsNeiDmatrix)]
ZosteropsNeiDheat <- reshape::melt(ZosteropsNeiDmatrix)
```

Examine the results of the Fst analysis by calling the object `ZosteropsFST`.

```{r}
#| label: Fst results
ZosteropsFST
```

Examine the results of the Nei's distance matrix by calling the object `ZosteropsNeiDmatrix`.

```{r}
#| label: Nei's distance results
# Show Nei's distance results
ZosteropsNeiDmatrix
```

### Pairwise Fst heatmap

Create a heatmap of the pairwise Fst estimates for each combination of taxa/populations.

```{r}
#| label: Fst heatmap
#| fig-height: 10
#| fig-width: 20
ggplot(data = ZosteropsFSTheat, aes(x = X1, y = X2, fill = value))+
  geom_tile()+
  geom_text(data = ZosteropsFSTheat, aes(label = round(value, 2)))+
  theme_minimal()+
  scale_fill_gradient2(low = "white", high = "red", space = "Lab", name = "Fst")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.text.y = element_text(angle = 45, hjust = 1))
```

Create a heatmap of the pairwise Fst measurements for each combination of taxa/populations but this time reorganize so that populations within the same taxon are kept together and they are ordered based on genetic similarity.

```{r}
#| label: Fst heatmap with taxa reordered
#| fig-height: 10
#| fig-width: 20
# Define the desired order for the axes
ZosteropsFST_reorder <- c("japonicusJAPAN", "japonicusKOREA", "japonicusRYUKYUn", "japonicusRYUKYUc", "japonicusRYUKYUs", "japonicusIZU", "japonicusOGASAWARA", "japonicusHAWAII", "meyeniLANYU", "meyeniBATAN", "montanusNORTH", "montanusPALAWAN", "montanusVISAYAS", "montanusMINDANAO", "montanusINDONESIA", "simplexASIA", "simplexNA", "erythropleurus", "everetti", "nigrorum", "palpebrosus")
# Convert the factors in the data to match the desired order
ZosteropsFSTheatRO <- ZosteropsFSTheat
ZosteropsFSTheatRO$X1 <- factor(ZosteropsFSTheat$X1, levels = ZosteropsFST_reorder)
ZosteropsFSTheatRO$X2 <- factor(ZosteropsFSTheat$X2, levels = ZosteropsFST_reorder)
# Create the heatmap with ggplot
library(ggplot2)
ggplot(data = ZosteropsFSTheatRO, aes(x = X1, y = X2, fill = value)) +
  geom_tile() +  
  geom_text(aes(label = round(value, 2))) +  
  theme_minimal() +  
  scale_fill_gradient2(low = "white", high = "red", space = "Lab", name = "Fst") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.text.y = element_text(angle = 45, hjust = 1))
```

### Pairwise Nei's distance heatmap

Create a heatmap of the Nei's distance measurements for each combination of taxa/populations.

```{r}
#| label: Nei's D heatmap
#| fig-height: 10
#| fig-width: 20
ZosteropsNeiPOP_mapping <- c("japonicusRYUKYUc", "japonicusRYUKYUs", "japonicusKOREA", "japonicusJAPAN", "japonicusOGASAWARA", "japonicusIZU", "japonicusRYUKYUn", "simplexASIA", "meyeniBATAN", "meyeniLANYU", "montanusVISAYAS", "montanusNORTH", "montanusMINDANAO", "montanusINDONESIA", "montanusPALAWAN", "nigrorum", "erythropleurus", "palpebrosus", "simplexNA", "everetti", "japonicusHAWAII")
# Reassign the X2 column with the mapped text values
ZosteropsNeiDheat$X2 <- ZosteropsNeiPOP_mapping[ZosteropsNeiDheat$X2]
ZosteropsNeiDheat$X2 <- factor(ZosteropsNeiDheat$X2, levels = unique(ZosteropsNeiDheat$X1))
# Create the heatmap matrix
ggplot(data = ZosteropsNeiDheat, aes(x = X1, y = X2, fill = value))+
  geom_tile()+
  geom_text(data = ZosteropsNeiDheat, aes(label = round(value, 4)))+
  theme_minimal()+
  scale_fill_gradient2(low = "white", high = "red", space = "Lab", name = "Nei's distance")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.text.y = element_text(angle = 45, hjust = 1))
```

Create a heatmap of the pairwise Nei's distances for each combination of taxa/populations but this time reorganize so that populations within the same taxon are kept together and they are ordered based on genetic distance.

```{r}
#| label: Nei's D heatmap with taxa reordered
#| fig-height: 10
#| fig-width: 20
ZosteropsNeiPOP_mapping <- c("japonicusRYUKYUc", "japonicusRYUKYUs", "japonicusKOREA", "japonicusJAPAN", "japonicusOGASAWARA", "japonicusIZU", "japonicusRYUKYUn", "simplexASIA", "meyeniBATAN", "meyeniLANYU", "montanusVISAYAS", "montanusNORTH", "montanusMINDANAO", "montanusINDONESIA", "montanusPALAWAN", "nigrorum", "erythropleurus", "palpebrosus", "simplexNA", "everetti", "japonicusHAWAII")
# Reassign the X2 column with the mapped text values
ZosteropsNeiDheat$X2 <- ZosteropsNeiPOP_mapping[ZosteropsNeiDheat$X2]
ZosteropsNeiDheat$X2 <- factor(ZosteropsNeiDheat$X2, levels = unique(ZosteropsNeiDheat$X1))
# Change the order for the heatmap to match the same order for the Fst heatmap
ZosteropsFST_reorder <- c("japonicusJAPAN", "japonicusKOREA", "japonicusRYUKYUn", "japonicusRYUKYUc", "japonicusRYUKYUs", "japonicusIZU", "japonicusOGASAWARA", "japonicusHAWAII", "meyeniLANYU", "meyeniBATAN", "montanusNORTH", "montanusPALAWAN", "montanusVISAYAS", "montanusMINDANAO", "montanusINDONESIA", "simplexASIA", "simplexNA", "erythropleurus", "everetti", "nigrorum", "palpebrosus")
# Convert the factors in the data to match the desired order
ZosteropsNeiDheatRO <- ZosteropsNeiDheat
ZosteropsNeiDheatRO$X1 <- factor(ZosteropsNeiDheat$X1, levels = ZosteropsFST_reorder)
ZosteropsNeiDheatRO$X2 <- factor(ZosteropsNeiDheat$X2, levels = ZosteropsFST_reorder)
# Create the heatmap matrix
ggplot(data = ZosteropsNeiDheatRO, aes(x = X1, y = X2, fill = value))+
  geom_tile()+
  geom_text(data = ZosteropsNeiDheatRO, aes(label = round(value, 4)))+
  theme_minimal()+
  scale_fill_gradient2(low = "white", high = "red", space = "Lab", name = "Nei's distance")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.text.y = element_text(angle = 45, hjust = 1))
```

### Fst map

Create a map showing the major biogeographic breaks for East Asian Zosterops and the pairwise Fst value estimated for the populations on either side of that break. In each case the Fst value was entered into the code manually from the table created in the previous step.

```{r}
#| label: Map pairwise Fst across biogeographic breaks. 
#| fig-height: 9
#| fig-width: 5
pac <- map_data("world")
ggplot()+
  geom_polygon(data = pac, aes(x=long, y = lat, group = group), fill="lightgrey", col="black", cex=.1)+
  coord_sf(xlim = c(116, 144.3), ylim = c(-5, 44))+
  # Biogeographic break between Minandao and Indonesia FST
  geom_segment(aes(x = 120, y = 4, xend = 132, yend = 6), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 133, y = 5, label = "0.22"), size = 5) +
  # Biogeographic break between Visayas and Minandao FST
  geom_segment(aes(x = 120, y = 7, xend = 129, yend = 13), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 129.3, y = 11.4, label = "0.12"), size = 5) +
  # Biogeographic break between Northern Philippines (Mindoro and Luzon) and Visayas FST
  geom_segment(aes(x = 120.2, y = 11.4, xend = 128, yend = 14), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 126.9, y = 14.7, label = "0.24"), size = 5) +
  # Biogeographic break between Northern Philippines (Mindoro and Luzon) and Palawan FST
  geom_segment(aes(x = 120.2, y = 9.4, xend = 119.7, yend = 15), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 118, y = 12.5, label = "0.26"), size = 5) +
  # Biogeographic break between Bantan Islands and Northen Philippines FST
  geom_segment(aes(x = 119, y = 19.7, xend = 125, yend = 20), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 124.5, y = 19, label = "0.15"), size = 5) +
  # Biogeographic break between Lanyu and Batan Islands FST
  geom_segment(aes(x = 121.2, y = 21.2, xend = 126, yend = 21.2), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 126.9, y = 21, label = "0.08"), size = 5) +
  # Biogeographic break between Southern Ryukyu Islands and Lanyu FST
  geom_segment(aes(x = 123, y = 29, xend = 123.6, yend = 22), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 125.5, y = 23, label = "0.31"), size = 5) +
  # Biogeographic break between Taiwan (simplex) and Lanyu FST
  geom_segment(aes(x = 122.5, y = 26, xend = 120.6, yend = 21), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 118.7, y = 23, label = "0.70"), size = 5) +
  # Biogeographic break between the central Ryukyu Islands and the Southern Ryukyu Islands FST (Hachisuka's Line)
  geom_segment(aes(x = 124.2, y = 28, xend = 129.2, yend = 22), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 130, y = 23.7, label = "0.04"), size = 5) +
  # Biogeographic break between the northern Ryukyu Islands and the central Ryukyu Islands FST (Watase's Line)
  geom_segment(aes(x = 126, y = 29.6, xend = 133, yend = 29.6), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 132.6, y = 28.5, label = "0.06"), size = 5) +
  # Biogeographic break between the main islands of Japan and the Northern Ryukyu Islands FST
  geom_segment(aes(x = 127, y = 30.8, xend = 133, yend = 31.2), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 134.3, y = 31.2, label = "0.05"), size = 5) +
  # Biogeographic break between the Korean Peninsula and surrounding islands and the main islands of Japan FST
  geom_segment(aes(x = 126, y = 31, xend = 131, yend = 35), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 125, y = 32.4, label = "0.02"), size = 5) +
  # Biogeographic break between the main islands of Japan and the Izu Islands FST
  geom_segment(aes(x = 136, y = 32.8, xend = 141, yend = 34.6), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 142, y = 33.6, label = "0.09"), size = 5) +
  # Biogeographic break between the Izu Islands and the Ogasawara and Volcano (Iwo) Islands FST
  geom_segment(aes(x = 139, y = 28, xend = 144, yend = 30), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 144, y = 28.6, label = "0.19"), size = 5) +
  theme_classic()
```

### Nei's distance map

Create a map showing the major biogeographic breaks for East Asian Zosterops and Nei's distance estimates for the populations on either side of that break. In each case the Nei's distance value was entered into the code manually from the table created in the previous step.

```{r}
#| label: Map Nei's distance across biogeographic breaks. 
#| fig-height: 9
#| fig-width: 5
pac <- map_data("world")
ggplot()+
  geom_polygon(data = pac, aes(x=long, y = lat, group = group), fill="lightgrey", col="black", cex=.1)+
  coord_sf(xlim = c(116, 144.3), ylim = c(-5, 44))+
  # Biogeographic break between Minandao and Indonesia Nei's distance
  geom_segment(aes(x = 120, y = 4, xend = 132, yend = 6), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 132, y = 5, label = "0.0146"), size = 3) +
  # Biogeographic break between Visayas and Minandao Nei's distance
  geom_segment(aes(x = 120, y = 7, xend = 129, yend = 13), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 127.9, y = 10.9, label = "0.0075"), size = 3) +
  # Biogeographic break between Northern Philippines (Mindoro and Luzon) and Visayas Nei's distance
  geom_segment(aes(x = 120.2, y = 11.4, xend = 128, yend = 14), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 127, y = 14.5, label = "0.0137"), size = 3) +
  # Biogeographic break between Northern Philippines (Mindoro and Luzon) and Palawan Nei's distance
  geom_segment(aes(x = 120.2, y = 9.4, xend = 119.7, yend = 15), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 118, y = 13, label = "0.0186"), size = 3) +
  # Biogeographic break between Bantan Islands and Northen Philippines Nei's distance
  geom_segment(aes(x = 119, y = 19.7, xend = 125, yend = 20), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 124.5, y = 19.3, label = "0.0094"), size = 3) +
  # Biogeographic break between Lanyu and Batan Islands Nei's distance
  geom_segment(aes(x = 121.2, y = 21.2, xend = 126, yend = 21.2), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 127, y = 21.3, label = "0.0068"), size = 3) +
  # Biogeographic break between Southern Ryukyu Islands and Lanyu Nei's distance
  geom_segment(aes(x = 122, y = 29, xend = 123.6, yend = 22), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 125.1, y = 23, label = "0.014"), size = 3) +
  # Biogeographic break between Taiwan (simplex) and Lanyu Nei's distance
  geom_segment(aes(x = 122.5, y = 26, xend = 120.6, yend = 21), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 118.7, y = 23, label = "0.0698"), size = 3) +
  # Biogeographic break between the central Ryukyu Islands and the Southern Ryukyu Islands Nei's distance (Hachisuka's Line)
  geom_segment(aes(x = 123.2, y = 28, xend = 129.2, yend = 22), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 129.9, y = 23.7, label = "0.0024"), size = 3) +
  # Biogeographic break between the northern Ryukyu Islands and the central Ryukyu Islands Nei's distance (Watase's Line)
  geom_segment(aes(x = 126, y = 29.8, xend = 133, yend = 29.6), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 132.6, y = 28.8, label = "0.0037"), size = 3) +
  # Biogeographic break between the main islands of Japan and the Northern Ryukyu Islands Nei's distance
  geom_segment(aes(x = 127, y = 30.8, xend = 134, yend = 31.2), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 135.3, y = 31.5, label = "0.004"), size = 3) +
  # Biogeographic break between the Korean Peninsula and surrounding islands and the main islands of Japan Nei's distance
  geom_segment(aes(x = 126, y = 31, xend = 131, yend = 35), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 125.5, y = 32.4, label = "0.0019"), size = 3) +
  # Biogeographic break between the main islands of Japan and the Izu Islands Nei's distance
  geom_segment(aes(x = 136, y = 32.8, xend = 141, yend = 34.6), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 142, y = 33.9, label = "0.0029"), size = 3) +
  # Biogeographic break between the Izu Islands and the Ogasawara and Volcano (Iwo) Islands Nei's distance
  geom_segment(aes(x = 139, y = 28, xend = 144, yend = 30), linetype = "dashed", size = 1, color = "red") +
  geom_text(aes(x = 144, y = 29, label = "0.0041"), size = 3) +
  theme_classic()
```

*Last updated: `r format(Sys.time(), '%B %d, %Y %H:%M')`*

## Close log

End the log.

```{r}
#| label: End log file. 
log_close()
```
