---
title: "Sampling Maps"

params:
  wrap_code: true
editor: visual
knitr: true
bibliography: ZosteropsMapsCitations.bib
---

## Goals

What follows is our workflow for mapping the sampling localities for all the individual samples included in the Zosterops single nucleotide polymorphism (SNP) data generated from a double digest restriction-site associated DNA (ddRAD) library and Sanger sequencing datasets (see DeRaad *et al*.2024 for details on the ddRAD library [@deraad2024].

## Getting started

We are using R version 4.4.2, RStudio version 2024.09.1 for the analysis. Quarto version 1.6.39 for Mac OS was used to document the workflow. Set the working directory in RStudio to one where the `txt` file denoting the identity and sampling location of the samples are accessible.

## General methods summary

Sampling included specimen vouchered tissue samples and museum archived blood sampling from individual banded birds in the field. Samples were sourced from the American Museum of Natural History (New York, NY, USA), Cincinnati Museum Center (Cincinnati, OH, USA), Field Museum of Natural History (Chicago, IL, USA), Institute of Zoology of the Guangdong Academy of Sciences (Guangzhou, People’s Republic of China), University of Kansas Biodiversity Institute and Natural History Museum (Lawrence, KS, USA), Natural History Museum of Los Angeles County (Los Angeles, CA, USA), National Museum of Nature and Science (Tokyo, Japan), Smithsonian National Museum of Natural History (Washington D.C., USA), and the University of Washington Burke Museum (Seattle, WA, USA). For the sake of clarity, we treat *Z. japonicus* and *Z. montanus* according to the former taxonomic arrangement (i.e., as separate species) as we test various species delimitation hypotheses. Our sampling includes multiple individuals within each of five out of the six recognized subspecies for *Z. japonicus* and seven out of the nine subspecies within *Z. montanus*, including sampling from all the recognized *Z. montanus* subspecies within the Philippines. A total of 208 genomic DNA (gDNA) samples were used in this study with 161 represented in the Sanger sequencing dataset, 124 represented in the filtered SNP dataset, and 77 samples overlapping both genetic datasets.

Latitude and longitude were available for most of the vouchered museum specimens and where not available were estimated from the locality description using the BerkeleyMapper 2.0 (<https://www.gbif.org/tool/81294/berkeleymapper>) georeferencing tool. Maps showing approximate locations of individual samples were created using the R packages *ggplot2* [@ggplot2], *maps* [@maps], and *sf* [@sf].

## Data files

This analysis requires the following data files.

- `ZosteropsPOPS4_LatLong.txt` A text file of individual samples included in the SNP dataset matched with taxon, locale, and latitude and longitude.

- `ZosteropsPOPS_SANGER_LatLong.txt` A text file of individual samples included in the Sanger sequencing dataset that do not appear in the SNP dataset matched with taxon, locale, and latitude and longitude.

### Individual samples and collecting localities: SNPs

A tab delimited text file with the individual sample IDs, taxon designations, the name of the sampling locales (designated by island or province), and latitude and longitude was created to generate maps. To reduce the amount of clutter on the maps subspecies were clustered together in shared geographic groups. *Zosterops japonicus* subspecies were designated as either *Z. japonicus alani* for the most remote subspecies in the Ogasawara and Volcano Islands or grouped within a single *Z. japonicus* designation for all the remaining subspecies. Subspecies for *Z. montanus* were clustered into three groups including a Palawan group (montanus PALAWAN) consisting of *Z. montanus parkesi*, a northern Philippine group (montanus NORTH) for those subspecies on Mindoro and Luzon (*Z. montanus halconensis* amd *Z. montanus whiteheadi*), and a southern Philippine group (montanus SOUTH) including the remaining *Z. montanus* subspecies included in the study (*Z. montanus diuatae*, *Z. montanus montanus*, *Z. montanus pectoralis*, *Z. montanus vulcani*, and an unnamed but suggested subspecies on Panay). Subspecies within the *Z. everetti* and *Z. nigrorum* species were lumped together under a single species designation for each in the sampling maps but distinguished in later analyses. Samples from introduced populations in Hawaii and California are included in this list, but they are excluded from the final figure. The individual samples and their accompanying data from the file `ZosteropsPOPS4_LatLong.txt` are listed below. All of the samples listed in this file appear in the SNP analyses.

```         
SAMPLE  TAXON   LOCALE  LAT LONG
ZJlo012 japonicus   KIKAI   28.317  129.9401
ZJlo015 japonicus   AMAMI   28.3192 129.2765
ZJlo016 japonicus   MIYAKOJIMA  24.7916 125.3302
ZJlo017 japonicus   IRIOMOTE    24.3301 123.8188
ZJlo021 japonicus   TOKUNOSHIMA 27.7904 128.9668
ZJlo022 japonicus   TOKUNOSHIMA 27.7904 128.9668
ZJlo024 japonicus   YONAGUMI    24.4680 123.0045
ZJlo031 japonicus   KUMEJIMA    26.3504 126.7712
ZJlo045 japonicus   AMAMI   28.3192 129.2765
ZJlo046 japonicus   MIYAKOJIMA  24.8218 125.3137
ZJlo050 japonicus   OKINOERABU  27.5737 128.5737
ZJlo052 japonicus   OKINOERABU  27.5737 128.5737
ZJlo055 japonicus   KIKAI   28.317  129.9401
ZJja001 japonicus   GEOJE   34.8806 128.6211
ZJja002 japonicus   GEOJE   34.8806 128.6211
ZJja003 japonicus   JEJU    33.489  126.4983
ZJja004 japonicus   JEJU    33.489  126.4983
ZJja005 japonicus   JEJU    33.489  126.4983
ZJja009 japonicus   HEUKSANDO   34.6707 125.4196
ZJja010 japonicus   HEUKSANDO   34.6707 125.4196
ZJja012 japonicus   HONSHU  36.0835 140.0764
ZJja013 japonicus   HONSHU  36.0835 140.0764
ZJja014 japonicus   HONSHU  35.9086 139.6248
ZJja016 japonicus   SHIKOKU 34.3708 133.9198
ZJja030 japonicus   HEUKSANDO   34.6707 125.4196
ZJal003 japonicus alani HAHAJIMA    26.6735 142.1541
Zjal004 japonicus alani HAHAJIMA    26.6735 142.1541
ZJal005 japonicus alani HAHAJIMA    26.6735 142.1541
ZJal010 japonicus alani ANEJIMA 26.5506 142.1581
ZJal011 japonicus alani ANEJIMA 26.5506 142.1581
ZJal012 japonicus alani ANEJIMA 26.5506 142.1581
ZJal020 japonicus alani IWOTO 24.7876 141.3156
ZJst007 japonicus   KOZUSHIMA   34.2055 139.1345
ZJst009 japonicus   KOZUSHIMA   34.2055 139.1345
ZJst010 japonicus   KOZUSHIMA   34.2055 139.1345
ZJst012 japonicus   MIYAKEJIMA  34.0788 139.5178
ZJst014 japonicus   MIYAKEJIMA  34.0788 139.5178
ZJst015 japonicus   MIYAKEJIMA  34.0788 139.5178
ZJin001 japonicus   KAGOSHIMA   30.3715 130.6646
ZJin003 japonicus   KAGOSHIMA   30.3715 130.6646
ZJsi002 simplex TAIWAN  22.7613 121.1438
ZJsi017 simplex GUANGXI 22.8152 108.3275
ZJsi032 simplex GUANGXI 22.8152 108.3275
ZJsi023 simplex GUANGDONG   23.1317 113.2663
ZJsi024 simplex GUANGDONG   23.1317 113.2663
ZJsi027 simplex GUANGXI 22.8152 108.3275
ZJsi028 simplex GUANGXI 22.8152 108.3275
ZJsi029 simplex GUANGXI 22.8152 108.3275
ZJsi031 simplex GUANGXI 22.8152 108.3275
ZPxx001 simplex SINGAPORE   1.3189  103.8512
ZPxx002 simplex SINGAPORE   1.3521  103.8198
ZMxx002 meyeni  BATAN   20.4700 121.9910
ZMxx003 meyeni  LANYU   22.0811 121.5048
ZMxx004 meyeni  LANYU   22.0811 121.5048
ZMxx005 meyeni  LANYU   22.0811 121.5048
ZMxx006 meyeni  LANYU   22.0093 121.5721
ZMOxx001    montanus SOUTH  PANAY   11.4062 122.1235
ZMOxx002    montanus SOUTH  PANAY   11.4062 122.1235
ZMOxx003    montanus SOUTH  PANAY   11.3936 122.1661
ZMOxx005    montanus SOUTH  PANAY   11.4062 122.1235
ZMOha001    montanus NORTH  MINDORO 13.3082 121.0765
ZMOwh002    montanus NORTH  LUZON   15.5286 119.9608
ZMOwh003    montanus NORTH  LUZON   17.4670 121.0670
ZMOwh004    montanus NORTH  LUZON   15.4819 120.1194
ZMOwh005    montanus NORTH  LUZON   17.0373 121.1010
ZMOwh006    montanus NORTH  LUZON   16.5976 120.8986
ZMOvu002    montanus SOUTH  MINDANAO    7.8514  126.0458
ZMOvu003    montanus SOUTH  MINDANAO    7.8514  126.0458
ZMOvu004    montanus SOUTH  MINDANAO    7.0189  125.4984
ZMOvu005    montanus SOUTH  MINDANAO    7.0189  125.4984
ZMOmo001    montanus SOUTH  SULAWESI    -5.4081 119.8989
ZMOpa001    montanus PALAWAN    PALAWAN 8.8183  117.6750
ZMOpa002    montanus PALAWAN    PALAWAN 8.8183  117.6750
ZNni001 nigrorum    PANAY   11.4062 122.1235
ZERxx002    erythropleurus  RUSSIAb 46.0200 135.2500
ZERxx004    erythropleurus  RUSSIAb 44.9200 131.8900
ZERxx005    erythropleurus  RUSSIAb 44.3674 133.0265
Zsim23588   simplex VIETNAM 22.3860 102.2380
Zsim28142   simplex VIETNAM 21.9490 104.2550
Zsim30897   palpebrosus VIETNAM 15.1100 107.8240
Zpal23498   palpebrosus VIETNAM 22.3860 102.2380
Zpal23522   palpebrosus VIETNAM 22.3860 102.2380
Zmon20893   montanus SOUTH  NEGROS  10.6743 123.1886
Zmon20892   montanus SOUTH  NEGROS  10.6743 123.1886
Zmon20902   montanus SOUTH  NEGROS  10.6743 123.1886
Zmon20909   montanus SOUTH  NEGROS  10.6655 123.1787
Zsim31166   montanus SOUTH  JAVA    -7.6145 110.7122
Zsim31159   montanus SOUTH  JAVA    -7.6145 110.7122
Zmey17876   meyeni  BATAN   20.4700 121.9910
Zmey17877   meyeni  BATAN   20.4700 121.9910
Zmey17852   meyeni  BATAN   20.4570 121.9900
Zmey17922   meyeni  SABTANG 20.2850 121.8710
Zmey17920   meyeni  SABTANG 20.2850 121.8710    
Zmey17925   meyeni  SABTANG 20.2850 121.8710
Zmey17923   meyeni  SABTANG 20.2850 121.8710
Zery28090   erythropleurus  VIETNAMnb   21.9490 104.2550
Zery28091   erythropleurus  VIETNAMnb   21.9490 104.2550
Zery28088   erythropleurus  VIETNAMnb   21.9490 104.2550
Zni10863    nigrorum    CAMIGUINNORTE   18.9290 121.8990
Zni14341    nigrorum    CAMIGUINSUR 9.1780  124.7197    
Zni19650    nigrorum    LUZON   15.6436 121.5149
Zni17984    nigrorum    LUZON   14.0390 122.7860
ZjaDOT-10981    palpebrosus VIETNAM 22.7601 104.8805
ZjaDOT-5235 simplex TAIWAN  23.8280 120.7925
Z_LA_122866_2   simplex LOSANGELES  33.6966 -118.0183
Z_LA_122577_2   simplex LOSANGELES  33.6966 -118.0183
Z_LA_122188_2   simplex LOSANGELES  33.6966 -118.0183
Zsim13809   simplex GUIZHOU 25.4114 107.8872
Zsim13773   simplex GUIZHOU 25.4114 107.8872
Zsim6797    simplex HUNAN   28.4167 114.1167
Zsim10336   simplex GUANGXI 21.8400 107.8800
Zsim11362   simplex GUIZHOU 29.1668 107.5750
Zsim11102   simplex GUIZHOU 28.2261 107.1596
Zsim11220   simplex GUIZHOU 28.2261 107.1596
Zmon20899   montanus SOUTH  NEGROS  10.6743 123.1886
Zmon28375   montanus SOUTH  MINDANAO    9.0790  125.6650
Zev13949    everetti    CAMIGUINSUR 9.1925  124.7085
Zev31650    everetti    SAMAR   11.8280 125.2760
Zev28451    everetti    MINDANAO    9.0740  125.6420
Z_HI_BRY431 japonicus   HAWAII  19.5660 -155.4530
Z_HI_NAN290 japonicus   HAWAII  19.5660 -155.4530
Z_HI_WAI087 japonicus   HAWAII  19.5660 -155.4530
Z_HI_WAI078 japonicus   HAWAII  19.5660 -155.4530
Z_HI_SOL783 japonicus   HAWAII  19.5660 -155.4530
```

### Individual samples and collecting localities: Sanger sequencing

Additional samples appear in the Sanger sequencing dataset that are not included in the SNP dataset. Those samples, their taxon designations, and collecting locality information are in the file `ZosteropsPOPS_SANGER_LatLong.txt` and listed below.

```         
SAMPLE  TAXON   LOCALE  LAT LONG
ZEba001 everetti    MINDANAO    7.0189  125.4984
ZEba002 everetti    MINDANAO    9.1925  124.7085
ZEba003 everetti    MINDANAO    9.0740  125.6420
ZEbo001 everetti    SAMAR   11.8280 125.2760
ZERxx001    erythropleurus  RUSSIAb 42.9700 132.8900
ZERxx003    erythropleurus  RUSSIAb 44.7500 132.7700
ZJal001 japonicus alani HAHAJIMA    26.6735 142.1541
ZJal002 japonicus alani HAHAJIMA    26.6735 142.1541
ZJal006 japonicus alani IMOTOJIMA   26.5587 142.2085
ZJal007 japonicus alani IMOTOJIMA   26.5587 142.2085
ZJal008 japonicus alani IMOTOJIMA   26.5587 142.2085
ZJal009 japonicus alani IMOTOJIMA   26.5587 142.2085
ZJal013 japonicus alani ANEJIMA 26.5506 142.1581
ZJal014 japonicus alani ANEJIMA 26.5506 142.1581
ZJal015 japonicus alani IMOTOJIMA   26.5587 142.2085
ZJal016 japonicus alani IWOTO   24.7876 141.3156
ZJal017 japonicus alani IWOTO   24.7876 141.3156
ZJal018 japonicus alani IWOTO   24.7876 141.3156
ZJal019 japonicus alani IWOTO   24.7876 141.3156
ZJba001 meyeni  LANYU   22.0436 121.5484
ZJba002 meyeni  LANYU   22.0436 121.5484
ZJja006 japonicus   JEJU    33.4890 126.4983
ZJja007 japonicus   JEJU    33.4890 126.4983
ZJja008 japonicus   HONSHU  35.6388 139.7256
ZJja011 japonicus   HUEKSANDO   34.6707 125.4196
ZJja015 japonicus   HONSHU  35.7263 139.7168
ZJja017 japonicus   HONSHU  36.6024 137.6088
ZJja018 japonicus   HOKKAIDO    42.9393 143.4477
ZJja019 japonicus   HONSHU  36.3402 139.4498
ZJja021 japonicus   HONSHU  35.6388 139.7256
ZJja022 japonicus   HONSHU  37.2832 138.9816
ZJja024 japonicus   HONSHU  36.7846 136.8348
ZJja026 japonicus   HONSHU  33.7030 135.3759
ZJja031 japonicus   HONSHU  35.7032 139.6975
ZJja032 japonicus   HONSHU  35.7032 139.6975
ZJja034 japonicus   TSUSHIMA    34.4081 129.2860
Zjja035 japonicus   TSUSHIMA    34.4081 129.2860
ZJlo001 japonicus alani CHICHIJIMA  27.0744 142.2178
ZJlo002 japonicus alani CHICHIJIMA  27.0744 142.2178
ZJlo003 japonicus alani CHICHIJIMA  27.0744 142.2178
ZJlo004 japonicus alani CHICHIJIMA  27.0744 142.2178
ZJlo005 japonicus alani CHICHIJIMA  27.0744 142.2178
ZJlo006 japonicus   IRIOMOTE    24.3301 123.8188
ZJlo007 japonicus   OKINAWA 26.7456 128.1779
ZJlo008 japonicus   OKINAWA 26.7456 128.1779
ZJlo009 japonicus   OKINAWA 26.7456 128.1779
ZJlo010 japonicus   ISHIGAKI    24.4064 124.1754
ZJlo011 japonicus   ISHIGAKI    24.4064 124.1754
ZJlo013 japonicus   KIKAI   28.317  129.9401
ZJlo014 japonicus   IRIOMOTE    24.3301 123.8188
ZJlo018 japonicus   IRIOMOTE    24.3301 123.8188
ZJlo020 japonicus   IRIOMOTE    24.3301 123.8188
ZJlo023 japonicus   OKINOERABU  27.5737 128.5737
ZJlo026 japonicus   MIYAKOJIMA  24.8218 125.3137
ZJlo027 japonicus   YONAGUNI    24.4680 123.0045
ZJlo032 japonicus   KUMEJIMA    26.3504 126.7712
ZJlo034 japonicus   MIYAKOJIMA  24.7354 125.3000
ZJlo035 japonicus   OKINAWA 26.7456 128.1779
ZJlo038 japonicus   OKINAWA 26.7456 128.1779
ZJlo039 japonicus   ISHIGAKI    24.4064 124.1754
ZJlo042 japonicus   AMAMI   28.3192 129.2765
ZJlo051 japonicus   OKINOERABU  27.5737 128.5737
ZJsi001 simplex TAIWAN  22.7613 121.1438
ZJsi003 simplex TAIWAN  22.7613 121.1438
ZJsi005 simplex TAIWAN  22.7613 121.1438
ZJsi015 simplex MACAO   22.1987 113.5439
ZJsi018 simplex SICHUAN 30.6509 104.0757
ZJsi025 simplex GUANGDONG   23.1317 113.2663
ZJsi026 simplex YUNNAN  25.0453 102.7097
ZJst001 japonicus   NIIJIMA 34.3771 139.2569
ZJst002 japonicus   AOGASHIMA   32.4570 139.7652
ZJst006 japonicus   AOGASHIMA   32.4570 139.7652
ZJst008 japonicus   KOZUSHIMA   34.2055 139.1345
ZJst011 japonicus   KOZUSHIMA   34.2055 139.1345
ZJst013 japonicus   MIYAKEJIMA  34.0788 139.5178
ZJst016 japonicus   MIYAKEJIMA  34.0788 139.5178
ZJst017 japonicus   NIIJIMA 34.3771 139.2569
ZMOmo004    montanus SOUTH  JAVA    -7.6145 110.7122
ZMOvu001    montanus SOUTH  MINDANAO    7.0189  125.4984
ZMOvu006    montanus SOUTH  MINDANAO    6.0234  124.6414
ZMOvu007    montanus SOUTH  MINDANAO    6.0234  124.6414
ZMOxx004    montanus SOUTH  PANAY   11.3936 122.1661
ZMOxx006    montanus SOUTH  PANAY   11.4109 122.2881
ZMOxx007    montanus SOUTH  PANAY   11.4109 122.2881
ZMOxx008    montanus SOUTH  PANAY   11.4109 122.2881
ZMxx007 meyeni  LANYU   22.0093 121.5721
ZMxx008 meyeni  LANYU   22.0093 121.5721
ZPxx001 simplex SINGAPORE   1.3189  103.8512
ZPxx002 simplex SINGAPORE   1.3521  103.8198
```

## Packages

Set the working directory to one where the data files `ZosteropsPOPS4_LatLong.txt` and `Zosterops_SANGER_LatLong.txt` are accessible. Creating the sampling maps requires the following R packages.

- logr [@logr] Optional if you wish to create a log file for the analysis.

- ggplot2 [@ggplot2]

- maps [@maps]

- sf [@sf]

Install the packages using the Tools/Install Packages selection from the RStudio menu and load the package libraries.

```{r}
#| label: Load libraries and create a log file
library(logr)
log_open(file_name = "ZosteropsMaps.log")
#load in each library used for mapping
library(ggplot2) #v3.4.1
library(maps) #v3.4.1
library(sf) #v1.0-12
```

## Read and edit data: SNP sampling

Read the data into a data frame using `read.csv` and indicate that the fields are tab delimited (`sep = "\t"`) and contain a header row (`header = TRUE`). First complete this process for the sampling data for the SNP dataset.

```{r}
#| label: Read tab delimited data file for SNP samples
# Read tab delimited (sep = "\t") file with latitude and longitude
ZLocations <- read.csv(
  "Data/Location files/ZosteropsPOPS4_LatLong.txt", 
                       sep = "\t", header = TRUE)
```

Remove the samples from the introduced populations. Those samples are from Los Angeles (*Z. simplex*) and Hawaii (*Z. japonicus*) and are included in analyses but will not appear on the sampling maps.

```{r}
#| label: Remove samples from introduced populations
# Remove Los Angeles and Hawaii samples
ZLocations_NATIVE <- ZLocations[-c(105,106,107,120,121,122,123,124),]
```

## Count individual samples per taxon: SNP sampling

To illustrate the relative number of individual samples from any particular site we will count the number of samples associated with a unique latitude and longitude. First split the data frame by the TAXON column.

```{r}
#| label: Split by taxon
# Split by TAXON
Zlocations_NATIVE_TAXON <- split(ZLocations_NATIVE, 
                                 ZLocations_NATIVE$TAXON)
```

Using a loop count samples grouped by LAT and LONG columns and save this as a new data frame called `sampling.df`.

```{r}
#| label: Loop to count individual samples
# initialize sampling.df which will be a df of samples grouped by unique lat/long
sampling.df<-data.frame(NULL)
# use a for loop to split the dataframe by species and summarize the number 
# of individuals sampled at each locality for each species
for (i in names(Zlocations_NATIVE_TAXON)){
  samps <- Zlocations_NATIVE_TAXON[[i]] %>% 
    dplyr::group_by(LAT, LONG) %>% 
    dplyr::summarize(count=dplyr::n())
  df <- cbind(rep(i, times=nrow(samps)), samps)
  sampling.df <- as.data.frame(rbind(sampling.df, df))
}
```

Edit the column names in the new `sampling.df` data frame.

```{r}
#| label: Edit the column names
#fix colnames
colnames(sampling.df)<-c("Species","Latitude","Longitude","count")
```

## Read and edit data: Sanger sampling

Repeat the process above for the file containing the additional samples included in the Sanger sequencing dataset. There are no samples from introduced populations in the Sanger sequencing data so there is no need to follow up with the editing step.

```{r}
#| label: Read tab delimited data file for Sanger samples
# Pull location data for samples in the SANGER sequencing dataset 
# but not in the NSP dataset
# Read tab delimited (sep = "\t") file with latitude and longitude
ZLocationsSANGER <- read.csv(
  "Data/Location files/ZosteropsPOPS_SANGER_LatLong.txt", 
                             sep = "\t", header = TRUE)
```

## Count individual samples per taxon: Sanger sampling

Split the data frame for the Sanger sequencing sampling by the TAXON column.

```{r}
#| label: Split the Sanger sampling by TAXON
# Split by TAXON
ZlocationsSANGER_TAXON <- split(ZLocationsSANGER, 
                                ZLocationsSANGER$TAXON)
```

Using a loop count samples grouped by LAT and LONG columns and save this as a new data frame called `samplingSANGER.df`.

```{r}
#| label: Loop to count individual Sanger samples
# init samplingSANGER.df which will be a df of samples 
# grouped by unique lat/long
samplingSANGER.df<-data.frame(NULL)
#use a for loop to split the dataframe by species and 
#summarize the number of individuals sampled at each 
#locality for each species
for (i in names(ZlocationsSANGER_TAXON)){
  samps <- ZlocationsSANGER_TAXON[[i]] %>% 
    dplyr::group_by(LAT, LONG) %>% 
    dplyr::summarize(count=dplyr::n())
  df <- cbind(rep(i, times=nrow(samps)), samps)
  samplingSANGER.df <- as.data.frame(rbind(samplingSANGER.df, df))
}
```

Edit the column names in the new `samplingSANGER.df` data frame.

```{r}
#| label: Edit the column names for the Sanger samples
#fix colnames
colnames(samplingSANGER.df)<-c("Species","Latitude","Longitude","count")
```

## Create a sampling map

### Mapping all samples

Download a world base map.

```{r}
#| label: Download a world base map
#make base world map
pac<-map_data("world")
```

First create a sampling map for all the Zosterops samples in the study (excluding those from introduced populations). Include two lines with the `geo_point` command. One line will define those points from the SNP sampling data frame (`sampling.df`) and identify those points as filled circles (`pch=21`). The next line will define the points using the `geo_point` command for the additional Sanger sequencing samples (`samplingSANGER.df`) and identify those points as triangles (`pch=24`). Because many of the sampling locations are the same for different taxa define `position = position_jitter(h=0.5, w=0.5)` to offset the points on the map so as they do not perfectly overlap. Also set `alpha=0.5` to make the points transparent. This will also help in showing those places where sampled taxa overlap in location.

```{r}
#| label: Sampling map for native range Zosterops samples
#| fig-height: 14
#| fig-width: 14
# plot all native samples
ggplot()+
  geom_polygon(data = pac, aes(x=long, y = lat, group = group), 
               fill="lightgrey", col="black", cex=.1)+
  coord_sf(xlim = c(90, 145), ylim = c(-10, 48)) + 
  geom_point(data = sampling.df, aes(x = Longitude, y = Latitude, 
                                     fill=Species, size=count), 
             position = position_jitter(h = 0.5, w = 0.5), pch=21, 
             color="black", alpha =0.5, show.legend=TRUE) +
  geom_point(data = samplingSANGER.df, aes(x = Longitude, y = Latitude, 
                                           fill=Species, size=count), 
             position = position_jitter(h = 0.5, w = 0.5), pch=24, 
             color="black", alpha =0.5, show.legend=TRUE) +
  theme_classic()+
  scale_fill_manual(values=c("#004949", "#DB6D00", "#D82632", 
                             "#290AD8", "#FED439", "#B66DFF", 
                             "#006DDB", "#FA9C88", "#924900", 
                             "#7BA9CB", "#008600"))+
  scale_size_continuous(range = c(3,6))+
  guides(fill = guide_legend(override.aes = list(size = 8), order=1, 
                             label.theme = element_text(face = "italic"), 
                             position = "left"),
         linewidth = guide_legend(order = 2))+
  theme(legend.position = "left")
```

Save the sampling map as a `png` file.

```{r}
#| label: Save the Zosterops sampling map as a png
# Save the sampling plot
ggsave("ZosteropsSamplingMapALL.png", device = "png", 
       width = 9, height = 6.5, units = "in", dpi = 500)
```

### Mapping Ryukyu Island samples

Because many of the samples were collected on small islands and may appear close together at a large scale, create more geographically focused maps for particular regions with heavy sampling. This may be accomplished by editing the `coord_sf` command to narrow the latitude and longitude for each map. First create a map that is focused on those samples from the Ryukyu Islands.

```{r}
#| label: Ryukyu islands sampling map
#| fig-height: 8
#| fig-width: 6
# plot all native samples in the Ryukyu Islands
ggplot()+
  geom_polygon(data = pac, aes(x=long, y = lat, group = group), 
               fill="lightgrey", col="black", cex=.1)+
  coord_sf(xlim = c(122, 131), ylim = c(24, 31)) + 
  geom_point(data = sampling.df, aes(x = Longitude, y = Latitude, 
                                     fill=Species, size=count), 
             position = position_jitter(h = 0.1, w = 0.1), pch=21, 
             color="black", alpha =0.5, show.legend=FALSE) +
  geom_point(data = samplingSANGER.df, aes(x = Longitude, y = Latitude, 
                                           fill=Species, size=count), 
             position = position_jitter(h = 0.1, w = 0.1), pch=24, 
             color="black", alpha =0.5, show.legend=FALSE) +
  theme_classic()+
  scale_fill_manual(values=c("#004949", "#DB6D00", "#D82632", 
                             "#290AD8", "#FED439", "#B66DFF", 
                             "#006DDB", "#FA9C88", "#924900", 
                             "#7BA9CB", "#008600"))+
  scale_size_continuous(range = c(3,6))
```

Save the Ryukyu Islands sampling map as a `png` file.

```{r}
#| label: Save Ryukyu map as a png file
# Save the sampling plot
ggsave("ZosteropsSamplingMapRYUKYU.png", device = "png", 
       width = 9, height = 6.5, units = "in", dpi = 500)
```

### Mapping Ogasawara and Volcano Island samples

Because the samples from the Ogasawara and Volcano Islands are isolated geographically make a separate map for those samples. Many of these islands are tiny and will not show up on the map but rest assured they are there.

```{r}
#| label: Ogasawara and Volcano Islands samping map
#| fig-height: 8
#| fig-width: 6
# plot all native samples in the Ogasawara and Volcano Islands
ggplot()+
  geom_polygon(data = pac, aes(x=long, y = lat, group = group), 
               fill="lightgrey", 
               col="black", cex=.1)+
  coord_sf(xlim = c(141, 143), ylim = c(24.4, 27.2)) + 
  geom_point(data = sampling.df, aes(x = Longitude, y = Latitude, 
                                     fill=Species, size=count), 
             position = position_jitter(h = 0.05, w = 0.05), pch=21, 
             color="black", alpha =0.5, show.legend=FALSE) +
  geom_point(data = samplingSANGER.df, aes(x = Longitude, y = Latitude, 
                                           fill=Species, size=count), 
             position = position_jitter(h = 0.05, w = 0.05), pch=24, 
             color="black", alpha =0.5, show.legend=FALSE) +
  theme_classic()+
  scale_fill_manual(values=c("#004949", "#DB6D00", "#D82632", 
                             "#290AD8", "#FED439", "#B66DFF", 
                             "#006DDB", "#FA9C88", "#924900", 
                             "#7BA9CB", "#008600"))+
  scale_size_continuous(range = c(3,6))
```

Save the Ogasawara and Volcano Islands sampling map as a `png` file.

```{r}
#| label: Save Ogasawara and Volcano Islands map as a png file
# Save the sampling plot
ggsave("ZosteropsSamplingMapOGASAWARA.png", device = "png", 
       width = 9, height = 6.5, units = "in", dpi = 500)
```

### Mapping Philippine samples

Many samples across multiple taxa were sampled in the Philippines for this study. Create a separate map that focuses on those samples.

```{r}
#| label: Philippine sampling map
#| fig-height: 8
#| fig-width: 6
# plot all native samples in the Philippines
ggplot()+
  geom_polygon(data = pac, aes(x=long, y = lat, group = group), 
               fill="lightgrey", 
               col="black", cex=.1)+
  coord_sf(xlim = c(115, 127), ylim = c(6, 20.3)) + 
  geom_point(data = sampling.df, aes(x = Longitude, y = Latitude, 
                                     fill=Species, size=count), 
             position = position_jitter(h = 0.2, w = 0.2), pch=21, 
             color="black", alpha =0.5, show.legend=FALSE) +
  geom_point(data = samplingSANGER.df, aes(x = Longitude, y = Latitude, 
                                           fill=Species, size=count), 
             position = position_jitter(h = 0.2, w = 0.2), pch=24, 
             color="black", alpha =0.5, show.legend=FALSE) +
  theme_classic()+
  scale_fill_manual(values=c("#004949", "#DB6D00", "#D82632", 
                             "#290AD8", "#FED439", "#B66DFF", 
                             "#006DDB", "#FA9C88", "#924900", 
                             "#7BA9CB", "#008600"))+
  scale_size_continuous(range = c(3,6))
```

Save the Philippine sampling map as a `png` file.

```{r}
#| label: Save Philppine map as a png file
# Save the sampling plot
ggsave("ZosteropsSamplingMapPHILIPPINES.png", device = "png", 
       width = 9, height = 6.5, units = "in", dpi = 500)
```

*Last updated: `r format(Sys.time(), '%B %d, %Y %H:%M')`*

## Close log

Close the log file.

```{r}
#| label: Close log file
log_close()
```
