---
title: "Distance trees, PCA, UMAP, and DAPC analyses"

format:
  html:
    fig-format: png
    fig-dpi: 700
    
editor: visual
knitr: true
bibliography: ZosteropsSNPanalysis.bib
---

## Goals

What follows is an analysis of the Zosterops single nucleotide polymorphism (SNP) data from our double digest restriction-site associated DNA (ddRAD) library or RADseq. This analysis seeks to describe general metrics of diversity and variation in East Asian Zosterops and makes some attempt to assess the utility of these data in clustering individuals by putative taxonomic and population groups.

## Getting started

We are using R version 4.4.2, RStudio version 2024.09.1 for the analysis. Quarto version 1.6.39 for Mac OS was used to document the workflow. Set the working directory in RStudio to one where the `vcf.gz` zipped data file and a `txt` file denoting the identity of the samples are accessible.

## General methods summary

### RADseq-derived SNP dataset

The RADseq-derived SNP dataset in this study was previously analyzed in DeRaad *et al.* [@deraad2024], which includes detailed methodological description. Briefly, restriction site associated DNA libraries were constructed by digesting gDNA using the enzyme *NdeI* and performing size selection for fragments between 495 – 605 base-pairs (bp). The complete library preparation protocol is available at <https://github.com/DevonDeRaad/zosterops.rad/blob/main/lab.protocols/MSG-NdeI_2plates-150samples_DAD-Moyle_230530.doc>. Pooled, bar-coded libraries were sequenced on the Illumina NextSeq2000 at the University of Kansas Genomic Sequencing Core using a P2 flow cell to generate 414,215,817 single-end 100 bp reads (approximately 41Gb raw sequence data). Raw genomic sequencing data is publicly available via the SRA BioProject PRJNA1079333 (<https://www.ncbi.nlm.nih.gov/sra/PRJNA1079333>).

### Read mapping, variant calls, and filtering

Mapping, variant calling, and filtering was performed as in DeRaad *et al.* (2023) and DeRaad *et al.* (2024) [@deraad2023; @deraad2024]. Raw data was demultiplexed and reads of low quality and with uncalled bases were removed using *Stacks* v2.41 [@rochette2019]. Reads were then mapped to a publicly available *Z. japonicus* genome assembly available at <https://www.ncbi.nlm.nih.gov/assembly/GCA_017612475.1> [@venkatraman2021] using the *Burrows-Wheeler Alignment tool* (*BWA*) v0.7.17 [@li2009]. *SAMtools* v1.3.1 [@li2009a] was used to convert files to a sorted *bam* format and RAD loci were identified in *Stacks* using the ‘gstacks’ module and exported as a variant call format (VCF) file resulting in an unfiltered SNP dataset containing 236,767 SNPs shared among 155 individual samples with 65.5% missing genotypes. The R packages *vcfR* v1.14.0 [@vcfR] and *SNPfiltR* v1.0.1 [@SNPfiltR] were used to visualize parameter distributions and filter the SNP data for quality. The final filtered VCF file included 15,704 SNPs across 124 individual samples with 5.3% missing data. A more detailed description of the filtering protocol may be found in DeRaad *et al.* [@deraad2024] and at: <https://devonderaad.github.io/zosterops.rad/zost.radseq.filtering.html>.

A pairwise distance matrix based on Nei’s D [@nei1972] was generated from the filtered SNP dataset using the R package *StAMPP* v1.6.3 [@StAMPP] and used to construct a neighbor-net phylogenetic network using the R package *SplitsTree4* v4.15.1 [@huson2005]. This allowed us to visualize the data and, in some cases, validate the identity of individual samples for subsequent analyses. For example, the sample designated Zsim30897 was initially identified as *Z. simplex*, and given an identification code to match, however in a phylogenetic network this sample clustered with the very divergent *Z. palpebrosus* and was therefore revised to match that identity. Associated R code and the phylogenetic network from *SplitsTree* v4.15.1 are described in DeRaad et al. [@deraad2024] and may be viewed at <https://devonderaad.github.io/zosterops.rad/splitstree.html>.

### General SNP visualization and multivariate analyses

Various approaches were employed to visualize variation in the SNP dataset. The filtered SNP VCF file was converted to a genlight object for subsequent analyses using the ‘vcfR2genlight’ command in the R package *vcfR* [@vcfR]. The R package *Poppr* [@poppr] was used to generate a minimum spanning network (MSN) and a distance tree based on the unweighted pair group method (UPGMA). Bootstrapping was performed for the UPGMA tree with 100 replicates.

A principal component analysis (PCA) and a discriminant function analysis of principal components (DAPC) was conducted using the R package *adegenet* [@adegenet] following the methodology from Jombart *et al.* [@jombart2010]. DAPC is a multivariate method that takes principal components (PCs) and partitions variance into within- and among-group components in a way that maximizes discrimination among the groups. DAPC makes no explicit assumptions regarding recombination or panmixia. This allows use of the full SNP dataset without additional filtering to reduce the likelihood of linkage as is required for other clustering analyses such as *Structure* [@pritchard2000; @porras-hurtado2013] or *Admixture* [@alexander2009].

The DAPC analysis was conducted in R and based from a tutorial available at [https://grunwaldlab.github.io/Population_Genetics_in_R/DAPC.html](http://127.0.0.1:14213/#0 "https://grunwaldlab.github.io/Population_Genetics_in_R/DAPC.html"). Samples were grouped by a combination of species, subspecies, and geography with *Z. japonicus* split into the most far-flung *Z. japonicus alani* subspecies in the Ogasawara and Volcano Islands and a broader *Z. japonicus* group encompassing the remaining subspecies. *Z. montanus* samples were grouped into a Northern Philippine group including *Z. montanus whiteheadi* on Luzon and *Z. montanus halconensis* on Mindoro, *Z. montanus parkesi* on Palawan, and a southern group including the remaining *Z. montanus* subspecies from the Visayas, Camiguin South, Mindanao, Sulawesi, and Java. The remaining samples were assigned to groups based on their species.

PCA was conducted on the filtered SNP dataset (15,704 SNPs, 124 individual samples) using the sample groups described above. PCs were generated using the ‘glPca’ function in *adegenet* [@adegenet]. We generated plots in two dimensions (PCs 1-2) and in three dimensions (PCs 1-3) using the R package *plotly* [@plotly].  

We conducted a cross-validation analysis to help determine the optimal number of PCs to be retained in the DAPC by examining the proportion of successful assignments for each PC for the first 50 PCs and 100 replicate runs per PC. This analysis showed that successful assignments greatly increased from PC1 to PC13 and leveled off until declining at approximately PC30. We also estimated the difference between the proportion of successful reassignments using the real groups and assignments of individual samples into randomly assigned groups (i.e. the a-score). The mean a-score across replicate runs therefore estimates the average percent improvement in discrimination over that based on random groupings. We used the ‘dapc’ function in *adegenet* to estimate the mean a-score across 30 replicate runs and for each of the first 40 PCs. This process was repeated 100 times generating an optimization graph for each run. The median optimal number of PCs was calculated from these 100 runs and used in the subsequent DAPC analysis. The DAPC analysis was conducted based on the optimal number of PCs, determined as described above, in which we estimated the posterior membership probability for each individual sample. We conducted two additional DAPC analyses based on 50% fewer and 50% greater than the optimal number of PCs.

In addition to PCA we conducted a second dimensionality reduction on the SNP data using the uniform manifold approximation and projection (UMAP) method [@uniform2024]. Analysis via both methods allowed for the visualization of the SNP data by a linear method emphasizing variation (PCA) and a non-linear method emphasizing global structure among groups (UMAP). UMAP was performed with the ‘*umap*’ function in the R package *umap* [@umap] and plots generated in two and three dimensions using the R package *plotly* [@plotly].

## Data files

This analysis requires the following data files.

- `zost.filtered.snps.vcf.gz` The gzipped file containing the filtered SNP data in variant call format (VCF) (see DeRaad et al. 2024 for details on creation of the ddRAD library and filtering the SNP dataset) [@deraad2024]

- `ZosteropsPOPS4.txt` The text file containing the taxon and locale data for each individual sample in the SNP dataset.

## Taxa and populations

For this run the taxon file we are using is `ZosteropsPOPS4.txt` containing the following samples. Samples of the migratory *Zosterops erythropleurus* are from breeding and non-breeding sites. Breeding sites for *Z. erythropleurus* are denoted RUSSIAb and non-breeding sites from Vietnam and denoted VIETNAMnb with *b* indicating breeding site samples and *nb* indicating non-breeding site samples.

```         
SAMPLE TAXON LOCALE 
ZJlo012 japonicus KIKAI 
ZJlo015 japonicus AMAMI 
ZJlo016 japonicus MIYAKOJIMA 
ZJlo017 japonicus IRIOMOTE 
ZJlo021 japonicus TOKUNOSHIMA 
ZJlo022 japonicus TOKUNOSHIMA 
ZJlo024 japonicus YONAGUMI 
ZJlo031 japonicus KUMEJIMA 
ZJlo045 japonicus AMAMI 
ZJlo046 japonicus MIYAKOJIMA 
ZJlo050 japonicus OKINOERABU 
ZJlo052 japonicus OKINOERABU 
ZJlo055 japonicus KIKAI 
ZJja001 japonicus GEOJE 
ZJja002 japonicus GEOJE 
ZJja003 japonicus JEJU 
ZJja004 japonicus JEJU 
ZJja005 japonicus JEJU 
ZJja009 japonicus HEUKSANDO 
ZJja010 japonicus HEUKSANDO 
ZJja012 japonicus HONSHU 
ZJja013 japonicus HONSHU 
ZJja014 japonicus HONSHU 
ZJja016 japonicus SHIKOKU 
ZJja030 japonicus HEUKSANDO 
ZJal003 japonicus_alani HAHAJIMA 
Zjal004 japonicus_alani HAHAJIMA 
ZJal005 japonicus_alani HAHAJIMA 
ZJal010 japonicus_alani ANEJIMA 
ZJal011 japonicus_alani ANEJIMA 
ZJal012 japonicus_alani ANEJIMA 
ZJal020 japonicus_alani IWOTO 
ZJst007 japonicus KOZUSHIMA 
ZJst009 japonicus KOZUSHIMA 
ZJst010 japonicus KOZUSHIMA 
ZJst012 japonicus MIYAKEJIMA 
ZJst014 japonicus MIYAKEJIMA 
ZJst015 japonicus MIYAKEJIMA 
ZJin001 japonicus KAGOSHIMA 
ZJin003 japonicus KAGOSHIMA 
ZJsi002 simplex TAIWAN 
ZJsi017 simplex GUANGXI 
ZJsi032 simplex GUANGXI 
ZJsi023 simplex GUANGDONG 
ZJsi024 simplex GUANGDONG 
ZJsi027 simplex GUANGXI 
ZJsi028 simplex GUANGXI 
ZJsi029 simplex GUANGXI 
ZJsi031 simplex GUANGXI 
ZPxx001 simplex SINGAPORE 
ZPxx002 simplex SINGAPORE 
ZMxx002 meyeni BATAN 
ZMxx003 meyeni LANYU 
ZMxx004 meyeni LANYU 
ZMxx005 meyeni LANYU 
ZMxx006 meyeni LANYU 
ZMOxx001 montanus_SOUTH PANAY 
ZMOxx002 montanus_SOUTH PANAY 
ZMOxx003 montanus_SOUTH PANAY 
ZMOxx005 montanus_SOUTH PANAY 
ZMOha001 montanus_NORTH MINDORO 
ZMOwh002 montanus_NORTH LUZON 
ZMOwh003 montanus_NORTH LUZON 
ZMOwh004 montanus_NORTH LUZON 
ZMOwh005 montanus_NORTH LUZON 
ZMOwh006 montanus_NORTH LUZON 
ZMOvu002 montanus_SOUTH MINDANAO 
ZMOvu003 montanus_SOUTH MINDANAO 
ZMOvu004 montanus_SOUTH MINDANAO 
ZMOvu005 montanus_SOUTH MINDANAO 
ZMOmo001 montanus_SOUTH SULAWESI 
ZMOpa001 montanus_PALAWAN PALAWAN 
ZMOpa002 montanus_PALAWAN PALAWAN 
ZNni001 nigrorum PANAY 
ZERxx002 erythropleurus RUSSIAb 
ZERxx004 erythropleurus RUSSIAb 
ZERxx005 erythropleurus RUSSIAb 
Zsim23588 simplex VIETNAM 
Zsim28142 simplex VIETNAM 
Zsim30897 palpebrosus VIETNAM 
Zpal23498 palpebrosus VIETNAM 
Zpal23522 palpebrosus VIETNAM 
Zmon20893 montanus_SOUTH NEGROS
Zmon20892 montanus_SOUTH NEGROS 
Zmon20902 montanus_SOUTH NEGROS 
Zmon20909 montanus_SOUTH NEGROS 
Zsim31166 montanus_SOUTH JAVA 
Zsim31159 montanus_SOUTH JAVA 
Zmey17876 meyeni BATAN 
Zmey17877 meyeni BATAN 
Zmey17852 meyeni BATAN 
Zmey17922 meyeni SABTANG 
Zmey17920 meyeni SABTANG 
Zmey17925 meyeni SABTANG 
Zmey17923 meyeni SABTANG 
Zery28090 erythropleurus VIETNAMnb 
Zery28091 erythropleurus VIETNAMnb 
Zery28088 erythropleurus VIETNAMnb 
Zni10863 nigrorum CAMIGUINNORTE 
Zni14341 nigrorum CAMIGUINSUR 
Zni19650 nigrorum LUZON 
Zni17984 nigrorum LUZON 
ZjaDOT-10981 palpebrosus VIETNAM 
ZjaDOT-5235 simplex TAIWAN 
Z_LA_122866_2 simplex LOSANGELES 
Z_LA_122577_2 simplex LOSANGELES 
Z_LA_122188_2 simplex LOSANGELES 
Zsim13809 simplex GUIZHOU 
Zsim13773 simplex GUIZHOU 
Zsim6797 simplex HUNAN 
Zsim10336 simplex GUANGXI 
Zsim11362 simplex GUIZHOU 
Zsim11102 simplex GUIZHOU 
Zsim11220 simplex GUIZHOU 
Zmon20899 montanus_SOUTH NEGROS 
Zmon28375 montanus_SOUTH MINDANAO 
Zev13949 everetti CAMIGUINSUR 
Zev31650 everetti SAMAR 
Zev28451 everetti MINDANAO 
Z_HI_BRY431 japonicus HAWAII 
Z_HI_NAN290 japonicus HAWAII 
Z_HI_WAI087 japonicus HAWAII 
Z_HI_WAI078 japonicus HAWAII 
Z_HI_SOL783 japonicus HAWAII
```

## Packages

Begin the analysis by restarting the R session (Session/Restart R in RStudio), loading the initial package libraries, and starting a log of the analysis. Other package libraries will be loaded as needed in later blocks. The following R packages should be installed using the Tools/Install Packages feature in RStudio.

- logr (If you are interested in keeping a log of the analysis.) [@logr]

- vcfR [@vcfR]

- poppr [@poppr]

- ape [@ape] [@paradis2004]

- adegenet (adgenet is automatically installed along with ape) [@adegenet]

- vegan [@vegan]

- RcolorBrewer [@RColorBrewer]

- popgenstuff (Not available currently in the R package repository in the RStudio drop-down list. To install go [here](https://github.com/gbradburd/popgenstuff/).)

- umap [@umap]

- igraph [@igraph]

- plotly [@plotly]

- htmlwidgets [@htmlwidgets]

- ggplot2 [@ggplot2]

- tidyr [@tidyr]

```{r}
#| label: Load libraries
library(logr)
log_open(file_name = "ZosteropsSNPs.log")
library(vcfR)
library(poppr)
library(ape)
library(adegenet)
library(vegan)
library(RColorBrewer)
library(popgenstuff)
library(umap)
library(igraph)
library(plotly)
library(htmlwidgets)
library(ggplot2)
library(tidyr)
```

## VCF file

Load the filtered VCF dataset from a compressed gzipped (`gz`) file.

```{r}
#| label: Load SNPs from VCF file
# Load SNPs from VCF file
ZosteropsSNPs <- read.vcfR("Data/VCF files/zost.filtered.snps.vcf.gz")
ZosteropsSNPs
```

Extract allelic states from the `ZosteropsSNPs` from the 'gt' portion of the VCF file and create a new object. Estimate Wu and Watterson's 𝜃 [@watterson1975] from this new object using the package [popgenstuff](https://rdrr.io/github/gbradburd/popgenstuff/man/calcThetaW.html) (calculation includes only the polymorphic sites).

```{r}
#| label: Estimate theta W
#Theta W for polymorphic SNPs. Only calculating Theta W for polymorphic sites. 
ZosteropsSNPsgt <- extract.gt(ZosteropsSNPs)
calcThetaW(ZosteropsSNPsgt)
```

Define the populations from a tab delimited (\\t) text file and confirm that the sample names match those in the VCF file.

```{r}
#| label: Define taxa
# Define taxa and populations from tab delimited (\t) text file with a header and confirm sample names match the VCF file.
ZosteropsPOPs <- read.table("Data/Taxon files/ZosteropsPOPS4.txt", sep = "\t", header = TRUE)
all(colnames(ZosteropsSNPs@gt)[-1] == ZosteropsPOPs$SAMPLE)
```

Convert the `ZosteropsSNPs` VCF object to a genlight (`gl`) object, indicate the ploidy (2 for diploid), and assign taxa to the genlight object using `ZosteropsPOPs` and the TAXON column (`ZosteropsPOPs$TAXON`).

```{r}
#| label: Create genlight object
# Convert to genlight object, determine ploidy, and assign taxa to genlight object.
gl.ZosteropsSNPs <- vcfR2genlight(ZosteropsSNPs)
ploidy(gl.ZosteropsSNPs) <- 2
pop(gl.ZosteropsSNPs) <- ZosteropsPOPs$TAXON
```

## Color settings

Define the color schema to be used throughout the figures for each taxon.

```{r}
#| label: Color palette 
ColorPal_CUSTOM <- c("#004949", "#DB6D00", "#D82632", "#290AD8", "#FED439", 
                     "#B66DFF", "#006DDB", "#FA9C88", "#924900", 
                     "#7BA9CB", "#008600")
```

## Minimum spanning network (MSN)

Create a minimum spanning network (MSN). Export the figure as either a `pdf` or `png` file using the export function in RStudio. Beginning here we will set a different random seed for each operation where a seed is required. Setting `inds` to some set of characters not in the dataset (in this instance just a blank space) and `nodelab` to some arbitrarily large number in `plot_poppr_msn` will leave blank points on the network. Removing these arguments will result in each point on the network labeled with the corresponding sample name.

```{r}
#| label: Minimum spanning network
#| fig-width: 20
#| fig-height: 15
MSNseed = sample(1:999, 1)
print(MSNseed)
set.seed(MSNseed)
ZosteropsSNPsDist <- bitwise.dist(gl.ZosteropsSNPs)
ZosteropsSNPsMSN <- poppr.msn(gl.ZosteropsSNPs, ZosteropsSNPsDist, showplot = FALSE, include.ties = TRUE)
node.size <- rep(2, times = nInd(gl.ZosteropsSNPs))
names(node.size) <- indNames(gl.ZosteropsSNPs)
vertex.attributes(ZosteropsSNPsMSN$graph)$size <- node.size
plot_poppr_msn(gl.ZosteropsSNPs, ZosteropsSNPsMSN, palette = ColorPal_CUSTOM, gadj = 70, label.cex = 8, pop.leg = TRUE, size.leg = FALSE, inds = "", nodelab = 100000, )
```

## Distance tree

Create a distance tree based on the unweighted pair group method with arithmetic mean (UPGMA). Export the figure as either a `pdf` or `png` file using the export function in RStudio.

```{r}
#| label: Distance tree
#| fig-width: 27
#| fig-height: 24
ZosteropsSNPsDTree <- aboot(gl.ZosteropsSNPs, tree = "upgma", distance = bitwise.dist, sample = 100, showtree = FALSE, cutoff = 50, quiet = FALSE)
plot.phylo(ZosteropsSNPsDTree, cex = 1, font = 2, adj = 0)
nodelabels(ZosteropsSNPsDTree$node.label, adj = c(1.3, -0.5), frame = "n", cex = 1.3,font = 3, xpd = TRUE)
axis(side = 1, cex.axis = 2)
title(xlab = "Genetic distance (proportion of loci that are different)", cex.lab = 2)
```

## Principal component analysis (PCA)

Conduct a principal component analysis (PCA) on the `gl.ZosteropsSNPs` object. Graph the percent of variance explained by each principal component (PC) and create two PC plots; one plot with PC1 and PC2 and a three-dimensional plot with PC1, PC2, and PC3.

```{r}
#| label: Principal component analysis
PCAseed = sample(1:999, 1)
print(PCAseed)
set.seed(PCAseed)
ZosteropsSNPsPCA <- glPca(gl.ZosteropsSNPs, nf = 5)
ZosteropsSNPs_PCA_VARexp <- (100*ZosteropsSNPsPCA$eig/sum(ZosteropsSNPsPCA$eig))
barplot(100*ZosteropsSNPsPCA$eig/sum(ZosteropsSNPsPCA$eig), col = heat.colors(200), main = "PCA Eigenvalues", ylim = c(0,30), cex.axis = 1.4)
title(ylab = "Percent of variance explained", line =2, cex.lab = 1.4)
title(xlab = "Eigenvalues", line = 2, cex.lab = 1.4)
ZosteropsSNPsPCAscores <- as.data.frame(ZosteropsSNPsPCA$scores)
ZosteropsSNPsPCAscores$pop <- pop(gl.ZosteropsSNPs)
PCAplot <- ZosteropsSNPsPCAscores
levels(PCAplot$pop)[levels(PCAplot$pop)=="erythropleurus"] <- "Z. erythropleurus"
levels(PCAplot$pop)[levels(PCAplot$pop)=="everetti"] <- "Z. everetti"
levels(PCAplot$pop)[levels(PCAplot$pop)=="japonicus"] <- "Z. japonicus"
levels(PCAplot$pop)[levels(PCAplot$pop)=="japonicus_alani"] <- "Z. japonicus alani"
levels(PCAplot$pop)[levels(PCAplot$pop)=="meyeni"] <- "Z. meyeni"
levels(PCAplot$pop)[levels(PCAplot$pop)=="montanus_NORTH"] <- "Z. montanus N. PHILIPPINES"
levels(PCAplot$pop)[levels(PCAplot$pop)=="montanus_PALAWAN"] <- "Z. montanus PALAWAN"
levels(PCAplot$pop)[levels(PCAplot$pop)=="montanus_SOUTH"] <- "Z. montanus S. PHILIPPINES"
levels(PCAplot$pop)[levels(PCAplot$pop)=="nigrorum"] <- "Z. nigrorum"
levels(PCAplot$pop)[levels(PCAplot$pop)=="palpebrosus"] <- "Z. palpebrosus"
levels(PCAplot$pop)[levels(PCAplot$pop)=="simplex"] <- "Z. simplex"
names(PCAplot)[names(PCAplot)=="pop"]  <- "Taxon"
PC2Dplot <- plot_ly(x = PCAplot$PC1, y = PCAplot$PC2, type = "scatter", mode = 'markers',
        marker = list(size = 10, line = list(width = 1, color = 'black')),
        color = PCAplot$Taxon, 
        colors = ColorPal_CUSTOM) %>%
  layout(xaxis = list(title = 'PC1'), yaxis = list(title = 'PC2'))
PC3Dplot <- plot_ly(x = PCAplot$PC1, y = PCAplot$PC2, z = PCAplot$PC3, type = "scatter3d", mode = 'markers', 
        marker = list(size = 10, line = list(width = 1, color = 'black')),
        color = PCAplot$Taxon, 
        colors = ColorPal_CUSTOM) %>%
  layout(scene = list(xaxis = list(title = 'PC1'),
                      yaxis = list(title = 'PC2'),
                      zaxis = list(title = 'PC3')))
PC2Dplot
PC3Dplot
ZosteropsSNPs_PCA_VARexp[1:5]
```

Calculate the centroids for each group based on the first three principal components and the pairwise Euclidean distances between each pair of groups.

```{r}
#| label: Centroid distances based on the first 3 PCs

Zosterops_PCAscores <- ZosteropsSNPsPCA$scores[, 1:3]
Zosterops_PCAgroups <- gl.ZosteropsSNPs@pop
Zosterops_PCAdf <- data.frame(Zosterops_PCAscores, Group = Zosterops_PCAgroups)
Zosterops_unique_PCAgroups <- unique(Zosterops_PCAgroups)
Zosterops_PCAcentroids <- data.frame(Group = character(), PC1 = numeric(), PC2 = numeric(), PC3 = numeric(), PC4 = numeric(), PC5 = numeric(), stringsAsFactors = FALSE)
# Loop through each unique group to calculate centroids
for (group in Zosterops_unique_PCAgroups) {
  group_scores <- Zosterops_PCAdf[Zosterops_PCAdf$Group == group, 1:3]
  centroid <- colMeans(group_scores, na.rm = TRUE)
  Zosterops_PCAcentroids <- rbind(Zosterops_PCAcentroids, 
                                   data.frame(Group = group, 
                                              PC1 = centroid[1], 
                                              PC2 = centroid[2], 
                                              PC3 = centroid[3]))
}
print(Zosterops_PCAcentroids)

# Calculate the Euclidean distances between centroids
Zosterops_PCAdistance_matrix <- as.matrix(dist(Zosterops_PCAcentroids[, 2:4]))
Zosterops_PCAdistance_df <- as.data.frame(Zosterops_PCAdistance_matrix)
rownames(Zosterops_PCAdistance_df) <- Zosterops_PCAcentroids$Group
colnames(Zosterops_PCAdistance_df) <- Zosterops_PCAcentroids$Group
print(Zosterops_PCAdistance_df)

# Plot PCA scores
Zosterops_PCAcentroids_plot <- ggplot(Zosterops_PCAdf, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(alpha = 0.6) +  # Add points for PCA scores
  geom_point(data = Zosterops_PCAcentroids, aes(x = PC1, y = PC2), 
             color = "black", size = 3, shape = 3) +  # Add centroids
  labs(title = "PCA of Zosterops SNPs", x = "PC1", y = "PC2") +
  scale_color_manual(values = ColorPal_CUSTOM) +
  theme_minimal()

# Display the plot
print(Zosterops_PCAcentroids_plot)
```

Export both the two principal components and three principal components plots (`PC2Dplot` and `PC3Dplot`) as html widgets.

```{r}
#| label: Export PCA plots
# Export PC3Dplot as html using the htmlWidgets package or from the menu in the RStudio viewer.
saveWidget(ggplotly(PC2Dplot), file = "Zosterops2DPCAplot.html")
saveWidget(ggplotly(PC3Dplot), file = "Zosterops3DPCAplot.html")
```

## Uniform manifold approximation and projection (UMAP)

Uniform manifold approximation and projection (UMAP) [@uniform2024] is a non-linear, stochastic algorithm to reduce dimensionality for visualization of multivariate data. UMAP seeks to preserve as much of the relationships among groups in the dataset as possible while reducing to a minimal number of components. In this case we are visualizing the SNP data with UMAP for two and three dimensions using the `umap` package and plotting the results for each analysis using the `plot_ly` command from the `plotly` package. The `n_neighbors`, `min_dist`, and `metric arguments` all reflect the default values. For each plot save as `html` files using the `saveWidget` command in the `htmlwidgets` package.

```{r}
#| label: UMAP
# Set random seed
UMAPseed = sample(1:999, 1)
print(UMAPseed)
set.seed(UMAPseed)

# Define color palette
ColorPal_UMAP <- c("#004949", "#DB6D00", "#D82632", "#290AD8", "#FED439", 
                     "#B66DFF", "#006DDB", "#FA9C88", "#924900", 
                     "#7BA9CB", "#008600")

# Extract SNP data
ZosteropsSNPs_UMAP <- as.data.frame(tab(gl.ZosteropsSNPs))  # Use tab() to get the SNP data

ZosteropsUMAP_results <- umap(ZosteropsSNPs_UMAP, n_neighbors = 15, min_dist = 0.1, metric = 'euclidean')
ZosteropsUMAP_df <- as.data.frame(ZosteropsUMAP_results$layout)
ZosteropsUMAP_df$Taxon <- pop(gl.ZosteropsSNPs)  # Add population information
ZosteropsUMAP_2Dplot <- plot_ly(data = ZosteropsUMAP_df, x = ~V1, y = ~V2, type = 'scatter', mode = 'markers',
                     marker = list(size = 15, line = list(width = 1, color = 'black')),
                     color = ~Taxon, colors = ColorPal_UMAP) %>%
  layout(xaxis = list(title = 'UMAP 1'),
         yaxis = list(title = 'UMAP 2'))
ZosteropsUMAP_2Dplot

ZosteropsUMAP_df

ZosteropsUMAP_3Dresults <- umap(ZosteropsSNPs_UMAP, n_neighbors = 15, min_dist = 0.1, metric = 'euclidean', n_components = 3)
ZosteropsUMAP_3Ddf <- as.data.frame(ZosteropsUMAP_3Dresults$layout)
ZosteropsUMAP_3Ddf$Taxon <- pop(gl.ZosteropsSNPs)
ZosteropsUMAP_3Dplot <- plot_ly(data = ZosteropsUMAP_3Ddf, x = ~V1, y = ~V2, z = ~V3, type = 'scatter3d', mode = 'markers',
                        marker = list(size = 10, line = list(width = 1, color = 'black')),  # Add line around markers
                        color = ~Taxon, colors = ColorPal_UMAP) %>%
  layout(scene = list(xaxis = list(title = 'UMAP 1'),
                      yaxis = list(title = 'UMAP 2'),
                      zaxis = list(title = 'UMAP 3')))
ZosteropsUMAP_3Dplot

ZosteropsUMAP_3Ddf

saveWidget(ggplotly(ZosteropsUMAP_2Dplot), file = "ZosteropsUMAP_2Dplot.html")
saveWidget(ggplotly(ZosteropsUMAP_3Dplot), file = "ZosteropsUMAP_3Dplot.html")
```

## Discriminant analysis of principal components (DAPC)

#### DAPC cross-validation, a-scores, and optimal number of PCs for discrimination

Discriminant analysis of principal components (DAPC) is a multivariate method that takes principal components (PCs) and partitions variance into within- and among-group components in a way that maximizes discrimination among the groups [@jombart2010]. DAPC makes no explicit assumptions regarding recombination or panmixia. This allows use of the full SNP dataset without worrying about filtering to reduce the likelihood of linkage as is required for STRUCTURE or ADMIXTURE. A tutorial on DAPC is available [here](https://grunwaldlab.github.io/Population_Genetics_in_R/DAPC.html). First conduct a cross-validation analysis to help determine the optimal number of PCs to be retained in the DAPC. Test between 1 and 50 PCs (`n.pca = 1:50`) with 100 replicate runs per PC (`n.rep = 100`) of the cross-validation analysis. Use the multicore option and 6 CPUs (`ncpus = 6L`), or potentially more if your machine allows it. Even running multiple cores this run can take a while to execute. Executing this analysis within the `system.time` command will return an elapsed time for the operation. This will return a graph showing the proportion of successful group predictions on the y-axis and the results from the individual replicate runs for each PC on the x-axis.

```{r}
#| label: DAPC analysis cross validation
DAPCseed = sample(1:999, 1)
print(DAPCseed)
set.seed(DAPCseed)
system.time (ZosteropsSNPsDAPCpramx <- xvalDapc(tab(gl.ZosteropsSNPs, NA.method = "mean"), pop(gl.ZosteropsSNPs), n.pca = 1:50, n.rep = 100, parallel = "multicore", ncpus = 6L))
```

#### *a*-scores

The DAPC cross validation provides some clues as to where we no longer need to add PCs to the DAPC to get increased discrimination, but we want to know exactly how many PCs is optimum for discrimination. The *a*-score is the difference between the proportion of successful reassignments using the real groups and assignments into randomly assigned groups. Thus the average *a*-score is the average percent improvement in discrimination over that based on random groupings. The code below will estimate the *a*-score across 30 replicate runs (`n.sim = 30`) and 40 PCs (`n.pca = 40`) showing the result per group for each run, the average across all replicates per group, and the overall average.

```{r}
#| label: Optimum PCs for DAPC
ZosteropsSNPsDAPC_OPTIMUM <- dapc(gl.ZosteropsSNPs, n.pca = 40, n.da = nPop(gl.ZosteropsSNPs)-1)
a.score(ZosteropsSNPsDAPC_OPTIMUM, n.pca = 1:ncol(ZosteropsSNPsDAPC_OPTIMUM$tab), n=30, plot = FALSE, n.sim = 30)
```

#### *a*-score optimization graphs

To get an optimal number of PCs for the DAPC iterate the calculation of the *a*-score across 100 replicate runs and for each, graph the distribution of the *a*-score for each number of PCs.

```{r}
#| label: A-score optimization graphs
#| layout-ncol: 4
ZosteropsSNPsDAPC_OPTIMUM_RESULTS <- list()
for(i in 1:100) {
  ZosteropsSNPsDAPC_OPTIMUM_RESULTS[[i]] <- optim.a.score(ZosteropsSNPsDAPC_OPTIMUM)
}
```

Across 100 replicate runs of our iterated *a*-score calculation print the optimal number of PCs for each run. In each run the optimal number of PCs is that which maximizes the *a*-score.

```{r}
#| label: Best number of PCs for DAPC
OptimumPC_list <- vector("list", 100)
for(i in 1:100) {
  OptimumPC <- ZosteropsSNPsDAPC_OPTIMUM_RESULTS[[i]]$best
  print(OptimumPC)
  OptimumPC_list[[i]] <- OptimumPC
}
```

Produce a summary of the 100 iterated a-score estimates denoting the minimum and maximum number of optimal PCs across all runs as well as the median and mean optimal PCs and first and third quartiles. Use the median optimal number of PCs as a starting point in subsequent DAPC analyses.

```{r}
#| label: Summarize best number of PCs for DAPC
write.table(OptimumPC_list, file = "OptimumPC_results_n100.txt", row.names = FALSE, col.names = FALSE)
OptimumPC_list <- unlist(OptimumPC_list)
OptimumPC_list_summary <- summary(OptimumPC_list)
OptimumPC_list_summary
```

#### DAPC scatter plots and bar graphs

Conduct the final DAPC analysis with group assignments for each individual sample. Three runs will be repeated for this analysis. One of those runs uses the optimal number of PCs from the prior iterated *a*-score estimations (the median optimal number of PCs across 100 replicate runs of the optimal *a*-score estimation). The other two runs of the DAPC analysis are four PCs less than the optimal and four PCs greater than the optimal number of PCs. Generate a scatter plot from the DAPC analysis and a bar plot showing the probability of assignment for each individual sample. Save both the scatter plot and bar graphs as `pdf` files and generate a tab-delimited table of the results for each individual sample with the columns being the original assigned population, the sample, the population assigned by the DAPC, and the posterior probability of that particular assignment. In addition, for each replicate run generate a loading plot displaying how each SNP contributes to the PCs for the optimal number of PCs and four less than and four greater than the optimal number. Export each loading plot as a `pdf`.

```{r}
#| label: DAPC scatter and bar graphs
#| fig-height: 6
#| fig-width: 14
n.pca.values <- c(4, 8, 12)
for (n.pca in n.pca.values) {
  DAPCseed = sample(1:999, 1)
  print(DAPCseed)
  set.seed(DAPCseed)
  ZosteropsSNPsDAPC <- dapc(gl.ZosteropsSNPs, n.pca = n.pca, n.da = nPop(gl.ZosteropsSNPs)-1)
  ZosteropsSNPsDAPC_contributions <- loadingplot(ZosteropsSNPsDAPC$var.contr, axis = 2, thres = 0.0025, lab.jitter = 1)
  DAPCscatter <- scatter(ZosteropsSNPsDAPC, col = ColorPal_CUSTOM, cex = 3, legend = TRUE, clabel = F, posi.leg = "bottomleft", scree.pca = TRUE,
          posi.pca = "topright", cleg = 0.75)
  DAPCscatter
  ZosteropsSNPsDAPCResults <- as.data.frame(ZosteropsSNPsDAPC$posterior)
  ZosteropsSNPsDAPCResults$pop <- pop(gl.ZosteropsSNPs)
  ZosteropsSNPsDAPCResults$indNames <- rownames(ZosteropsSNPsDAPCResults)
  ZosteropsSNPsDAPCResults <- pivot_longer(ZosteropsSNPsDAPCResults, -c(pop, indNames))
  head(ZosteropsSNPsDAPCResults, n = 11)
  nPCAtitle <- paste(n.pca, "principal components")
  print(nPCAtitle)
  colnames(ZosteropsSNPsDAPCResults) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")
  DAPCbar <- ggplot(ZosteropsSNPsDAPCResults, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
  DAPCbar <-  DAPCbar + geom_bar(stat='identity') 
  DAPCbar <-  DAPCbar + ggtitle(nPCAtitle)
  DAPCbar <-  DAPCbar + scale_fill_manual(values = ColorPal_CUSTOM) 
  DAPCbar <-  DAPCbar + facet_grid(~Original_Pop, scales = "free")
  DAPCbar <-  DAPCbar + theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
  print(DAPCbar)
  write.table(ZosteropsSNPsDAPCResults, file = paste("ZosteropsSNPsDAPCResults","_",n.pca,"PCs.txt"), row.names = FALSE, sep = "\t", quote = FALSE)
}
```

*Last updated: `r format(Sys.time(), '%B %d, %Y %H:%M')`*

## Close log

Close the log file.

```{r}
log_close()
```
