---
title: "Linear morphometrics"

editor: visual
knitr: true
bibliography: ZosteropsMorphometrics.bib
---

## Goals

Included here are analyses of morphological phenotypes in East Asian Zosterops taken from linear measurements of museum skins at the Cincinnati Museum Center, National Museum of Natural History at the Smithsonian, the American Museum of Natural History, and the Field Museum of Natural History.

## Getting started

We are using R version 4.4.2, RStudio version 2024.09.1 for the analysis. Quarto version 1.6.39 for Mac OS was used to document the workflow. Set the working directory in RStudio to one where the `Zosterops_linear_measurements.csv` data file is accessible.

## General methods summary

Linear morphometric data including culmen length, nares-to-tip bill length, bill depth, bill width, tarsus length, wing cord, and tail length were collected from 388 round museum skins for all the Zosterops taxa in this study from the American Museum of Natural History (New York, NY, USA), Cincinnati Museum Center (Cincinnati, OH, USA), The Field Museum (Chicago, IL, USA), and the Smithsonian National Museum of Natural History (Washington DC, USA). All measurements were recorded by the same researcher (HLM) in mm using either digital calipers or a wing rule. Culmen length was measured as exposed culmen from the tip of the bill to the point where the bill meets the forehead plumage. Bill length from nares-to-tip was measured from the distal end of the right nare to the tip of the bill. Bill depth and width were measured at the nares. Tarsus (tarsometatarsus) length was measured from the base of the phalanges to the tibiotarsal joint. Wing cord was measured unflattened with a wing rule. Tail length was measured from the body to the tip of the longest rectrix by placing the same wing rule in the center of the tail. Skins where poor condition or positioning prevented accurate measurements for one or more traits were excluded from analysis making the final analyzed dataset 354 individual specimens.

A PCA was conducted using the `prcomp` function and plotted for PC1-2 and for PC1-3 using *plotly* in R with specimens grouped solely according to species. Specimens were further divided into 11 groups for pairwise statistics including a 'continental' group of *Z. japonicus* specimens from the main islands of Japan, the Ryukyu Archipelago, and the Korean Peninsula and surrounding islands, an 'oceanic' *Z. japonicus* group including those subspecies *stejnegeri* and *alani* on oceanic islands in the Pacific (Izu, Ogasawara, and Volcano Islands), two *Z. montanus* groups including a northern Philippine group on the islands of Luzon and Mindoro, and a southern *Z. montanus* group in the central and southern Philippines and Indonesia, and the *Z. nigrorum catarmanensis* subspecies on the Philippine island of Camiguin South. The remaining samples, *Z. erythropleurus*, *Z. everetti*, *Z. meyeni*, *Z. nigrorum* (excluding *catarmanesnis*), *Z. palpebrosus*, and *Z. simplex*, were each grouped according to species. Box-and-whisker plots were generated for both grouping schemas, solely by species and by a combination of geography and species, tests for normality were conducted, and pairwise Wilcoxon tests were conducted for each pair of groups.

## Packages

Set the working directory to one where the file `Zosterops_linear_measurements.csv` is accessible. Install the following packages.

- logr [@logr] Optional if you wish to create a log file for the analysis.

- ggplot2 [@ggplot2]

- plotly [@plotly]

- dplyr [@dplyr]

- tidyr [@tidyr]

- gt [@gt]

- factoextra [@factoextra]

- broom [@broom]

- readr [@readr]

- gridExtra [@gridExtra]

- htmlwidgets [@htmlwidgets]

Start a log file.

```{r}
#| label: Start a log file.
library(logr)
log_open(file_name = "ZosteropsMorphometrics.log")
```

Load the package libraries.

```{r}
#| label: Load package libraries. 
# Load necessary libraries
library(ggplot2)
library(plotly)
library(dplyr)
library(tidyr)
library(gt)
library(factoextra)
library(broom)
library(readr)
library(gridExtra)
library(grid)
library(htmlwidgets)
```

## Data files

### Final morphometric dataset

Morphological data for this analysis is in the comma delimited file titled `Zosterops_linear_measurements2.csv`. Import the morphological dataset using `read.csv` and summarize. In total 388 individual specimens were measured. Measurements for 354 skins were retained for analysis.

```{r}
#| label: Load data from .csv file.
# Import data from CSV file
ZosteropsMorph <- read.csv('./Data/Morphometrics/Zosterops_linear_measurements2.csv')
summary(ZosteropsMorph)
```

Create a table displaying all the data in the file `Zosterops_linear_measurements.csv`.

```{r}
#| label: Table for morphometric dataset.
ZosteropsMorph %>%
  gt(auto_align = FALSE)
```

## Principal component analyses (PCA)

### PCA for species

Conduct a principal component analysis (PCA) in which the groupings are species. Plot both the first two principal components (PCs) and the first three PCs. Create additional plots for three PCs in which the *Z. japonicus* samples are highlighted in black and different taxonomic groups are isolated and highlighted in grey for comparison (*Z. meyeni*, *Z. montanus*, *Z. simplex*, and *Z. palpebrosus*) while keeping all other taxa as open circles.

```{r}
#| label: PCA for species.
# Group data by 'species' and select relevant variables for PCA
ZosteropsMorph_species <- ZosteropsMorph %>% 
  select(Species, BillCulmen, BillNareTip, BillDepth, BillWidth, Tarsus, Wing, Tail) %>%
  group_by(Species)

# Perform PCA on the selected variables
pca_result <- prcomp(ZosteropsMorph_species[, -1], center = TRUE, scale. = TRUE)

# Convert PCA result to a data frame for plotting
pca_data <- as.data.frame(pca_result$x)

# Add 'species' variable to PCA data
pca_data$Species <- ZosteropsMorph_species$Species

# Plot the first two principal components
ColorPal_CUSTOM <- c("#004949", "#DB6D00", "#D82632", "#FED439", "#FA9C88", "#924900", "#8DB6CD", "#008600")
# Create the 2D scatter plot for PC1 and PC2 with outlines
PC2Dplot <- plot_ly(
  x = pca_data$PC1, 
  y = pca_data$PC2, 
  type = "scatter",  # Change to scatter for 2D plot
  mode = 'markers', 
  marker = list(size = 14, 
                line = list(color = 'black', width = 1)),  # Add black outline
  color = pca_data$Species, 
  colors = ColorPal_CUSTOM
) %>%
layout(
  xaxis = list(title = 'PC1'),
  yaxis = list(title = 'PC2'),
  title = "PCA of Zosterops Morphological Measurements"
)
PC2Dplot
# Plot the first three principal components
PC3Dplot <- plot_ly(x = pca_data$PC1, y = pca_data$PC2, z = pca_data$PC3, type = "scatter3d", mode = 'markers', 
        marker = list(size = 14, line = list(color = 'black', width = 1)),
        color = pca_data$Species, 
        colors = ColorPal_CUSTOM) %>%
  layout(scene = list(xaxis = list(title = 'PC1'),
                      yaxis = list(title = 'PC2'),
                      zaxis = list(title = 'PC3')))
PC3Dplot
# Plot of 3 PCs highlighting just the Z. japonicus and Z. meyeni taxa
# Define your custom color palette
ColorPal_japo_meye <- c("black", "grey", "transparent")  # Black for japonicus, grey for meyeni, transparent for others

# Create a vector for marker colors
MarkerCol_japo_meye <- ifelse(pca_data$Species == 'japonicus', 'black',
                               ifelse(pca_data$Species == 'meyeni', 'grey', 'transparent'))

# Create the 3D scatter plot
PC3D_japo_meye <- plot_ly(
  x = pca_data$PC1, 
  y = pca_data$PC2, 
  z = pca_data$PC3, 
  type = "scatter3d", 
  mode = 'markers', 
  marker = list(size = 10, color = MarkerCol_japo_meye, line = list(color = 'black', width = 2)),
  text = ifelse(pca_data$Species == 'japonicus', 'Z. japonicus',
                ifelse(pca_data$Species == 'meyeni', 'Z. meyeni', 'Other Species')),
  hoverinfo = "text"  # Show species name on hover
) %>%
layout(
  scene = list(
    xaxis = list(title = 'PC1'),
    yaxis = list(title = 'PC2'),
    zaxis = list(title = 'PC3')
  ),
  legend = list(title = list(text = "Species"))  # Add legend title
)

# Display the plot
PC3D_japo_meye
# Plot of 3 PCs highlighting just the Z. japonicus and Z. montanus taxa
# Define your custom color palette
ColorPal_japo_mont <- c("black", "grey", "transparent")  # Black for japonicus, grey for montanus, transparent for others

# Create a vector for marker colors
MarkerCol_japo_mont <- ifelse(pca_data$Species == 'japonicus', 'black',
                               ifelse(pca_data$Species == 'montanus', 'grey', 'transparent'))

# Create the 3D scatter plot
PC3D_japo_mont <- plot_ly(
  x = pca_data$PC1, 
  y = pca_data$PC2, 
  z = pca_data$PC3, 
  type = "scatter3d", 
  mode = 'markers', 
  marker = list(size = 10, color = MarkerCol_japo_mont, line = list(color = 'black', width = 2)),
  text = ifelse(pca_data$Species == 'japonicus', 'Z. japonicus',
                ifelse(pca_data$Species == 'montanus', 'Z. montanus', 'Other Species')),
  hoverinfo = "text"  # Show species name on hover
) %>%
layout(
  scene = list(
    xaxis = list(title = 'PC1'),
    yaxis = list(title = 'PC2'),
    zaxis = list(title = 'PC3')
  ),
  legend = list(title = list(text = "Species"))  # Add legend title
)

# Display the plot
PC3D_japo_mont
# Plot of 3 PCs highlighting just the Z. japonicus and Z. simplex taxa
# Define your custom color palette
ColorPal_japo_simp <- c("black", "grey", "transparent")  # Black for japonicus, grey for simplex, transparent for others

# Create a vector for marker colors
MarkerCol_japo_simp <- ifelse(pca_data$Species == 'japonicus', 'black',
                               ifelse(pca_data$Species == 'simplex', 'grey', 'transparent'))

# Create the 3D scatter plot
PC3D_japo_simp <- plot_ly(
  x = pca_data$PC1, 
  y = pca_data$PC2, 
  z = pca_data$PC3, 
  type = "scatter3d", 
  mode = 'markers', 
  marker = list(size = 10, color = MarkerCol_japo_simp, line = list(color = 'black', width = 2)),
  text = ifelse(pca_data$Species == 'japonicus', 'Z. japonicus',
                ifelse(pca_data$Species == 'simplex', 'Z. simplex', 'Other Species')),
  hoverinfo = "text"  # Show species name on hover
) %>%
layout(
  scene = list(
    xaxis = list(title = 'PC1'),
    yaxis = list(title = 'PC2'),
    zaxis = list(title = 'PC3')
  ),
  legend = list(title = list(text = "Species"))  # Add legend title
)

# Display the plot
PC3D_japo_simp
# Plot of 3 PCs highlighting just the Z. japonicus and Z. palpebrosus taxa
# Define your custom color palette
ColorPal_japo_palp <- c("black", "grey", "transparent")  # Black for japonicus, grey for palpebrosus, transparent for others

# Create a vector for marker colors
MarkerCol_japo_palp <- ifelse(pca_data$Species == 'japonicus', 'black',
                               ifelse(pca_data$Species == 'palpebrosus', 'grey', 'transparent'))

# Create the 3D scatter plot
PC3D_japo_palp <- plot_ly(
  x = pca_data$PC1, 
  y = pca_data$PC2, 
  z = pca_data$PC3, 
  type = "scatter3d", 
  mode = 'markers', 
  marker = list(size = 10, color = MarkerCol_japo_palp, line = list(color = 'black', width = 2)),
  text = ifelse(pca_data$Species == 'japonicus', 'Z. japonicus',
                ifelse(pca_data$Species == 'palpebrosus', 'Z. palpebrosus', 'Other Species')),
  hoverinfo = "text"  # Show species name on hover
) %>%
layout(
  scene = list(
    xaxis = list(title = 'PC1'),
    yaxis = list(title = 'PC2'),
    zaxis = list(title = 'PC3')
  ),
  legend = list(title = list(text = "Species"))  # Add legend title
)

# Display the plot
PC3D_japo_palp
# Export PC3Dplots as html using the htmlWidgets package or from the menu in the RStudio viewer.
saveWidget(ggplotly(PC3Dplot), file = "Zosterops3DPCAplot_morph_ALL.html")
saveWidget(ggplotly(PC3D_japo_meye), file = "Zosterops3DPCAplot_morph_japo_meye.html")
saveWidget(ggplotly(PC3D_japo_mont), file = "Zosterops3DPCAplot_morph_japo_mont.html")
saveWidget(ggplotly(PC3D_japo_simp), file = "Zosterops3DPCAplot_morph_japo_simp.html")
saveWidget(ggplotly(PC3D_japo_palp), file = "Zosterops3DPCAplot_morph_japo_palp.html")
```

For PCs 1 -3 calculate the centroids and Euclidean distances for pairwise comparisons between species.

```{r}
#| label: Calculate the centroids and Euclidean pairwise distances.
# Initialize an empty data frame to store centroids
ZosteropsMorph_PCAcentroids <- data.frame(Group = character(), PC1 = numeric(), PC2 = numeric(), PC3 = numeric(), stringsAsFactors = FALSE)

# Loop through each unique species to calculate centroids
for (species in unique(pca_data$Species)) {
  group_scores <- pca_data[pca_data$Species == species, 1:3]  # Select PCA scores for the species
  centroid <- colMeans(group_scores, na.rm = TRUE)  # Calculate the centroid
  ZosteropsMorph_PCAcentroids <- rbind(ZosteropsMorph_PCAcentroids, 
                                        data.frame(Group = species, 
                                                   PC1 = centroid[1], 
                                                   PC2 = centroid[2], 
                                                   PC3 = centroid[3]))
}

# Print the centroids
print(ZosteropsMorph_PCAcentroids)
# Calculate the Euclidean distances between centroids
ZosteropsMorph_PCAdistance_matrix <- as.matrix(dist(ZosteropsMorph_PCAcentroids[, 2:4]))  # Exclude the Group column
ZosteropsMorph_PCAdistance_df <- as.data.frame(ZosteropsMorph_PCAdistance_matrix)

# Set row and column names for the distance data frame
rownames(ZosteropsMorph_PCAdistance_df) <- ZosteropsMorph_PCAcentroids$Group
colnames(ZosteropsMorph_PCAdistance_df) <- ZosteropsMorph_PCAcentroids$Group

# Print the distance data frame
print(ZosteropsMorph_PCAdistance_df)
```

For the PCA produced from the SNP data Euclidean distances were also calculated between species (see [**Distance trees, PCA, UMAP, and DAPC analyses**](https://hermmays.github.io/Zosterops1/ZosteropsSNPs_analysis.html)). Compare the genetic distances with those from the morphometric dataset using Spearman's ⍴. Because different groupings were used in the genetic versus morphometric PCAs in some cases we are using the mean genetic distance. For example, in the genetic PCA *Z. montanus* was divided into north, south, and Palawan groups. Here we are taking the mean from the genetic distances between those groups and using that in pairwise comparisons of genetic distance between *Z. montanus* and other species. Below is the data manually entered into a comma delimited (`csv`) `txt` file (`Zosterops_genetic_morpho_dist.csv`) used in analyses of correlation between genetic and morphological pairwise distances.

```         
pairwise_comp,genetic_dist,morpho_dist
MONT_MEYE,1.60,1.36
MONT_EVER,36.23,0.88
MONT_NIGR,28.42,0.27
MONT_ERYT,18.93,2.44
MONT_JAPO,4.37,3.00
MONT_SIMP,26.40,2.07
MONT_PALP,35.81,2.60
MEYE_EVER,37.13,1.04
MEYE_NIGR,29.33,1.45
MEYE_ERYT,20.43,3.29
MEYE_JAPO,2.75,2.13
MEYE_SIMP,27.33,3.22
MEYE_PALP,37.22,3.81
EVER_NIGR,7.82,0.76
EVER_ERYT,28.28,2.52
EVER_JAPO,38.43,2.87
EVER_SIMP,37.80,2.43
EVER_PALP,39.36,2.89
NIGR_ERYT,21.81,2.23
NIGR_JAPO,30.69,3.08
NIGR_SIMP,32.50,1.90
NIGR_PALP,35.39,2.41
ERYT_JAPO,22.96,3.99
ERYT_SIMP,25.85,0.93
ERYT_PALP,18.54,1.48
JAPO_SIMP,29.36,4.18
JAPO_PALP,39.66,5.02
SIMP_PALP,39.09,0.94
```

Correlation analysis using Spearman's ⍴ between pairwise comparisons for genetic and morphological distance was calculated as follows. A plot of morphological and genetic distances was produced and pairwise comparisons involving *Z. japonicus* were highlighted (black points) from the remaining pairwise comparisons (open points).

```{r}
#| label: Spearman correlation between morphometric and genetic distances.
# Read the data from a CSV file
Zosterops_distance_data <- read.csv("Data/Morphometrics/Zosterops_genetic_morpho_dist.csv")

# Calculate the correlation for all comparisons using Spearman
correlation_all <- cor(Zosterops_distance_data$genetic_dist, Zosterops_distance_data$morpho_dist, method = "spearman")
print(paste("Correlation coefficient (all comparisons):", round(correlation_all, 3)))
# Perform a statistical significance test for all comparisons using Spearman
cor_test_all <- cor.test(Zosterops_distance_data$genetic_dist, Zosterops_distance_data$morpho_dist, method = "spearman")
print(cor_test_all)
# Filter data for JAPO comparisons
Zosterops_JAPO_data <- Zosterops_distance_data[grepl("JAPO", Zosterops_distance_data$pairwise_comp), ]

# Calculate the correlation for JAPO comparisons using Spearman
correlation_JAPO <- cor(Zosterops_JAPO_data$genetic_dist, Zosterops_JAPO_data$morpho_dist, method = "spearman")
print(paste("Correlation coefficient (JAPO comparisons):", round(correlation_JAPO, 3)))
# Perform a statistical significance test for JAPO comparisons using Spearman
cor_test_JAPO <- cor.test(Zosterops_JAPO_data$genetic_dist, Zosterops_JAPO_data$morpho_dist, method = "spearman")
print(cor_test_JAPO)
# Fit linear models for R-squared calculation
lm_model_all <- lm(morpho_dist ~ genetic_dist, data = Zosterops_distance_data)
r_squared_all <- summary(lm_model_all)$r.squared

lm_model_JAPO <- lm(morpho_dist ~ genetic_dist, data = Zosterops_JAPO_data)
r_squared_JAPO <- summary(lm_model_JAPO)$r.squared

# Plot the data
Zosterops_correlation_plot <- ggplot(Zosterops_distance_data, aes(x = genetic_dist, y = morpho_dist)) +
  geom_point(aes(shape = ifelse(grepl("JAPO", pairwise_comp), "filled", "open")), size = 3, color = "black", show.legend = FALSE) +  # Points based on shape, no legend
  scale_shape_manual(values = c("open" = 1, "filled" = 16)) +  # Open circles and filled circles
  labs(
       x = "Genetic Distance",
       y = "Morphological Distance") +
  theme_minimal() +
  theme(legend.position = "none")

# Display the plot
print(Zosterops_correlation_plot)
```

### PCA for subspecies

Conduct a PCA in which the groupings are subspecies. Only show a single plot for two PCs and all subspecies.

```{r}
#| label: Morphometric PCA for subspecies.
# Group data by 'Name' (subspecies) and select relevant variables for PCA
ZosteropsMorph_subspecies <- ZosteropsMorph %>%
  select(Name, BillCulmen, BillNareTip, BillDepth, BillWidth, Tarsus, Wing, Tail) %>%
  group_by(Name)

# Perform PCA on the selected variables
pca_result <- prcomp(ZosteropsMorph_subspecies[, -1], center = TRUE, scale. = TRUE)

# Convert PCA result to a data frame for plotting
pca_data <- as.data.frame(pca_result$x)

# Add 'Name' variable to PCA data for coloring
pca_data$Name <- ZosteropsMorph_subspecies$Name

# Plot the first two principal components
ggplot(pca_data, aes(x = PC1, y = PC2, color = Name)) +
  geom_point(size = 3) + 
  labs(title = "PCA of Zosterops Morphological Measurements",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

## Box-and-whisker plots and pairwise tests

### Species plots and summaries

Define color values for box plots to match prior color schemes.

```{r}
#| label: Define color scheme for box-and-whisker plots.
species_colors <- c("#004949", "#DB6D00", "#D82632", "#FED439", "#FA9C88", "#924900", "#8DB6CD", "#008600")
```

Create box-and-whisker plots for each individual trait and every species.

```{r}
#| label: Box-and-whisker plots for species and each trait.
# Define a function to create boxplots for each trait
create_boxplot <- function(ZosteropsMorph_species, trait) {
  ggplot(ZosteropsMorph_species, aes_string(x = "Species", y = trait)) + 
    geom_boxplot(outlier.shape = NA) + # Boxplot without outliers
    geom_jitter(width = 0.2, aes(color = Species), size = 2, alpha = 0.5) + # Add jitter for individual points
    labs(title = paste("Boxplot of", trait, "by Species"), x = "Species", y = trait) + 
    scale_color_manual(values = species_colors) +
    theme_minimal() # Minimal theme for better aesthetics
}
# List of traits to plot
traits <- c("BillCulmen", "BillNareTip", "BillDepth", "BillWidth", "Tarsus", "Wing", "Tail")
# Loop through each trait and create boxplots
for (trait in traits) {
  print(create_boxplot(ZosteropsMorph_species, trait)) # Print each boxplot
}
```

Descriptive statistics for morphometric traits by species.

```{r}
#| label: Descriptive statistics for morphometric traits by species.
ZosteropsMorph_descriptive_stats_species <- ZosteropsMorph_species %>%
  select(Species, BillCulmen, BillNareTip, BillDepth, BillWidth, Tarsus, Wing, Tail) %>%  # Select relevant columns
  group_by(Species) %>%  # Group by 'Group'
  summarise(  # Summarize to get descriptive statistics
    Mean_BillCulmen = mean(BillCulmen, na.rm = TRUE),
    Median_BillCulmen = median(BillCulmen, na.rm = TRUE),
    SD_BillCulmen = sd(BillCulmen, na.rm = TRUE),
    
    Mean_BillNareTip = mean(BillNareTip, na.rm = TRUE),
    Median_BillNareTip = median(BillNareTip, na.rm = TRUE),
    SD_BillNareTip = sd(BillNareTip, na.rm = TRUE),
    
    Mean_BillDepth = mean(BillDepth, na.rm = TRUE),
    Median_BillDepth = median(BillDepth, na.rm = TRUE),
    SD_BillDepth = sd(BillDepth, na.rm = TRUE),
    
    Mean_BillWidth = mean(BillWidth, na.rm = TRUE),
    Median_BillWidth = median(BillWidth, na.rm = TRUE),
    SD_BillWidth = sd(BillWidth, na.rm = TRUE),
    
    Mean_Tarsus = mean(Tarsus, na.rm = TRUE),
    Median_Tarsus = median(Tarsus, na.rm = TRUE),
    SD_Tarsus = sd(Tarsus, na.rm = TRUE),
    
    Mean_Wing = mean(Wing, na.rm = TRUE),
    Median_Wing = median(Wing, na.rm = TRUE),
    SD_Wing = sd(Wing, na.rm = TRUE),
    
    Mean_Tail = mean(Tail, na.rm = TRUE),
    Median_Tail = median(Tail, na.rm = TRUE),
    SD_Tail = sd(Tail, na.rm = TRUE)
  )
# View the descriptive statistics table
print(ZosteropsMorph_descriptive_stats_species)
write.csv(ZosteropsMorph_descriptive_stats_species)
ZosteropsMorph_descriptive_stats_species %>%
  gt(auto_align = FALSE)
```

### Taxonomic/biogeographic group plots, summaries, and t-tests

Define colors for box-and-whisker plots for taxonomic/biogeographic groups.

```{r}
#| label: Define the color schemes for taxonomic/biogeographic groups.
group_colors <- c("#004949", "#DB6D00", "#D82632", "#290AD8", "#FED439", "#B66DFF", "#FA9C88", "#924900", "#CDAA7D", "#8DB6CD", "#008600")
```

Test each trait within each taxonomic/biogeographic group for normality.

```{r}
#| label: Test each morphometric trait for normality.
# Select relevant variables and group by taxonomic group separating montanus into NORTH and SOUTH groups and japonicus into CONTINENTAL and OCEANIC groups and create datasets with the original data and log transformed data. 
ZosteropsMorph_groups <- ZosteropsMorph %>%
  select(Group, BillCulmen, BillNareTip, BillDepth, BillWidth, Tarsus, Wing, Tail) %>%
  group_by(Group)

# Function to perform Kolmogorov-Smirnov test for each variable within each group
perform_ks_test <- function(group_data, variable) {
  # Create a list to store results
  results <- list()
  
  # Get unique groups
  unique_groups <- unique(group_data$Group)
  
  # Iterate over each unique group
  for (i in 1:length(unique_groups)) {
    # Filter data for the specific group
    group_name <- unique_groups[i]
    group_values <- group_data %>% filter(Group == group_name) %>% pull(variable)
    
    # Perform KS test with the first group's data as reference
    if (i == 1) {
      reference_values <- group_values
    } else {
      ks_test_result <- ks.test(group_values, reference_values)
      results[[group_name]] <- ks_test_result
    }
  }
  return(results)
}
```

Box and whisker plots by taxonomic/biogeographic group splitting *Z. japonicus* into continental and oceanic groups, *Z. montanus* into north and south groups, and *Z. nigrorum* into the mismatched *Z. nigrorum catarmanensis* subspecies on Camiguin South and the remaining *Z. nigrorum* samples.

```{r}
#| label: Box-and-whisker plots by taxonomic/biogeographic group.
# Define a function to create boxplots for each trait
create_boxplot <- function(data, trait) {
  ggplot(data, aes_string(x = "Group", y = trait)) +
    geom_boxplot(outlier.shape = NA) + # Boxplot without outliers
    geom_jitter(width = 0.2, aes(color = Group), size = 2, alpha = 0.5) + # Add jitter for individual points
    scale_color_manual(values = group_colors) + # Define colors for groups
    theme_minimal() + # Minimal theme for better aesthetics
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Adjust labels on X axis
}

# List of traits to plot
traits <- c("BillCulmen", "BillNareTip", "BillDepth", "BillWidth", "Tarsus", "Wing", "Tail")

# Loop through each trait and create boxplots
for (trait in traits) {
  assign(paste0("boxplot_groups_", trait), create_boxplot(ZosteropsMorph_groups, trait)) # Create and save boxplot as an object named by trait
}
boxplot_groups_BillCulmen
boxplot_groups_BillNareTip
boxplot_groups_BillDepth
boxplot_groups_BillWidth
boxplot_groups_Tarsus
boxplot_groups_Wing
boxplot_groups_Tail
```

Isolate just the legend from one of the group box-and-whisker figures.

```{r}
#| label: Isolate the legend for the combined figure.
Zosterops_groups_legend <- cowplot::get_legend(boxplot_groups_Tail + theme(legend.position = "right"))
ggsave("boxplot_legend.png", plot = Zosterops_groups_legend, width = 1.75, height = 3)
```

Combine plots into a single figure and add the figure legend.

```{r}
#| label: Combine the box-and-whisker plots into a single legend.
#| fig-width: 18
#| fig-height: 25
ZosteropsMorph_groups_combined <- grid.arrange(
  # Plot A: Bill Culmen
  boxplot_groups_BillCulmen + 
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_blank(), 
          axis.title.y = element_text(size = 19),
          axis.text.y = element_text(size = 15)) +
    scale_y_continuous(limits = c(7.5, NA), name = "Bill length culmen (mm)") + 
    annotate("text", x = Inf, y = Inf, label = "a.", hjust = 1, vjust = 1.3, size = 12) +
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1)),
  
  # Plot B: Bill Nare Tip
  boxplot_groups_BillNareTip + 
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_blank(), 
          axis.title.y = element_text(size = 19),
          axis.text.y = element_text(size = 15)) + 
    scale_y_continuous(limits = c(5, NA), name = "Bill length nare-to-tip (mm)") + 
    annotate("text", x = Inf, y = Inf, label = "b.", hjust = 1, vjust = 1.2, size = 12) +
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1)),
  
  # Plot C: Bill Depth
  boxplot_groups_BillDepth + 
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_blank(), 
          axis.title.y = element_text(size = 19),
          axis.text.y = element_text(size = 15)) +
    scale_y_continuous(limits = c(1.5, NA), name = "Bill depth (mm)") + 
    annotate("text", x = Inf, y = Inf, label = "c.", hjust = 1, vjust = 1.3, size = 12) +
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1)),
  
  # Plot D: Bill Width
  boxplot_groups_BillWidth + 
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_blank(), 
          axis.title.y = element_text(size = 19),
          axis.text.y = element_text(size = 15)) + 
    scale_y_continuous(limits = c(2, NA), name = "Bill width (mm)") + 
    annotate("text", x = Inf, y = Inf, label = "d.", hjust = 1, vjust = 1.2, size = 12) +
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1)),
  
  # Plot E: Tarsus
  boxplot_groups_Tarsus + 
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_blank(), 
          axis.title.y = element_text(size = 19),
          axis.text.y = element_text(size = 15)) +
    scale_y_continuous(limits = c(10, NA), name = "Tarsus length (mm)") + 
    annotate("text", x = Inf, y = Inf, label = "e.", hjust = 1, vjust = 1.3, size = 12) +
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1)),
  
  # Plot F: Wing
  boxplot_groups_Wing + 
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_blank(), 
          axis.title.y = element_text(size = 19),
          axis.text.y = element_text(size = 15)) + 
    scale_y_continuous(limits = c(48, NA), name = "Wing cord length (mm)") + 
    annotate("text", x = Inf, y = Inf, label = "f.", hjust = 1, vjust = 1.2, size = 12) +
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1)),
  
  # Plot G: Tail
  boxplot_groups_Tail + 
    theme(legend.position = "none",
          axis.title.x = element_blank(),
          axis.text.x = element_blank(), 
          axis.title.y = element_text(size = 19),
          axis.text.y = element_text(size = 15)) +
    scale_y_continuous(limits = c(25, NA), name = "Tail length (mm)") + 
    annotate("text", x = Inf, y = Inf, label = "g.", hjust = 1, vjust = 0.9, size = 12) +
    theme(panel.border = element_rect(colour = "black", fill = NA, size = 1)),
  
  # Add image in 4th row, 2nd column
  grid::rasterGrob(png::readPNG("boxplot_legend.png"), interpolate = TRUE), 
  ncol = 2, 
  layout_matrix = rbind(c(1, 2), 
                        c(3, 4), 
                        c(5, 6), 
                        c(7, 8))
)
```

Produce descriptive statistics for morphometric traits by taxonomic/biogeographic group.

```{r}
#| label: Descriptive statistics by taxonomic/biogeographic groups.
ZosteropsMorph_descriptive_stats_groups <- ZosteropsMorph_groups %>%
  select(Group, BillCulmen, BillNareTip, BillDepth, BillWidth, Tarsus, Wing, Tail) %>%  # Select relevant columns
  group_by(Group) %>%  # Group by 'Group'
  summarise(  # Summarize to get descriptive statistics
    Mean_BillCulmen = mean(BillCulmen, na.rm = TRUE),
    Median_BillCulmen = median(BillCulmen, na.rm = TRUE),
    SD_BillCulmen = sd(BillCulmen, na.rm = TRUE),
    
    Mean_BillNareTip = mean(BillNareTip, na.rm = TRUE),
    Median_BillNareTip = median(BillNareTip, na.rm = TRUE),
    SD_BillNareTip = sd(BillNareTip, na.rm = TRUE),
    
    Mean_BillDepth = mean(BillDepth, na.rm = TRUE),
    Median_BillDepth = median(BillDepth, na.rm = TRUE),
    SD_BillDepth = sd(BillDepth, na.rm = TRUE),
    
    Mean_BillWidth = mean(BillWidth, na.rm = TRUE),
    Median_BillWidth = median(BillWidth, na.rm = TRUE),
    SD_BillWidth = sd(BillWidth, na.rm = TRUE),
    
    Mean_Tarsus = mean(Tarsus, na.rm = TRUE),
    Median_Tarsus = median(Tarsus, na.rm = TRUE),
    SD_Tarsus = sd(Tarsus, na.rm = TRUE),
    
    Mean_Wing = mean(Wing, na.rm = TRUE),
    Median_Wing = median(Wing, na.rm = TRUE),
    SD_Wing = sd(Wing, na.rm = TRUE),
    
    Mean_Tail = mean(Tail, na.rm = TRUE),
    Median_Tail = median(Tail, na.rm = TRUE),
    SD_Tail = sd(Tail, na.rm = TRUE)
  )
# View the descriptive statistics table
print(ZosteropsMorph_descriptive_stats_groups)
write.csv(ZosteropsMorph_descriptive_stats_groups)
library(gt)
ZosteropsMorph_descriptive_stats_groups %>%
  gt(auto_align = FALSE)
```

Conduct Wilcoxon rank-order pairwise tests among taxnonomic/biogeographic groups for each morphometric trait.

```{r}
#| label: Wilcoxon rank order tests aong taxonomic/biogeographic groups.
# Conduct a nonparametric Wilcoxon paired test for each measurement. 
WILCOX_BillCulmen <- pairwise.wilcox.test(ZosteropsMorph_groups$BillCulmen, ZosteropsMorph_groups$Group, p.adjust.method="bonferroni", exact = FALSE)
WILCOX_BillNareTip <- pairwise.wilcox.test(ZosteropsMorph_groups$BillNareTip, ZosteropsMorph_groups$Group, p.adjust.method="bonferroni", exact = FALSE)
WILCOX_BillDepth <- pairwise.wilcox.test(ZosteropsMorph_groups$BillDepth, ZosteropsMorph_groups$Group, p.adjust.method="bonferroni", exact = FALSE)
WILCOX_BillWidth <- pairwise.wilcox.test(ZosteropsMorph_groups$BillWidth, ZosteropsMorph_groups$Group, p.adjust.method="bonferroni", exact = FALSE)
WILCOX_Tarsus <- pairwise.wilcox.test(ZosteropsMorph_groups$Tarsus, ZosteropsMorph_groups$Group, p.adjust.method="bonferroni", exact = FALSE)
WILCOX_Wing <- pairwise.wilcox.test(ZosteropsMorph_groups$Wing, ZosteropsMorph_groups$Group, p.adjust.method="bonferroni", exact = FALSE)
WILCOX_Tail <- pairwise.wilcox.test(ZosteropsMorph_groups$Tail, ZosteropsMorph_groups$Group, p.adjust.method="bonferroni", exact = FALSE)
WILCOX_BillCulmen
WILCOX_BillNareTip
WILCOX_BillDepth
WILCOX_BillWidth
WILCOX_Tarsus
WILCOX_Wing
WILCOX_Tail
# Function to convert test results to a table format
generate_WILCOX_table <- function(test_result) {
  p_values <- format(test_result$p.value, nsmall = 2, digits = 2)
  data.frame(p_values)
}
# Generate tables for each test result
WILCOX_table_BillCulmen <- generate_WILCOX_table(WILCOX_BillCulmen)
WILCOX_table_BillNareTip <- generate_WILCOX_table(WILCOX_BillNareTip)
WILCOX_table_BillDepth <- generate_WILCOX_table(WILCOX_BillDepth)
WILCOX_table_BillWidth <- generate_WILCOX_table(WILCOX_BillWidth)
WILCOX_table_Tarsus <- generate_WILCOX_table(WILCOX_Tarsus)
WILCOX_table_Wing <- generate_WILCOX_table(WILCOX_Wing)
WILCOX_table_Tail <- generate_WILCOX_table(WILCOX_Tail)

WILCOX_table_BillCulmen
WILCOX_table_BillNareTip
WILCOX_table_BillDepth
WILCOX_table_BillWidth
WILCOX_table_Tarsus
WILCOX_table_Wing
WILCOX_table_Tail
```

*Last updated: `r format(Sys.time(), '%B %d, %Y %H:%M')`*

Close the log.

```{r}
#| label: Close the log file.
log_close()
```
